### 2018-03-30 wiw-study1-nb.Rmd
added first-pass analysis

---
title: "R Notebook"
output: html_notebook
---

```{r}

library(tidyverse)

d_raw <- read.csv("data/WIW+Judgment+Task+Preregistered_March+30%2C+2018_11.17.csv")

d_wide <- d_raw %>%
  rename(duration = Duration..in.seconds.) %>%
  select(workerId, age, sex, attn, duration, contains("_trial"), -contains("_trial_")) %>%
  filter(Check_trial == 0, attn == 1) %>%
  rename(
    Corruption_trial = corruption_trial,
    Clerk_trial = clerk_trial,
    Flattire_trial = flattire_trial,
    Rape_trial = rape_trial)

reverse_items <- c( # sadly, only 6 of 10 items were reversed! :(
  "Cancer_trial",
  "Heater_trial",
  "Flattire_trial",
  "Clerk_trial",
  "Corruption_trial",
  "Rape_trial"
)

d_tidy <- d_wide %>%
  gather(trial, response, Earthquake_trial:BigDiff5_trial) %>%
  mutate(trial_type = ifelse(grepl("[0-9]", trial), "filler", "experimental")) %>%
  mutate(trial_type = ifelse(grepl("BigDiff",trial), "filler-diff", trial_type)) %>%
  mutate(likely_2nd = ifelse((trial %in% reverse_items), 1, 0))

d <- d_tidy %>%
  filter(trial_type == "experimental", trial != "Check_trial") %>%
  mutate(response = ifelse(likely_2nd==1, (1-response), response))
```

```{r}
# normal = more likely is 1
# reverse = less likely is 1

d %>%
  group_by(trial) %>%
  summarize(mean_response = mean(response)) %>%
  ggplot(aes(x=trial, y = mean_response)) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
library(brms)
fit <- brm(
  response ~ 0 + intercept + (1|workerId) + (1|trial),
  data = d,
  family = bernoulli(), # student(), #cumulative(), #bernoulli(), etc
  control = list(adapt_delta = .80),
  cores = parallel::detectCores(),
  iter = 2000)
```

```{r}
summary(fit)
```

### 2018-03-30 wiw-study1-nb.Rmd
checked reverse coef

@@
 fit <- brm(
-  response ~ 0 + intercept + (1|workerId) + (1|trial),
+  response ~ 0 + intercept + likely_2nd + (1|workerId) + (1|trial),
   data = d,

### 2018-04-21 wiw-study2-nb.Rmd
added study 2 prelim analyses

---
title: "R Notebook"
output: html_notebook
---

```{r}

library(tidyverse)

d_raw <- read.csv("data/WIW+Judgment+Task+Preregistered+-+Likert_April+20%2C+2018_17.08.csv")

d_wide <- d_raw %>%
  rename(duration = Duration..in.seconds.) %>%
  select(workerId, age, sex, attn, duration, contains("_trial"), -contains("_trial_")) %>%
  filter(
    Check1_trial == -1, 
    Check2_trial == -2, 
    Check3_trial == 1, 
    attn == 1
    ) %>%
  rename(
    Corruption_trial = corruption_trial,
    Clerk_trial = clerk_trial,
    Flattire_trial = flattire_trial,
    Rape_trial = rape_trial)

# normal = more likely is 1
# reverse = less likely is 1

reverse_items <- c(
  "Cancer_trial",
  "Heater_trial",
  "Flattire_trial",
  "Clerk_trial",
  "Corruption_trial",
  "Rape_trial",
  "Earthquake_trial",
  "Fastfood_trial",
  "Police_trial"
)

d_tidy <- d_wide %>%
  select(-Check1_trial, -Check2_trial, -Check3_trial) %>%
  gather(trial, response, Tornado_trial:BigDiff5_trial) %>%
  mutate(trial_type = ifelse(grepl("[0-9]", trial), "filler", "experimental")) %>%
  mutate(trial_type = ifelse(grepl("BigDiff",trial), "filler-diff", trial_type)) %>%
  mutate(likely_2nd = ifelse((trial %in% reverse_items), 1, 0))

d <- d_tidy %>%
  filter(trial_type == "experimental") %>%
  mutate(response = ifelse(likely_2nd==1, response*-1, response))
```

```{r}
# normal = more likely is 1
# reverse = less likely is 1

d %>%
  group_by(trial) %>%
  summarize(mean_response = mean(response)) %>%
  ggplot(aes(x=trial, y = mean_response)) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
library(brms)

devtools::source_gist(id = "f1994c0f8325abbc5d300600744af39d", filename="cbrm.R")

fit_exp2 <- cbrm(
  response ~ trial + (1|workerId),
  data = d %>% mutate(response = response + 3),
  family = cumulative(), # student(), #cumulative(), #bernoulli(), etc
  prior = prior(normal(0,3),class="b"),
  control = list(adapt_delta = .80),
  cores = parallel::detectCores(),
  sample_prior = TRUE,
  iter = 2000,
  cached_file = "fit_exp2.rds"
  )
```

```{r}
summary(fit_exp2)

# So we get four intercepts representing the cumulative log odds of the different responses. 
# The question is whether the cumulative odds of responses 1 and 2 are greater than the cumulative log odds of responses 4 and 5. I'm getting this by comparing the `intercept[2]` 
# representing log odds of 1 and 2 against the inverse of `intercept[3]`, where `intercept[3]` itself represents the log odds of responses 1, 2, and 3 and therefore its inverse represents # the log odds of the remaining responses 4 and 5.

# test whether intercepts are different
h <- hypothesis(fit_exp2, "Intercept[2] = -Intercept[3]")

h$hypothesis$Evid.Ratio
```

```{r}
fit_exp2_re <- cbrm(
  response ~ (1|trial) + (1|workerId),
  data = d %>% mutate(response = response + 3),
  family = cumulative(), # student(), #cumulative(), #bernoulli(), etc
  # prior = prior(normal(0,3), class="Intercept"),
  control = list(adapt_delta = .80),
  cores = parallel::detectCores(),
  sample_prior = TRUE,
  iter = 2000,
  cached_file = "fit_exp2_re.rds"
  )
```

```{r}
summary(fit_exp2_re)

h2 <- hypothesis(fit_exp2_re, "Intercept[2] = -Intercept[3]")

h2$hypothesis$Evid.Ratio
```

### 2018-04-21 wiw-study2-nb.Rmd
minor update

@@
     Check3_trial == 1, 
-    attn == 1
+    attn == 1,
+    !is.na(workerId)
     ) %>%

### 2018-06-02 paper/paper.Rmd
added initial skeleton for rmarkdown paper

---
title: "Expectations bias moral evaluations"
author: "Derek Powell and Zachary Horne"
output: 
  pdf_document:
    keep_tex: TRUE
bibliography: library.bib
csl: apa6.csl

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

On the evening of November 13th, 2015, a terrorist attack in Paris left 130 people dead and injured over 300 more. In the aftermath, millions took to Twitter to express their shock, horror, and outrage at this tragedy under hashtags like `#parisattacks` and `#jesuisparis`. Yet, most of those mourning had little to say 15 hours earlier, when another tragic attack killed at least 43 people in Beirut. Several factors are surely at play in these different reactions  [e.g. group affiliations, @Brewer1999; construal-level theory, @Trope2010], yet one potentially fundamental factor has gone unmentioned: the fact that the Paris attack was more surprising than the attack in Beirut. In contrast to France, Lebanon had experienced dozens of terrorist bombings and attacks in recent years. Consequently, Beirut may seem to many like the sort of place where “these things happen,” whereas Paris is perceived as being stable and safe.

We can see in everyday experience that people’s evaluations of events often depend on their expectations about those events. There is often disappointment when events fail to meet expectations, and there is a special thrill to having one’s expectations exceeded. Anecdotally, these forces seem to drive people’s tendencies to root for the underdog, hold surprise parties, and foreshadow bad news to ease its delivery [@Bell1985]. Indeed, laboratory studies suggest that expectations play an important role in people’s evaluations of the utility of an event. For instance, Mellers and colleagues [-@Mellers1997] found that expectations influenced affective reactions during a gambling task: Given a gamble with a 10% chance to win \$30 and 90% chance to win \$0, participants felt little disappointment with the \$0 outcome, but were considerably more excited when they won \$30. Conversely, given a gamble with a 90% chance of winning \$60 and 10% chance of winning \$0, participants were disappointed with the \$0 outcome and showed more muted enjoyment of the \$60 outcome. In fact, in gambles similar to these, Mellers and colleagues [-@Mellers1997] found that people were happier with the smaller unexpected gain than with the larger but more expected gain [also see @Shepperd2002].

Just as expectations affect the utility of simple gambles, we suspect they are equally likely to shape how people react to morally harmful events, such as acts of terrorism (Waldmann Cite and Rai and Fiske cite --domain general peeps; us also). However, unlike in the context of gambles, in these contexts the effects of expectations on evaluations may have harmful consequences. When events are shocking, we predict that people will perceive them as more severe and consequently be roused to action. In contrast, when events harm victims who are generally considered to be at greater risk--the poor, sick, or those living in unstable regions of the world, we predicted that reactions may be more muted. However, prior research has yet to articulate whether these same effects emerge in the moral domain and the implications of this possibility. For instance, it is unclear if these same mechanisms can help account for the divergent reactions people have to terrorism in France and Beirut. Further, people’s moral evaluations are in fact impacted by expectations what is their source and what does that tell us about how to overcome the biases they may impose?

A number of researchers have sought to develop theories of disappointment--the psychological reactions that result when experiences fail to meet expectations--and its role in evaluation and decision-making [e.g., @Bell1985; @Gul1991; @Loomes1986]. These researchers have generally argued that expectations modify people's reactions to events, so that they are jointly determined by the context-free utilities of options or events (e.g., economic utilities), and the contextually-dependent disappointment that individuals experience as a function of their expectations. Similarly, Mellers and colleagues (1997) interpreted the divergence they observed between the utility of an outcome (i.e., monetary gains or losses) and the felt experience of that outcome as a result of post-decision affect.

Each of these theories has been rigorously defined and supported within the small-world confines they were meant to capture. However, we seek to craft a more general explanation of how and why expectations influence evaluations and the range of contexts in which they should be expected to do so. Several accounts of utility evaluation, including early theories like Prospect Theory [@Kahneman1979; @Tversky1992], have emphasized the role of relative comparisons in evaluations of utility. In a similar spirit, we argue that expectations can set the reference points against which people compare future outcomes. On this view, evaluating the utility of some event consists in comparing the state of the world now that the event has occurred to the state of the world just prior to the event. Further, we assume that our knowledge of the world is uncertain, so that it is really just our (probabilistic) expectations about what has happened, what is happening, and what will happen in the future. Thus, the evaluation of an event might be driven by what is learned about the state of the world as a result of that event’s occurrence.

The account outlined above provides an information-theoretic route to understanding how expectations directly influence evaluation. A fundamental insight of Information Theory is that the information carried by an event is a function of its prior probability. This means that low probability events carry more information than high probability events [@Shannon1948]. Along similar lines, violations of expectations have long been recognized as fundamental to associative and animal learning models; a larger delta means people learn more [e.g., @Rescorla1972]. Science progresses most when new findings impugn widely-accepted theories or when a theory’s extreme predictions are validated. Likewise, people learn more about the state of the world when their expectations are violated by shocking world events as compared to when they are affirmed by less surprising events. For example, if a bombing occurs in Paris -- an unexpected location -- rather than Lebanon --a more expected location--we learn that the world is more dangerous than we had previously believed.

A more formal treatment of these issues reveals the generality of these conclusions. Formally, before some binary event $x_i$ (which either occurs or does not occur), an evaluator has a mental model of the current state of affairs, S. This mental model includes expectations about potential future events and uncertainty over current and past events. Thus S represents a joint probability distribution over possible states of affairs $p(x_1, x_2, ..., x_n)$. The evaluator's subjective utility function can assign an expected utility to S--representing how positive or negatively the evaluator views the current state of affairs. The occurrence of $x_i$ brings about some new state of affairs, S', to which the evaluator can again apply her utility function. The evaluation of event  $x_i$ then consists in the comparison between these two states of affairs, those before the event (reflecting the agents' expectations), and the new state of affairs brought about by the event:

$$ V(x_i) = EU(S') -EU(S) $$

We let S be represented by a random vector X, whose members are any number of discrete events $x_1, x_2, ..., x_n$, where $x_i$ represents the event of interest so that $x_i$ occurs in S with some probability and occurs in S' with probability 1. That is, S represents $X$ and S' represents $X|x_i=1$. We write this as:

$$
\begin{aligned}
V(x_i) &= EU(X|x_i) - EU(X) \\
&= \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x)
\end{aligned}$$

Applying the chain rule of probability and some algebra (see appendix) we eventually obtain:

$$V(x_i) = \big(1 - p(x_i) \big) \sum_{x \in X} U(x)p(x|x_i)$$

Thus, the value assigned to event $x_i$ is proportional to the prior probability of $x_i$. Further, this holds true regardless of the nature of the event in question and of the form of the utility function over events.

In the context of our reaction to surprise parties, it might seem desirable that expectations would influence our experience of events.  Much joy is tied to the psychological marriage between surprise and evaluation. However, this basic cognitive process may have negative moral repercussions. When harms are expected, these dynamics may lead people to draw less extreme evaluations, feel less concern for victims, and be less driven to help. To test this hypothesis, we examined whether people’s evaluations of morally harmful events are affected by their expectations about those events. We asked people to compare pairs of simple events where a victim suffered an identical harm, but where the events differed in their prior probability. For each pair of events, participants were asked to judge which event was more upsetting. Two studies provided evidence for the predicted effect of expectations on moral evaluations: people tended to view unexpected negative events as more upsetting, even when the harm to victims was identical.

# General Methods

Here we present two preregistered studies examining the role of expectations in moral evaluations. Study 1 used a forced-choice task to test the hypothesis that people would be more upset about moral outcomes that were unexpected than expected. In Study 2, we tested of our hypothesis using a more conservative judgment task to increase the generalizability of our results.

## Materials

In both studies, participants were presented with a series of trials where they read brief (one sentence) descriptions of two different events and were asked to indicate which of the two events seemed more upsetting. In “experimental” trials, the two events were highly similar, but differed in their prior probabilities: one event was more expected and one more unexpected. (The perceived likelihood of a given event was confirmed in prior norming studies). These prior expectations were manipulated by changing the context in which the events occurred. For example, participants considered the following stimulus:

* “A 30 year old man in California dies in an earthquake” [Expected]

* “A 30 year old man in Oklahoma dies in an earthquake” [Unexpected]

In each event, the harm to the victim is the same (here, death) but one event is more expected than the other, given the different likelihoods of earthquakes occurring in California versus Oklahoma.

Each study contained between 16 experimental event-pairs that spanned a variety of different events and contexts. All experimental materials for these studies are available as supplemental online materials at [OSF LINK].

Both studies included “equivalent” filler trials. In these trials, the two events differed in trivial contextual details that we did not expect would affect participants’ judgments. For example:

* “A man in Connecticut starts a house fire.” [Equally expected]

* “A man in New Hampshire starts a house fire.” [Equally expected]

These filler trials were meant to prevent participants from becoming explicitly aware of the structure of the experimental trials. Finally, both studies included “non-equivalent” filler trials, the two events differed substantially in the degree of harm suffered by a victim, so that one event was expected to be seen as considerably more upsetting than the other. For example:

* “An 11-year-old child sets a doll on fire” [Less severe]

* “A 12-year-old child sets a cat on fire”  [More severe]

These trials were included to allow participants a chance to use the extremes of the response scale and to reduce any task demands that might drive them to make artificially fine-grained distinctions between the severity of events.

## Exclusions

In both studies, we excluded participants who failed to incorrectly answer attention-check questions. These questions asked participants to enter a particular response to ensure that they were paying attention and reading the items as they proceeded through the study. A final question asked participants if they had paid attention and taken the study seriously, encouraging them to be honest in their replies.

## Data Analysis

We analyzed our data by performing Bayesian estimation using the probabilistic programming language Stan [@Carpenter2017]. We tested our predictions by computing Bayes Factors (i.e. BF01) on the intercept term of our regression model using the hypothesis function in the R package brms. Bayes Factors express the ratio of the probability of data under the null hypothesis to the probability of the data under an alternative hypothesis. Therefore, larger Bayes Factors indicate that the data are more likely under the null hypothesis (e.g., that the intercept is not different from zero) than the alternative hypothesis (e.g., that the intercept is different from zero), and vice versa. Bayes Factors can be influenced by prior choices (Gelman, Simpson, & Betancourt, 2017) so we also performed prior robustness checks to confirm that our conclusions were unchanged under different prior assumptions.

# Appendix

$$ \begin{aligned}
V(x_i) &= EU(X|x_i) - EU(X) \\
&=  \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x) \\
&= \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x|x_i)p(x_i) \\
&= \sum_{x \in X} U(x)p(x|x_i) - p(x_i)\sum_{x \in X} U(x)p(x|x_i) \\
&= \big(1 - p(x_i) \big) \sum_{x \in X} U(x)p(x|x_i) \\
\end{aligned}$$

The first step utilizes the Law of the Unconscious Statistician, the next comes from the chain rule of probability, and the final two steps are simple algebra.

# References

\setlength{\parindent}{-0.25in} 
\setlength{\leftskip}{0.125in}
\noindent

### 2018-06-02 paper/paper.tex
added initial skeleton for rmarkdown paper

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Expectations bias moral evaluations},
            pdfauthor={Derek Powell and Zachary Horne},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Expectations bias moral evaluations}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Derek Powell and Zachary Horne}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
    \date{}
    \predate{}\postdate{}

\begin{document}
\maketitle

\section{Introduction}\label{introduction}

On the evening of November 13th, 2015, a terrorist attack in Paris left
130 people dead and injured over 300 more. In the aftermath, millions
took to Twitter to express their shock, horror, and outrage at this
tragedy under hashtags like \texttt{\#parisattacks} and
\texttt{\#jesuisparis}. Yet, most of those mourning had little to say 15
hours earlier, when another tragic attack killed at least 43 people in
Beirut. Several factors are surely at play in these different reactions
(construal-level theory, {\textbf{???}}; e.g.~group affiliations,
Brewer, 1999), yet one potentially fundamental factor has gone
unmentioned: the fact that the Paris attack was more surprising than the
attack in Beirut. In contrast to France, Lebanon had experienced dozens
of terrorist bombings and attacks in recent years. Consequently, Beirut
may seem to many like the sort of place where ``these things happen,''
whereas Paris is perceived as being stable and safe.

We can see in everyday experience that people's evaluations of events
often depend on their expectations about those events. There is often
disappointment when events fail to meet expectations, and there is a
special thrill to having one's expectations exceeded. Anecdotally, these
forces seem to drive people's tendencies to root for the underdog, hold
surprise parties, and foreshadow bad news to ease its delivery (Bell,
1985). Indeed, laboratory studies suggest that expectations play an
important role in people's evaluations of the utility of an event. For
instance, Mellers and colleagues (1997) found that expectations
influenced affective reactions during a gambling task: Given a gamble
with a 10\% chance to win \$30 and 90\% chance to win \$0, participants
felt little disappointment with the \$0 outcome, but were considerably
more excited when they won \$30. Conversely, given a gamble with a 90\%
chance of winning \$60 and 10\% chance of winning \$0, participants were
disappointed with the \$0 outcome and showed more muted enjoyment of the
\$60 outcome. In fact, in gambles similar to these, Mellers and
colleagues (1997) found that people were happier with the smaller
unexpected gain than with the larger but more expected gain (also see
Shepperd \& Mcnulty, 2002).

Just as expectations affect the utility of simple gambles, we suspect
they are equally likely to shape how people react to morally harmful
events, such as acts of terrorism (Waldmann Cite and Rai and Fiske cite
--domain general peeps; us also). However, unlike in the context of
gambles, in these contexts the effects of expectations on evaluations
may have harmful consequences. When events are shocking, we predict that
people will perceive them as more severe and consequently be roused to
action. In contrast, when events harm victims who are generally
considered to be at greater risk--the poor, sick, or those living in
unstable regions of the world, we predicted that reactions may be more
muted. However, prior research has yet to articulate whether these same
effects emerge in the moral domain and the implications of this
possibility. For instance, it is unclear if these same mechanisms can
help account for the divergent reactions people have to terrorism in
France and Beirut. Further, people's moral evaluations are in fact
impacted by expectations what is their source and what does that tell us
about how to overcome the biases they may impose?

A number of researchers have sought to develop theories of
disappointment--the psychological reactions that result when experiences
fail to meet expectations--and its role in evaluation and
decision-making (e.g., Bell, 1985; Gul, 1991; Loomes \& Sugden, 1986).
These researchers have generally argued that expectations modify
people's reactions to events, so that they are jointly determined by the
context-free utilities of options or events (e.g., economic utilities),
and the contextually-dependent disappointment that individuals
experience as a function of their expectations. Similarly, Mellers and
colleagues (1997) interpreted the divergence they observed between the
utility of an outcome (i.e., monetary gains or losses) and the felt
experience of that outcome as a result of post-decision affect.

Each of these theories has been rigorously defined and supported within
the small-world confines they were meant to capture. However, we seek to
craft a more general explanation of how and why expectations influence
evaluations and the range of contexts in which they should be expected
to do so. Several accounts of utility evaluation, including early
theories like Prospect Theory (Kahneman \& Tversky, 1979; Tversky \&
Kahneman, 1992), have emphasized the role of relative comparisons in
evaluations of utility. In a similar spirit, we argue that expectations
can set the reference points against which people compare future
outcomes. On this view, evaluating the utility of some event consists in
comparing the state of the world now that the event has occurred to the
state of the world just prior to the event. Further, we assume that our
knowledge of the world is uncertain, so that it is really just our
(probabilistic) expectations about what has happened, what is happening,
and what will happen in the future. Thus, the evaluation of an event
might be driven by what is learned about the state of the world as a
result of that event's occurrence.

The account outlined above provides an information-theoretic route to
understanding how expectations directly influence evaluation. A
fundamental insight of Information Theory is that the information
carried by an event is a function of its prior probability. This means
that low probability events carry more information than high probability
events (Shannon, 1948). Along similar lines, violations of expectations
have long been recognized as fundamental to associative and animal
learning models; a larger delta means people learn more (e.g., Rescorla
\& Wagner, 1972). Science progresses most when new findings impugn
widely-accepted theories or when a theory's extreme predictions are
validated. Likewise, people learn more about the state of the world when
their expectations are violated by shocking world events as compared to
when they are affirmed by less surprising events. For example, if a
bombing occurs in Paris -- an unexpected location -- rather than Lebanon
--a more expected location--we learn that the world is more dangerous
than we had previously believed.

A more formal treatment of these issues reveals the generality of these
conclusions. Formally, before some binary event \(x_i\) (which either
occurs or does not occur), an evaluator has a mental model of the
current state of affairs, S. This mental model includes expectations
about potential future events and uncertainty over current and past
events. Thus S represents a joint probability distribution over possible
states of affairs \(p(x_1, x_2, ..., x_n)\). The evaluator's subjective
utility function can assign an expected utility to S--representing how
positive or negatively the evaluator views the current state of affairs.
The occurrence of \(x_i\) brings about some new state of affairs, S', to
which the evaluator can again apply her utility function. The evaluation
of event \(x_i\) then consists in the comparison between these two
states of affairs, those before the event (reflecting the agents'
expectations), and the new state of affairs brought about by the event:

\[ V(x_i) = EU(S') -EU(S) \]

We let S be represented by a random vector X, whose members are any
number of discrete events \(x_1, x_2, ..., x_n\), where \(x_i\)
represents the event of interest so that \(x_i\) occurs in S with some
probability and occurs in S' with probability 1. That is, S represents
\(X\) and S' represents \(X|x_i=1\). We write this as:

\[
\begin{aligned}
V(x_i) &= EU(X|x_i) - EU(X) \\
&= \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x)
\end{aligned}\]

Applying the chain rule of probability and some algebra (see appendix)
we eventually obtain:

\[V(x_i) = \big(1 - p(x_i) \big) \sum_{x \in X} U(x)p(x|x_i)\]

Thus, the value assigned to event \(x_i\) is proportional to the prior
probability of \(x_i\). Further, this holds true regardless of the
nature of the event in question and of the form of the utility function
over events.

In the context of our reaction to surprise parties, it might seem
desirable that expectations would influence our experience of events.
Much joy is tied to the psychological marriage between surprise and
evaluation. However, this basic cognitive process may have negative
moral repercussions. When harms are expected, these dynamics may lead
people to draw less extreme evaluations, feel less concern for victims,
and be less driven to help. To test this hypothesis, we examined whether
people's evaluations of morally harmful events are affected by their
expectations about those events. We asked people to compare pairs of
simple events where a victim suffered an identical harm, but where the
events differed in their prior probability. For each pair of events,
participants were asked to judge which event was more upsetting. Two
studies provided evidence for the predicted effect of expectations on
moral evaluations: people tended to view unexpected negative events as
more upsetting, even when the harm to victims was identical.

\section{General Methods}\label{general-methods}

Here we present two preregistered studies examining the role of
expectations in moral evaluations. Study 1 used a forced-choice task to
test the hypothesis that people would be more upset about moral outcomes
that were unexpected than expected. In Study 2, we tested of our
hypothesis using a more conservative judgment task to increase the
generalizability of our results.

\subsection{Materials}\label{materials}

In both studies, participants were presented with a series of trials
where they read brief (one sentence) descriptions of two different
events and were asked to indicate which of the two events seemed more
upsetting. In ``experimental'' trials, the two events were highly
similar, but differed in their prior probabilities: one event was more
expected and one more unexpected. (The perceived likelihood of a given
event was confirmed in prior norming studies). These prior expectations
were manipulated by changing the context in which the events occurred.
For example, participants considered the following stimulus:

\begin{itemize}
\item
  ``A 30 year old man in California dies in an earthquake''
  {[}Expected{]}
\item
  ``A 30 year old man in Oklahoma dies in an earthquake''
  {[}Unexpected{]}
\end{itemize}

In each event, the harm to the victim is the same (here, death) but one
event is more expected than the other, given the different likelihoods
of earthquakes occurring in California versus Oklahoma.

Each study contained between 16 experimental event-pairs that spanned a
variety of different events and contexts. All experimental materials for
these studies are available as supplemental online materials at {[}OSF
LINK{]}.

Both studies included ``equivalent'' filler trials. In these trials, the
two events differed in trivial contextual details that we did not expect
would affect participants' judgments. For example:

\begin{itemize}
\item
  ``A man in Connecticut starts a house fire.'' {[}Equally expected{]}
\item
  ``A man in New Hampshire starts a house fire.'' {[}Equally expected{]}
\end{itemize}

These filler trials were meant to prevent participants from becoming
explicitly aware of the structure of the experimental trials. Finally,
both studies included ``non-equivalent'' filler trials, the two events
differed substantially in the degree of harm suffered by a victim, so
that one event was expected to be seen as considerably more upsetting
than the other. For example:

\begin{itemize}
\item
  ``An 11-year-old child sets a doll on fire'' {[}Less severe{]}
\item
  ``A 12-year-old child sets a cat on fire'' {[}More severe{]}
\end{itemize}

These trials were included to allow participants a chance to use the
extremes of the response scale and to reduce any task demands that might
drive them to make artificially fine-grained distinctions between the
severity of events.

\subsection{Exclusions}\label{exclusions}

In both studies, we excluded participants who failed to incorrectly
answer attention-check questions. These questions asked participants to
enter a particular response to ensure that they were paying attention
and reading the items as they proceeded through the study. A final
question asked participants if they had paid attention and taken the
study seriously, encouraging them to be honest in their replies.

\subsection{Data Analysis}\label{data-analysis}

We analyzed our data by performing Bayesian estimation using the
probabilistic programming language Stan (Carpenter et al., 2017). We
tested our predictions by computing Bayes Factors (i.e.~BF01) on the
intercept term of our regression model using the hypothesis function in
the R package brms. Bayes Factors express the ratio of the probability
of data under the null hypothesis to the probability of the data under
an alternative hypothesis. Therefore, larger Bayes Factors indicate that
the data are more likely under the null hypothesis (e.g., that the
intercept is not different from zero) than the alternative hypothesis
(e.g., that the intercept is different from zero), and vice versa. Bayes
Factors can be influenced by prior choices (Gelman, Simpson, \&
Betancourt, 2017) so we also performed prior robustness checks to
confirm that our conclusions were unchanged under different prior
assumptions.

\section{Appendix}\label{appendix}

\[ \begin{aligned}
V(x_i) &= EU(X|x_i) - EU(X) \\
&=  \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x) \\
&= \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x|x_i)p(x_i) \\
&= \sum_{x \in X} U(x)p(x|x_i) - p(x_i)\sum_{x \in X} U(x)p(x|x_i) \\
&= \big(1 - p(x_i) \big) \sum_{x \in X} U(x)p(x|x_i) \\
\end{aligned}\]

The first step utilizes the Law of the Unconscious Statistician, the
next comes from the chain rule of probability, and the final two steps
are simple algebra.

\section{References}\label{references}

\setlength{\parindent}{-0.25in} \setlength{\leftskip}{0.125in} \noindent

\hypertarget{refs}{}
\hypertarget{ref-Bell1985}{}
Bell, D. (1985). Disappointment In Decision Making Under Uncertainty.
\emph{Operations Research}, \emph{33}(1), 1--27.
\url{http://doi.org/10.1287/opre.33.1.1}

\hypertarget{ref-Brewer1999}{}
Brewer, M. B. (1999). The psychology of prejudice: Ingroup love or
outgroup hate? \emph{Journal of Social Issues}, \emph{55}(3), 429--444.
\url{http://doi.org/10.1111/0022-4537.00126}

\hypertarget{ref-Carpenter2017}{}
Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B.,
Betancourt, M., \ldots{} Riddell, A. (2017). Stan: A probabilistic
programming language. \emph{Journal of Statistical Software},
\emph{76}(1). \url{http://doi.org/10.18637/jss.v076.i01}

\hypertarget{ref-Gul1991}{}
Gul, F. (1991). A theory of disappointment aversion.
\emph{Econometrica}, \emph{59}(3), 667--686.
\url{http://doi.org/10.2307/2938223}

\hypertarget{ref-Kahneman1979}{}
Kahneman, D., \& Tversky, A. (1979). Prospect Theory: An Analysis of
Decision under Risk. \emph{Econometrica}, \emph{47}(2), 263--292.

\hypertarget{ref-Loomes1986}{}
Loomes, G., \& Sugden, R. (1986). Disappointment and Dynamic Consistency
in Choice under Uncertainty. \emph{Review of Economic Studies},
\emph{53}(2), 271--282. \url{http://doi.org/10.2307/2297651}

\hypertarget{ref-Mellers1997}{}
Mellers, B. a, Schwartz, a., Ho, K., \& Ritov, I. (1997). Decision
Affect Theory: Emotional Reactions to the Outcomes of Risky Options.
\emph{Psychological Science}, \emph{8}(6), 423--429.
\url{http://doi.org/10.1111/j.1467-9280.1997.tb00455.x}

\hypertarget{ref-Rescorla1972}{}
Rescorla, R. A., \& Wagner, A. R. (1972). A theory of Pavlovian
conditioning: Variations in the effectiveness of reinforcement and
nonreinforcement. In \emph{Classical conditioning ii current research
and theory} (Vol. 21, pp. 64--99).
\url{http://doi.org/10.1101/gr.110528.110}

\hypertarget{ref-Shannon1948}{}
Shannon, C. E. (1948). A Mathematical Theory of Communication.
\emph{Bell System Technical Journal}, \emph{27}(3), 379--423.
\url{http://doi.org/10.1002/j.1538-7305.1948.tb01338.x}

\hypertarget{ref-Shepperd2002}{}
Shepperd, J. a, \& Mcnulty, J. K. (2002). The affective consequences of
expected and unexpected outcomes. \emph{Psychological Science},
\emph{13}(1), 85--88. \url{http://doi.org/10.1111/1467-9280.00416}

\hypertarget{ref-Tversky1992}{}
Tversky, A., \& Kahneman, D. (1992). Advances in prospect theory:
Cumulative representation of uncertainty. \emph{Journal of Risk and
Uncertainty}, \emph{5}, 297--323.

\end{document}

### 2018-06-04 paper/paper.Rmd
updated paper with results

@@
 csl: apa6.csl
-
+fontsize: 12pt
+header-includes:
+    - \usepackage{fancyhdr}
+    - \pagestyle{fancy}
+    - \fancyhead{}
+    - \fancyhead[LO,LE]{PREPRINT DRAFT - PLEASE DO NOT CITE WITHOUT PERMISSION}
+    - \fancyfoot[CO,CE]{}
 ---
@@
 
+```{r load data, include=FALSE}
+# analysis
+
+library(tidyverse)
+
+d_raw <- read.csv("../data/WIW+Judgment+Task+Preregistered_March+30%2C+2018_11.17.csv")
+
+d_wide <- d_raw %>%
+  rename(duration = Duration..in.seconds.) %>%
+  select(workerId, age, sex, attn, duration, contains("_trial"), -contains("_trial_")) %>%
+  filter(Check_trial == 0, attn == 1) %>%
+  rename(
+    Corruption_trial = corruption_trial,
+    Clerk_trial = clerk_trial,
+    Flattire_trial = flattire_trial,
+    Rape_trial = rape_trial) %>%
+  mutate(sex = if_else(sex == 1, "male", "female"))
+
+reverse_items <- c( # sadly, only 6 of 10 items were reversed! :(
+  "Cancer_trial",
+  "Heater_trial",
+  "Flattire_trial",
+  "Clerk_trial",
+  "Corruption_trial",
+  "Rape_trial"
+)
+
+d_tidy <- d_wide %>%
+  gather(trial, response, Earthquake_trial:BigDiff5_trial) %>%
+  mutate(trial_type = ifelse(grepl("[0-9]", trial), "filler", "experimental")) %>%
+  mutate(trial_type = ifelse(grepl("BigDiff",trial), "filler-diff", trial_type)) %>%
+  mutate(likely_2nd = ifelse((trial %in% reverse_items), 1, 0))
+
+d <- d_tidy %>%
+  filter(trial_type == "experimental", trial != "Check_trial") %>%
+  mutate(response = ifelse(likely_2nd==1, (1-response), response))
+
+d2_raw <- read.csv("../data/WIW+Judgment+Task+Preregistered+-+Likert_April+20%2C+2018_17.08.csv")
+
+d2_wide <- d2_raw %>%
+  rename(duration = Duration..in.seconds.) %>%
+  select(workerId, age, sex, attn, duration, contains("_trial"), -contains("_trial_")) %>%
+  filter(
+    Check1_trial == -1, 
+    Check2_trial == -2, 
+    Check3_trial == 1, 
+    attn == 1,
+    !is.na(workerId)
+    ) %>%
+  rename(
+    Corruption_trial = corruption_trial,
+    Clerk_trial = clerk_trial,
+    Flattire_trial = flattire_trial,
+    Rape_trial = rape_trial) %>%
+  mutate(sex = if_else(sex == 1, "male", "female"))
+
+# normal = more likely is 1
+# reverse = less likely is 1
+
+reverse_items2 <- c(
+  "Cancer_trial",
+  "Heater_trial",
+  "Flattire_trial",
+  "Clerk_trial",
+  "Corruption_trial",
+  "Rape_trial",
+  "Earthquake_trial",
+  "Fastfood_trial",
+  "Police_trial"
+)
+
+d2_tidy <- d2_wide %>%
+  select(-Check1_trial, -Check2_trial, -Check3_trial) %>%
+  gather(trial, response, Tornado_trial:BigDiff5_trial) %>%
+  mutate(trial_type = ifelse(grepl("[0-9]", trial), "filler", "experimental")) %>%
+  mutate(trial_type = ifelse(grepl("BigDiff",trial), "filler-diff", trial_type)) %>%
+  mutate(likely_2nd = ifelse((trial %in% reverse_items2), 1, 0))
+
+d2 <- d2_tidy %>%
+  filter(trial_type == "experimental") %>%
+  mutate(response = ifelse(likely_2nd==1, response*-1, response))
+```
+
+```{r compute descriptives, include=FALSE}
+gender_exp1 <- d %>% 
+  group_by(workerId) %>% 
+  summarize(sex=first(sex)) %>% 
+  ungroup() %>% 
+  count(sex) %>% 
+  spread(sex, n)
+
+gender_exp2 <- d2 %>% 
+  group_by(workerId) %>% 
+  summarize(sex=first(sex)) %>% 
+  ungroup() %>% 
+  count(sex) %>% 
+  spread(sex, n)
+
+age_exp1 <- d %>% 
+  group_by(workerId) %>% 
+  summarize(age=first(age)) %>% 
+  ungroup() %>% 
+  summarize(age = median(age)) %>% 
+  .$age
+
+age_exp2 <- d2 %>% 
+  group_by(workerId) %>% 
+  summarize(age=first(age)) %>% 
+  ungroup() %>% 
+  summarize(age = median(age)) %>% 
+  .$age
+```
+
+```{r fit brms models, include=FALSE}
+library(brms)
+devtools::source_gist(id = "f1994c0f8325abbc5d300600744af39d", filename="cbrm.R")
+
+fit_exp1 <- cbrm(
+  response ~ 0 + intercept + trial + (1|workerId), # another model adds likely_2nd
+  data = d,
+  family = bernoulli(),
+  prior = prior(normal(0,3), class="b"),
+  control = list(adapt_delta = .80),
+  cores = parallel::detectCores(),
+  iter = 2000,
+  cached_file = "fit_exp1.rds"
+  )
+
+fit_exp2 <- cbrm(
+  response ~ trial + (1|workerId), # another model adds likely_2nd
+  data = d2 %>% mutate(response = response + 3),
+  family = cumulative(),
+  prior = prior(normal(0,3), class="Intercept"),
+  control = list(adapt_delta = .80),
+  cores = parallel::detectCores(),
+  sample_prior = TRUE,
+  iter = 2000,
+  cached_file = "fit_exp2.rds"
+  )
+```
+
+```{r hypothesis tests, include=FALSE}
+print_bf <- function(hypothesis) {
+
+  bf <- hypothesis$hypothesis$Evid.Ratio
+
+  if (is.na(bf)) {
+    return("< .001")
+  }
+  else if (bf < .001) {
+    return("< .001")
+  }
+  else if (bf > 1000) {
+    return("> 1000")
+  }
+  else {
+    return(paste("=", as.character(bf)))
+  }
+}
+
+coefs_exp1 <- broom::tidy(fit_exp1) %>%
+  filter(term=="b_intercept")
+
+# coefs_exp2 <- broom::tidy(fit_exp2) %>%
+#   filter(term=="b_intercept")
+
+bf_exp1 <- hypothesis(fit_exp1, "intercept = 0")
+# bf_exp1 <- h1$hypothesis$Evid.Ratio
+
+bf_exp2 <- hypothesis(fit_exp2, "Intercept[2] = -Intercept[3]")
+# bf_exp2 <- h2$hypothesis$Evid.Ratio
+```
+
+```{r make figures, include=FALSE}
+# plt <- plot(marginal_effects(fit_exp1))
+```
+
 # Introduction
 
-On the evening of November 13th, 2015, a terrorist attack in Paris left 130 people dead and injured over 300 more.
-In the aftermath, millions took to Twitter to express their shock, horror, and outrage at this tragedy under hashtags like `#parisattacks` and `#jesuisparis`.
-Yet, most of those mourning had little to say 15 hours earlier, when another tragic attack killed at least 43 people in Beirut.
-Several factors are surely at play in these different reactions  [e.g.
-group affiliations, @Brewer1999; construal-level theory, @Trope2010], yet one potentially fundamental factor has gone unmentioned: the fact that the Paris attack was more surprising than the attack in Beirut.
-In contrast to France, Lebanon had experienced dozens of terrorist bombings and attacks in recent years.
-Consequently, Beirut may seem to many like the sort of place where “these things happen,” whereas Paris is perceived as being stable and safe.
+When we learn of tragic events--a friend's loss of a loved-one, a violent killing, or a terrorist attack--we are mournful, concerned, angry, and driven to action.
+But our responses to these events differ depending on the context.
+In particular, it often seems as if our expectations, how surprising or unsurprising the event, play an important role.
+This seems true of both the mundane--When a friend's loved-one passes, we often ask, "was it sudden?"-- the dramatic--a murder in a sleepy town prompts a very different reaction than a similar crime in a major city--and even the catastrophic: acts of terrorism seem to evoke more outrage, horror, and concern when they occurs in peaceful rather than unstable regions.
+For instance, the 2015 Paris attacks were felt around the world.
+Yet, most of those mourning had little to say 15 hours earlier, when another tragic attack killed at least 43 people in Beirut (citation).
+Likewise, stories of the Virginia Tech massacre, which killed 32 people, wildly overshadowed coverage of attacks killing nearly 200 people in Iraq--one of the bloodiest days since the 2003 invasion (cite).
+In each of these situations, the surprise elicited by an event appears to affect evaluations.
+In some contexts, harm is tragic and in others, “these things happen.”
 
-We can see in everyday experience that people’s evaluations of events often depend on their expectations about those events.
-There is often disappointment when events fail to meet expectations, and there is a special thrill to having one’s expectations exceeded.
-Anecdotally, these forces seem to drive people’s tendencies to root for the underdog, hold surprise parties, and foreshadow bad news to ease its delivery [@Bell1985].
-Indeed, laboratory studies suggest that expectations play an important role in people’s evaluations of the utility of an event.
+Although this pattern seems so intuitive as to be obvious, there is no code of ethics that would justify this pattern of judgment, nor is there any moral psychological framework that can clearly explain it.
+(i think we could cite that kurt gray thing saying that weirdness is part of the response to purity violations).
+Outside of the domain of morality, however, a parallel set of everyday experiences and laboratory work shows that people’s evaluations of events unambiguously depend on their expectations about those events.
+There is disappointment when events fail to meet expectations, and there is a special thrill to having one’s expectations exceeded: we root for the underdog, hold surprise parties, and foreshadow bad news to ease its delivery [@Bell1985].
+Indeed, laboratory work suggests that expectations play an important role in people’s evaluations of the utility of an event.
 For instance, Mellers and colleagues [-@Mellers1997] found that expectations influenced affective reactions during a gambling task: Given a gamble with a 10% chance to win \$30 and 90% chance to win \$0, participants felt little disappointment with the \$0 outcome, but were considerably more excited when they won \$30.
@@
 
-Just as expectations affect the utility of simple gambles, we suspect they are equally likely to shape how people react to morally harmful events, such as acts of terrorism (Waldmann Cite and Rai and Fiske cite --domain general peeps; us also).
-However, unlike in the context of gambles, in these contexts the effects of expectations on evaluations may have harmful consequences.
-When events are shocking, we predict that people will perceive them as more severe and consequently be roused to action.
-In contrast, when events harm victims who are generally considered to be at greater risk--the poor, sick, or those living in unstable regions of the world, we predicted that reactions may be more muted.
-However, prior research has yet to articulate whether these same effects emerge in the moral domain and the implications of this possibility.
-For instance, it is unclear if these same mechanisms can help account for the divergent reactions people have to terrorism in France and Beirut.
-Further, people’s moral evaluations are in fact impacted by expectations what is their source and what does that tell us about how to overcome the biases they may impose?
+Expectations seem to affect people's reactions to personal gains and losses, but can these findings be extended to the types of moral situations we have highlighted thus far?
+That is, do expectations directly affect reactions to morally consequential events, as intuition and anecdote seem to suggest?
+And if so, why should expectations inform these reactions, and what does this say about the dependability of human moral judgments?
 
@@
 Much joy is tied to the psychological marriage between surprise and evaluation.
-However, this basic cognitive process may have negative moral repercussions.
+Indeed, it seems that for information reasons alone, attending to events that tell us the most seems like an efficient way to navigate the world.
+
+However, the basic cognitive process we have just discussed may have negative moral repercussions.
 When harms are expected, these dynamics may lead people to draw less extreme evaluations, feel less concern for victims, and be less driven to help.
 To test this hypothesis, we examined whether people’s evaluations of morally harmful events are affected by their expectations about those events.
-We asked people to compare pairs of simple events where a victim suffered an identical harm, but where the events differed in their prior probability.
+The information-theoretic account we have sketched makes only directional predictions--predicting that, all else equal, more surprising events will elicit stronger responses than less surprising events.
+Thus, we examined the effects of expectations in a comparison task, where we asked people to compare pairs of simple events where a victim suffered an identical harm, but where the events differed in their prior probability.
 For each pair of events, participants were asked to judge which event was more upsetting.
@@
 
+# Study 1
+
+Study 1 examined the hypothesis that changes in the prior probability of an event would affect people’s evaluation of that event.
+
+## Participants
+
+A total of `r n_distinct(d_raw$workerId)` participants were recruited from Amazon’s Mechanical Turk work distribution website (mTurk).
+Of these, `r n_distinct(d$workerId)` passed attention checks and were included in the final analyses (`r gender_exp1$male` male, `r gender_exp1$female` female, median age = `r age_exp1` years old).
+All participants were paid $0.60 for their participation.
+
+## Materials and procedure
+
+Participants judged 16 experimental event-pairs and 10 equivalent filler event-pairs.
+The events were described in the passive voice, and participants were asked to judge which event seemed more upsetting.
+For example:
+
+* A 32 year old woman gets food poisoning after eating a hamburger at a fast food restaurant. [Expected]
+
+* A 32 year old woman gets food poisoning after eating a hamburger at a four star restaurant. [Unexpected]
+
+On each trial, participants were presented with the event-pair stimulus and had to judge which outcome was more upsetting in a two-alternative forced choice task.
+The two events were labeled “Outcome 1” and “Outcome 2” and their order was randomized.
+
+## Results and discussion
+
+We predicted that people would think that events where unexpected harm occurred were more upsetting than events that entail similar harm but were comparatively more expected.
+As indicated in Figure 1, people judged that unexpected negative events were more upsetting than expected events.
+To confirm this difference formally, we fit a Bayesian logistic random effects model with participants’ responses as the dependent variable (1 = unexpected event is more upsetting; 0 = expected event is more upsetting) and random intercepts for item and subject, allowing us to generalize to items and subjects we did not test.
+The intercept in this model represents the log-odds of selecting the unexpected event as being more upsetting.
+Thus, by examining the population-level intercept, we can test whether participants were biased toward selecting the unexpected event ($\beta$ > 0), the expected event ($\beta$ < 0), or were unbiased ($\beta$ = 0).
+Consistent with our hypothesis, we found that people were much more likely to think that unexpected events were more upsetting than events that were expected, Intercept = `r round(coefs_exp1$estimate, 3)`, 95% CI [`r round(coefs_exp1$lower, 3)`, `r round(coefs_exp1$upper, 3)`], BF01 `r print_bf(bf_exp1)`.
+Bayes factors and the estimate of the intercept were similar under different prior choices.
+
+```{r figure1, echo=FALSE, fig.cap="Marginal probability of choosing more likely response as more upsetting in Experiment 1. (DRAFT FIGURE)"}
+
+plot(marginal_effects(fit_exp1, "trial"))
+```
+
+Figure 1 shows participants' responses broken-down by individual items. Participants' bias toward selecting the unexpected event as more upsetting was consistent across the 16 experimental items.
+
+# Study 2
+
+In Study 1 we found that people’s judgments of events were biased by their expectations about those events.
+When forced to choose between two events, participants decided that unexpected events were more upsetting than expected events.
+In Study 2, we sought to test our hypothesis using a more conservative method: we provided participants with a more expressive response scale so that if they viewed the events under consideration as equally harmful, their responses could reflect their attitude.
+
+## Participants
+
+A total of `r n_distinct(d2_raw$workerId)` participants were recruited from Amazon’s Mechanical Turk work distribution website (mTurk).
+Of these, `r n_distinct(d$workerId)` passed attention checks and were included in the final analyses (`r gender_exp2$male` male, `r gender_exp2$female` female, median age = `r age_exp2` years old).
+All participants were paid $0.60 for their participation.
+
+## Materials and procedure
+
+Participants judged 16 experimental event-pairs.
+One “equivalent” filler item was changed because participants did not, in fact, view the two outcomes as equally upsetting.
+(Specifically, participants thought it was more upsetting to have one’s right foot amputated than their left foot) [OSF LINK].
+
+As in Study 1, on each trial of the study, participants were presented with a pair of actions labeled “Outcome 1” and “Outcome 2” and were asked, “Which outcome seems more upsetting?” However, unlike previous studies, in Study 2 participants made their rating on a five-point scale (Outcome 1 seems more upsetting, Outcome 1 seems a little more upsetting, neither seems more upsetting than the other, Outcome 2 seems a little more upsetting, Outcome 2 seems more upsetting).
+
+## Results and discussion
+
+```{r figure2, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Marginal expected response (1-5 scale) in Experiment 2.
+A response of 3 indicates indifference, a response below 3 indicates a bias to choose the less likely response as more upsetting.
+(DRAFT FIGURE)"}
+
+plot(marginal_effects(fit_exp2, "trial"))
+```
+
+The events participants were asked to compare were, by design, highly similar.
+Consequently, we expected that participants' would typically indicate that neither event seemed more upsetting.
+Indeed, we think that this response is, by most people’s lights, the right answer.
+This was by far the choice participants most frequently made (see Figure 1, panel 3).
+However, we also observed that when participants did perceive one event was more upsetting than the other, they were biased to perceive unexpected negative events as more upsetting than more-expected negative events.
+
+To examine these findings formally, we performed cumulative (ordinal) regression using a Bayesian random effects model with participants’ scale responses as the dependent variable (1 to 5) and random intercepts for item and subject.
+This model produces four intercept coefficients, representing the cumulative log-odds of responses at each scale point or higher.
+For instance, the second coefficient represents the log-odds participants chose a 2 ("outcome 2 seems slightly more upsetting") or lower on the scale.
+Similarly, the third intercept coefficient represents the log-odds participants chose a 3 or lower on the scale.
+By comparing the second intercept coefficient to the inverse of the third intercept coefficient (thereby representing the log-odds *not* choosing a 3 or lower--i.e., choosing a 4 or 5), we can test whether participants were more likely to choose the expected or unexpected event as being more upsetting in cases where they did not choose the neither option.
+This analysis indicated that people were more likely to think that events that were unexpected were more upsetting than events that were expected, BF01 `r print_bf(bf_exp2)` (see supplemental online materials for full model results) -- when participants did exhibit a bias in their responses about which event was more upsetting, they reliably chose the unexpected event was more upsetting than the expected event.
+
+# Discussion
+
+On the evening of November 13th, 2015, a terrorist attack in Paris left 130 people dead and injured over 300 more.
+In the aftermath, millions took to Twitter to express their shock, horror, and outrage at this tragedy under hashtags like `#parisattacks` and `#jesuisparis`.
+Yet, most of those mourning had little to say 15 hours earlier, when another tragic attack killed at least 43 people in Beirut.
+Several factors are surely at play in these different reactions  [e.g.
+group affiliations, @Brewer1999; construal-level theory, @Trope2010], yet one potentially fundamental factor has gone unmentioned: the fact that the Paris attack was more surprising than the attack in Beirut.
+In contrast to France, Lebanon had experienced dozens of terrorist bombings and attacks in recent years.
+Consequently, Beirut may seem to many like the sort of place where “these things happen,” whereas Paris is perceived as being stable and safe.
+
+Consistent with this hypothesis, the results of two preregistered studies suggest that people view unexpected harmful events as more upsetting than expected harmful events.
+Just as people react more strongly to unexpected monetary gains and losses [@Mellers1997], people similarly are more upset by unexpected moral harm than expected moral harm.
+However, unlike in the context of gambles, in these contexts the effects of expectations on evaluations may have harmful consequences.
+When events are shocking, people may perceive them as more severe and consequently be roused to action.
+In contrast, when events harm victims who are generally considered to be at greater risk--the poor, sick, or those living in unstable regions of the world, reactions are likely to be more muted.
+
+What is the source of this bias: prior judgment and decision-making research along with Information Theory suggests that  these effects are unlikely to be a domain-specific quirk of moral thinking and more likely to be a product of information-theoretic mechanisms.
+The striking consequence of this is that it is quite likely that the bias to ignore expected harm exists because it allows us to efficiently navigation of the world around us; paying attention to events that surprise us allows us to learn the most about the world.
+But the consequences of this bias seems unambiguously sub-optimal.
+
+Understood in this way, the inconsistency in people’s attitudes about otherwise similar acts of terrorism seems almost intuitive.
+An act of terrorism in a peaceful nation might be less expected because of our beliefs about the effectiveness of security forces, the stability of national relationships, and positive relationships among people groups in that nation.
+In contrast, a surprising act of terrorism casts doubt on our deeper understanding of the geopolitical situation, and thereby leads us to suspect similar acts are more likely to occur in the future.
+This could rightly increase our concern about this event.
+
+However, this bias--regardless of whether it is rooted in otherwise reasonable information-theoretic mechanisms--threatens to impose a vicious and morally pernicious cycle.
+For instance, people living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and are the very people in greatest need of assistance and concern from the world at-large.
+However, for these very reasons, it is often unsurprising when harm befalls people living in these circumstances.
+Stated another way, expected but nonetheless devastating events elicit little concern, and thus cause people to be less likely to donate time or money to aid victims, or to take political action.
+Understanding the mechanism that leads people to exhibit a bias where they will be least likely to help those in most need of attention can guide policymakers and psychologists to develop interventions which allow people to overcome this default tendency.
+
+The information-theoretic account has further implications, including offering a unifying explanation for related moral judgment behavior.
+For instance, the account we’ve discussed can elucidate why people appear to distinguish between acts of omission (e.g., failing to save someone’s life) and commission (e.g., actively killing someone).
+Most people feel (Spranca, Minsk, & Baron, 1991), and many ethicists argue (e.g., Foot, 1967; Kagan, 1988; Kamm, 1993; Kant 1780/1965; Thomson, 1986), that immoral acts of commission may seem like more severe moral violations than are acts of omission that produce similar outcomes.
+Our account may shed light on this phenomena: when a person is in a position to be harmed through another’s inaction, that person tends to be at greater risk of harm than a person who can only be harmed through direct action.
+Therefore, omission generally leads to a smaller change in moral utility than do acts of commission, potentially explaining its less severe moral implications.
+
+To consider one more example, our account may provide an explanation of certain types of victim blaming.
+Often, female victims of sexual harassment and assault are held partially responsible for the actions of their male aggressors.
+This is especially true if they were dressed or acting promiscuously, or if they willfully became intoxicated prior to the assault.
+People seemingly believe that a woman who behaves in these ways increases the probability that she will suffer sexual assault.
+Research by Janoff-Bullman, Timko, and Carli (1984) indicates that people are more likely to hold victims accountable for their own rape when the rape was perceived as having a greater prior probability, even when a victim’s behaviors were the same.
+Not only are victims sometimes held accountable for their own assaults, but victim blaming also serves to shift blame away from the aggressor, leading some people to partially or completely excuse their transgressions.
+Although these counter-normative attitudes are often extrinsically motivated by misogyny (cite), our account may explain some aspects of this tendency.
+When people perceive a victims’ actions as increasing the probability of their eventual assault, this affects their utility evaluations of an aggressor’s actions: an aggressor’s actions produce a smaller change in utility, thus warranting weaker condemnation.
+
+# Acknowledgements
+
+We would like to acknowledge the help of Larisa Hussak, Keith Holyoak, Alan Fiske, Matthew Lieberman, Hongjing Lu, John Hummel, Andrei Cimpian, and Ellen Markman for their comments and support.
+
 # Appendix
@@
 
-\setlength{\parindent}{-0.25in} 
+\setlength{\parindent}{-0.125in} 
 \setlength{\leftskip}{0.125in}

### 2018-06-04 paper/paper.tex
updated paper with results

@@
-\documentclass[]{article}
+\documentclass[12pt,]{article}
 \usepackage{lmodern}
@@
 
+\usepackage{fancyhdr}
+\pagestyle{fancy}
+\fancyhead{}
+\fancyhead[LO,LE]{PREPRINT DRAFT - PLEASE DO NOT CITE WITHOUT PERMISSION}
+\fancyfoot[CO,CE]{}
+
 \begin{document}
@@
 
-On the evening of November 13th, 2015, a terrorist attack in Paris left
-130 people dead and injured over 300 more. In the aftermath, millions
-took to Twitter to express their shock, horror, and outrage at this
-tragedy under hashtags like \texttt{\#parisattacks} and
-\texttt{\#jesuisparis}. Yet, most of those mourning had little to say 15
+When we learn of tragic events--a friend's loss of a loved-one, a
+violent killing, or a terrorist attack--we are mournful, concerned,
+angry, and driven to action. But our responses to these events differ
+depending on the context. In particular, it often seems as if our
+expectations, how surprising or unsurprising the event, play an
+important role. This seems true of both the mundane--When a friend's
+loved-one passes, we often ask, ``was it sudden?''-- the dramatic--a
+murder in a sleepy town prompts a very different reaction than a similar
+crime in a major city--and even the catastrophic: acts of terrorism seem
+to evoke more outrage, horror, and concern when they occurs in peaceful
+rather than unstable regions. For instance, the 2015 Paris attacks were
+felt around the world. Yet, most of those mourning had little to say 15
 hours earlier, when another tragic attack killed at least 43 people in
-Beirut. Several factors are surely at play in these different reactions
-(construal-level theory, {\textbf{???}}; e.g.~group affiliations,
-Brewer, 1999), yet one potentially fundamental factor has gone
-unmentioned: the fact that the Paris attack was more surprising than the
-attack in Beirut. In contrast to France, Lebanon had experienced dozens
-of terrorist bombings and attacks in recent years. Consequently, Beirut
-may seem to many like the sort of place where ``these things happen,''
-whereas Paris is perceived as being stable and safe.
+Beirut (citation). Likewise, stories of the Virginia Tech massacre,
+which killed 32 people, wildly overshadowed coverage of attacks killing
+nearly 200 people in Iraq--one of the bloodiest days since the 2003
+invasion (cite). In each of these situations, the surprise elicited by
+an event appears to affect evaluations. In some contexts, harm is tragic
+and in others, ``these things happen.''
 
-We can see in everyday experience that people's evaluations of events
-often depend on their expectations about those events. There is often
-disappointment when events fail to meet expectations, and there is a
-special thrill to having one's expectations exceeded. Anecdotally, these
-forces seem to drive people's tendencies to root for the underdog, hold
-surprise parties, and foreshadow bad news to ease its delivery (Bell,
-1985). Indeed, laboratory studies suggest that expectations play an
-important role in people's evaluations of the utility of an event. For
-instance, Mellers and colleagues (1997) found that expectations
-influenced affective reactions during a gambling task: Given a gamble
-with a 10\% chance to win \$30 and 90\% chance to win \$0, participants
-felt little disappointment with the \$0 outcome, but were considerably
-more excited when they won \$30. Conversely, given a gamble with a 90\%
-chance of winning \$60 and 10\% chance of winning \$0, participants were
+Although this pattern seems so intuitive as to be obvious, there is no
+code of ethics that would justify this pattern of judgment, nor is there
+any moral psychological framework that can clearly explain it. (i think
+we could cite that kurt gray thing saying that weirdness is part of the
+response to purity violations). Outside of the domain of morality,
+however, a parallel set of everyday experiences and laboratory work
+shows that people's evaluations of events unambiguously depend on their
+expectations about those events. There is disappointment when events
+fail to meet expectations, and there is a special thrill to having one's
+expectations exceeded: we root for the underdog, hold surprise parties,
+and foreshadow bad news to ease its delivery (Bell, 1985). Indeed,
+laboratory work suggests that expectations play an important role in
+people's evaluations of the utility of an event. For instance, Mellers
+and colleagues (1997) found that expectations influenced affective
+reactions during a gambling task: Given a gamble with a 10\% chance to
+win \$30 and 90\% chance to win \$0, participants felt little
+disappointment with the \$0 outcome, but were considerably more excited
+when they won \$30. Conversely, given a gamble with a 90\% chance of
+winning \$60 and 10\% chance of winning \$0, participants were
 disappointed with the \$0 outcome and showed more muted enjoyment of the
@@
 
-Just as expectations affect the utility of simple gambles, we suspect
-they are equally likely to shape how people react to morally harmful
-events, such as acts of terrorism (Waldmann Cite and Rai and Fiske cite
---domain general peeps; us also). However, unlike in the context of
-gambles, in these contexts the effects of expectations on evaluations
-may have harmful consequences. When events are shocking, we predict that
-people will perceive them as more severe and consequently be roused to
-action. In contrast, when events harm victims who are generally
-considered to be at greater risk--the poor, sick, or those living in
-unstable regions of the world, we predicted that reactions may be more
-muted. However, prior research has yet to articulate whether these same
-effects emerge in the moral domain and the implications of this
-possibility. For instance, it is unclear if these same mechanisms can
-help account for the divergent reactions people have to terrorism in
-France and Beirut. Further, people's moral evaluations are in fact
-impacted by expectations what is their source and what does that tell us
-about how to overcome the biases they may impose?
+Expectations seem to affect people's reactions to personal gains and
+losses, but can these findings be extended to the types of moral
+situations we have highlighted thus far? That is, do expectations
+directly affect reactions to morally consequential events, as intuition
+and anecdote seem to suggest? And if so, why should expectations inform
+these reactions, and what does this say about the dependability of human
+moral judgments?
 
@@
 Much joy is tied to the psychological marriage between surprise and
-evaluation. However, this basic cognitive process may have negative
-moral repercussions. When harms are expected, these dynamics may lead
-people to draw less extreme evaluations, feel less concern for victims,
-and be less driven to help. To test this hypothesis, we examined whether
-people's evaluations of morally harmful events are affected by their
-expectations about those events. We asked people to compare pairs of
-simple events where a victim suffered an identical harm, but where the
-events differed in their prior probability. For each pair of events,
-participants were asked to judge which event was more upsetting. Two
-studies provided evidence for the predicted effect of expectations on
-moral evaluations: people tended to view unexpected negative events as
-more upsetting, even when the harm to victims was identical.
+evaluation. Indeed, it seems that for information reasons alone,
+attending to events that tell us the most seems like an efficient way to
+navigate the world.
+
+However, the basic cognitive process we have just discussed may have
+negative moral repercussions. When harms are expected, these dynamics
+may lead people to draw less extreme evaluations, feel less concern for
+victims, and be less driven to help. To test this hypothesis, we
+examined whether people's evaluations of morally harmful events are
+affected by their expectations about those events. The
+information-theoretic account we have sketched makes only directional
+predictions--predicting that, all else equal, more surprising events
+will elicit stronger responses than less surprising events. Thus, we
+examined the effects of expectations in a comparison task, where we
+asked people to compare pairs of simple events where a victim suffered
+an identical harm, but where the events differed in their prior
+probability. For each pair of events, participants were asked to judge
+which event was more upsetting. Two studies provided evidence for the
+predicted effect of expectations on moral evaluations: people tended to
+view unexpected negative events as more upsetting, even when the harm to
+victims was identical.
 
@@
 
+\section{Study 1}\label{study-1}
+
+Study 1 examined the hypothesis that changes in the prior probability of
+an event would affect people's evaluation of that event.
+
+\subsection{Participants}\label{participants}
+
+A total of 255 participants were recruited from Amazon's Mechanical Turk
+work distribution website (mTurk). Of these, 230 passed attention checks
+and were included in the final analyses (105 male, 125 female, median
+age = 33 years old). All participants were paid \$0.60 for their
+participation.
+
+\subsection{Materials and procedure}\label{materials-and-procedure}
+
+Participants judged 16 experimental event-pairs and 10 equivalent filler
+event-pairs. The events were described in the passive voice, and
+participants were asked to judge which event seemed more upsetting. For
+example:
+
+\begin{itemize}
+\item
+  A 32 year old woman gets food poisoning after eating a hamburger at a
+  fast food restaurant. {[}Expected{]}
+\item
+  A 32 year old woman gets food poisoning after eating a hamburger at a
+  four star restaurant. {[}Unexpected{]}
+\end{itemize}
+
+On each trial, participants were presented with the event-pair stimulus
+and had to judge which outcome was more upsetting in a two-alternative
+forced choice task. The two events were labeled ``Outcome 1'' and
+``Outcome 2'' and their order was randomized.
+
+\subsection{Results and discussion}\label{results-and-discussion}
+
+We predicted that people would think that events where unexpected harm
+occurred were more upsetting than events that entail similar harm but
+were comparatively more expected. As indicated in Figure 1, people
+judged that unexpected negative events were more upsetting than expected
+events. To confirm this difference formally, we fit a Bayesian logistic
+random effects model with participants' responses as the dependent
+variable (1 = unexpected event is more upsetting; 0 = expected event is
+more upsetting) and random intercepts for item and subject, allowing us
+to generalize to items and subjects we did not test. The intercept in
+this model represents the log-odds of selecting the unexpected event as
+being more upsetting. Thus, by examining the population-level intercept,
+we can test whether participants were biased toward selecting the
+unexpected event (\(\beta\) \textgreater{} 0), the expected event
+(\(\beta\) \textless{} 0), or were unbiased (\(\beta\) = 0). Consistent
+with our hypothesis, we found that people were much more likely to think
+that unexpected events were more upsetting than events that were
+expected, Intercept = -0.647, 95\% CI {[}-0.923, -0.378{]}, BF01
+\textless{} .001. Bayes factors and the estimate of the intercept were
+similar under different prior choices.
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/figure1-1.pdf}
+\caption{Marginal probability of choosing more likely response as more
+upsetting in Experiment 1. (DRAFT FIGURE)}
+\end{figure}
+
+Figure 1 shows participants' responses broken-down by individual items.
+Participants' bias toward selecting the unexpected event as more
+upsetting was consistent across the 16 experimental items.
+
+\section{Study 2}\label{study-2}
+
+In Study 1 we found that people's judgments of events were biased by
+their expectations about those events. When forced to choose between two
+events, participants decided that unexpected events were more upsetting
+than expected events. In Study 2, we sought to test our hypothesis using
+a more conservative method: we provided participants with a more
+expressive response scale so that if they viewed the events under
+consideration as equally harmful, their responses could reflect their
+attitude.
+
+\subsection{Participants}\label{participants-1}
+
+A total of 504 participants were recruited from Amazon's Mechanical Turk
+work distribution website (mTurk). Of these, 230 passed attention checks
+and were included in the final analyses (174 male, 199 female, median
+age = 34 years old). All participants were paid \$0.60 for their
+participation.
+
+\subsection{Materials and procedure}\label{materials-and-procedure-1}
+
+Participants judged 16 experimental event-pairs. One ``equivalent''
+filler item was changed because participants did not, in fact, view the
+two outcomes as equally upsetting. (Specifically, participants thought
+it was more upsetting to have one's right foot amputated than their left
+foot) {[}OSF LINK{]}.
+
+As in Study 1, on each trial of the study, participants were presented
+with a pair of actions labeled ``Outcome 1'' and ``Outcome 2'' and were
+asked, ``Which outcome seems more upsetting?'' However, unlike previous
+studies, in Study 2 participants made their rating on a five-point scale
+(Outcome 1 seems more upsetting, Outcome 1 seems a little more
+upsetting, neither seems more upsetting than the other, Outcome 2 seems
+a little more upsetting, Outcome 2 seems more upsetting).
+
+\subsection{Results and discussion}\label{results-and-discussion-1}
+
+\begin{figure}
+\centering
+\includegraphics{paper_files/figure-latex/figure2-1.pdf}
+\caption{Marginal expected response (1-5 scale) in Experiment 2. A
+response of 3 indicates indifference, a response below 3 indicates a
+bias to choose the less likely response as more upsetting. (DRAFT
+FIGURE)}
+\end{figure}
+
+The events participants were asked to compare were, by design, highly
+similar. Consequently, we expected that participants' would typically
+indicate that neither event seemed more upsetting. Indeed, we think that
+this response is, by most people's lights, the right answer. This was by
+far the choice participants most frequently made (see Figure 1, panel
+3). However, we also observed that when participants did perceive one
+event was more upsetting than the other, they were biased to perceive
+unexpected negative events as more upsetting than more-expected negative
+events.
+
+To examine these findings formally, we performed cumulative (ordinal)
+regression using a Bayesian random effects model with participants'
+scale responses as the dependent variable (1 to 5) and random intercepts
+for item and subject. This model produces four intercept coefficients,
+representing the cumulative log-odds of responses at each scale point or
+higher. For instance, the second coefficient represents the log-odds
+participants chose a 2 (``outcome 2 seems slightly more upsetting'') or
+lower on the scale. Similarly, the third intercept coefficient
+represents the log-odds participants chose a 3 or lower on the scale. By
+comparing the second intercept coefficient to the inverse of the third
+intercept coefficient (thereby representing the log-odds \emph{not}
+choosing a 3 or lower--i.e., choosing a 4 or 5), we can test whether
+participants were more likely to choose the expected or unexpected event
+as being more upsetting in cases where they did not choose the neither
+option. This analysis indicated that people were more likely to think
+that events that were unexpected were more upsetting than events that
+were expected, BF01 \textless{} .001 (see supplemental online materials
+for full model results) -- when participants did exhibit a bias in their
+responses about which event was more upsetting, they reliably chose the
+unexpected event was more upsetting than the expected event.
+
+\section{Discussion}\label{discussion}
+
+On the evening of November 13th, 2015, a terrorist attack in Paris left
+130 people dead and injured over 300 more. In the aftermath, millions
+took to Twitter to express their shock, horror, and outrage at this
+tragedy under hashtags like \texttt{\#parisattacks} and
+\texttt{\#jesuisparis}. Yet, most of those mourning had little to say 15
+hours earlier, when another tragic attack killed at least 43 people in
+Beirut. Several factors are surely at play in these different reactions
+(e.g.~group affiliations, Brewer, 1999; construal-level theory, Trope \&
+Liberman, 2010), yet one potentially fundamental factor has gone
+unmentioned: the fact that the Paris attack was more surprising than the
+attack in Beirut. In contrast to France, Lebanon had experienced dozens
+of terrorist bombings and attacks in recent years. Consequently, Beirut
+may seem to many like the sort of place where ``these things happen,''
+whereas Paris is perceived as being stable and safe.
+
+Consistent with this hypothesis, the results of two preregistered
+studies suggest that people view unexpected harmful events as more
+upsetting than expected harmful events. Just as people react more
+strongly to unexpected monetary gains and losses (Mellers et al., 1997),
+people similarly are more upset by unexpected moral harm than expected
+moral harm. However, unlike in the context of gambles, in these contexts
+the effects of expectations on evaluations may have harmful
+consequences. When events are shocking, people may perceive them as more
+severe and consequently be roused to action. In contrast, when events
+harm victims who are generally considered to be at greater risk--the
+poor, sick, or those living in unstable regions of the world, reactions
+are likely to be more muted.
+
+What is the source of this bias: prior judgment and decision-making
+research along with Information Theory suggests that these effects are
+unlikely to be a domain-specific quirk of moral thinking and more likely
+to be a product of information-theoretic mechanisms. The striking
+consequence of this is that it is quite likely that the bias to ignore
+expected harm exists because it allows us to efficiently navigation of
+the world around us; paying attention to events that surprise us allows
+us to learn the most about the world. But the consequences of this bias
+seems unambiguously sub-optimal.
+
+Understood in this way, the inconsistency in people's attitudes about
+otherwise similar acts of terrorism seems almost intuitive. An act of
+terrorism in a peaceful nation might be less expected because of our
+beliefs about the effectiveness of security forces, the stability of
+national relationships, and positive relationships among people groups
+in that nation. In contrast, a surprising act of terrorism casts doubt
+on our deeper understanding of the geopolitical situation, and thereby
+leads us to suspect similar acts are more likely to occur in the future.
+This could rightly increase our concern about this event.
+
+However, this bias--regardless of whether it is rooted in otherwise
+reasonable information-theoretic mechanisms--threatens to impose a
+vicious and morally pernicious cycle. For instance, people living in
+geo-politically unstable regions or in the developing world are often
+those who are most affected by terrorism, famine, and natural disasters,
+and are the very people in greatest need of assistance and concern from
+the world at-large. However, for these very reasons, it is often
+unsurprising when harm befalls people living in these circumstances.
+Stated another way, expected but nonetheless devastating events elicit
+little concern, and thus cause people to be less likely to donate time
+or money to aid victims, or to take political action. Understanding the
+mechanism that leads people to exhibit a bias where they will be least
+likely to help those in most need of attention can guide policymakers
+and psychologists to develop interventions which allow people to
+overcome this default tendency.
+
+The information-theoretic account has further implications, including
+offering a unifying explanation for related moral judgment behavior. For
+instance, the account we've discussed can elucidate why people appear to
+distinguish between acts of omission (e.g., failing to save someone's
+life) and commission (e.g., actively killing someone). Most people feel
+(Spranca, Minsk, \& Baron, 1991), and many ethicists argue (e.g., Foot,
+1967; Kagan, 1988; Kamm, 1993; Kant 1780/1965; Thomson, 1986), that
+immoral acts of commission may seem like more severe moral violations
+than are acts of omission that produce similar outcomes. Our account may
+shed light on this phenomena: when a person is in a position to be
+harmed through another's inaction, that person tends to be at greater
+risk of harm than a person who can only be harmed through direct action.
+Therefore, omission generally leads to a smaller change in moral utility
+than do acts of commission, potentially explaining its less severe moral
+implications.
+
+To consider one more example, our account may provide an explanation of
+certain types of victim blaming. Often, female victims of sexual
+harassment and assault are held partially responsible for the actions of
+their male aggressors. This is especially true if they were dressed or
+acting promiscuously, or if they willfully became intoxicated prior to
+the assault. People seemingly believe that a woman who behaves in these
+ways increases the probability that she will suffer sexual assault.
+Research by Janoff-Bullman, Timko, and Carli (1984) indicates that
+people are more likely to hold victims accountable for their own rape
+when the rape was perceived as having a greater prior probability, even
+when a victim's behaviors were the same. Not only are victims sometimes
+held accountable for their own assaults, but victim blaming also serves
+to shift blame away from the aggressor, leading some people to partially
+or completely excuse their transgressions. Although these
+counter-normative attitudes are often extrinsically motivated by
+misogyny (cite), our account may explain some aspects of this tendency.
+When people perceive a victims' actions as increasing the probability of
+their eventual assault, this affects their utility evaluations of an
+aggressor's actions: an aggressor's actions produce a smaller change in
+utility, thus warranting weaker condemnation.
+
+\section{Acknowledgements}\label{acknowledgements}
+
+We would like to acknowledge the help of Larisa Hussak, Keith Holyoak,
+Alan Fiske, Matthew Lieberman, Hongjing Lu, John Hummel, Andrei Cimpian,
+and Ellen Markman for their comments and support.
+
 \section{Appendix}\label{appendix}
@@
 
-\setlength{\parindent}{-0.25in} \setlength{\leftskip}{0.125in} \noindent
+\setlength{\parindent}{-0.125in} \setlength{\leftskip}{0.125in}
+\noindent
 
@@
 
+\hypertarget{ref-Trope2010}{}
+Trope, Y., \& Liberman, N. (2010). Construal-level theory of
+psychological distance. \emph{Psychological Review}, \emph{117}(2),
+440--63. \url{http://doi.org/10.1037/a0018963}
+
 \hypertarget{ref-Tversky1992}{}

### 2018-06-04 paper/paper.Rmd
made draft figures readable

@@
 ```{r make figures, include=FALSE}
-# plt <- plot(marginal_effects(fit_exp1))
+fig1 <- plot(marginal_effects(fit_exp1, "trial"))
+fig2 <- plot(marginal_effects(fit_exp2, "trial"))
 ```
@@
 
-plot(marginal_effects(fit_exp1, "trial"))
+fig1$trial +
+  coord_cartesian(ylim=c(0,1)) +
+  geom_hline(yintercept = .5, linetype = "dashed") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1))
 ```
@@
 (DRAFT FIGURE)"}
-
-plot(marginal_effects(fit_exp2, "trial"))
+fig2$trial +
+  coord_cartesian(ylim=c(2.25,3.25)) +
+  geom_hline(yintercept = 3, linetype = "dashed") +
+  theme_bw() +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1))
 ```

### 2018-06-29 norming.Rmd
added norming data

---
title: "Item Norming"
output: html_notebook
---

```{r}
library(tidyverse)

df_raw <- read.csv("data/WIW_Norming_for_preregistered_study.csv")

reverse_items <- c(
  "Earthquake",
  "Tornado",
  "Fastfood",
  "Marine",
  "Police",
  "Mugging",
  "Crushed",
  "Dysentery",
  "Drunk",
  "Football"
)

df_wide <- df_raw %>%
  filter(Status==0) %>%
  filter(Check_trial == 5) %>%
  select(-Check_trial) %>%
  mutate(sex = if_else(sex == 1, "male", "female"))

df <- df_wide %>%
  gather(trial, response, Earthquake:Rape) %>%
  mutate(response = ifelse(trial %in% reverse_items, 6-response, response))
```

As predicted, in all items the "less expected" outcome was judged to be less likely. Plot below shows mean likelihood rating by item along with approximate 95% confidence intervals.

```{r}
# standard error function
stderr <- function(x) {
          sqrt(var(x[!is.na(x)]) / length(x[!is.na(x)]))
}

df %>%
  group_by(trial) %>%
  summarize(
    Mean = mean(response),
    UL = mean(response) + 1.96*stderr(response),
    LL = mean(response) - 1.96*stderr(response)
    ) %>%
  ggplot() +
  aes(x = reorder(trial, -Mean), y = Mean, ymin = LL, ymax=UL) +
  geom_pointrange() +
  geom_hline(yintercept = 3, linetype="dashed") +
  coord_cartesian(ylim=c(1,5)) +
  theme_bw()+
  labs(x = "Trial", y="Mean Likelihoods Judgment") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

### 2018-06-29 paper/paper.Rmd
added norming data

--- 2018-06-02 paper/paper.Rmd
@@
 ---
-title: "Expectations bias moral evaluations"
-author: "Derek Powell and Zachary Horne"
-output: 
-  pdf_document:
-    keep_tex: TRUE
-bibliography: library.bib
-csl: apa6.csl
+title: Expectations bias moral evaluations
 
+# Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
+author:
+  - name: Derek Powell
+    affiliation: a,1
+  - name: Zachary Horne
+    affiliation: b
+address:
+  - code: a
+    address: Stanford University, Department of Psychology, 450 Serra Mall, Stanford, CA, 94305
+  - code: b
+    address: Arizona State University, School of Social and Behavioral Sciences, 13591 N 47th Ave, Phoenix, AZ 85051
+
+corresponding_author:
+  - code: 1
+    text: "To whom correspondence should be addressed. E-mail: derekpowell@stanford.edu"
+
+# For footer text
+lead_author_surname: Powell
+
+## Remove this if not required - talk to zach 
+# equal_authors:
+#   - code: 0
+#     text:
+
+author_contributions: |
+  Both authors jointly conceived of the project and theory, designed and conducted experiments, and analyzed results. D.P. developed the formal aspects of the theory.
+
+## Remove this if not required
+conflict_of_interest: |
+  The authors declare no conflicts of interest.
+
+abstract: |
+ When we learn of tragic events--a violent killing or a terrorist attack--we are mournful and often driven to action.
+But our responses to these events can differ substantially depending on the context of the event.
+When tragedy takes us by surprise, we are shocked and outraged, but when it strikes in unstable regions or rough neighborhoods, we are often resigned to feel that “these things happen.” Here we explain the role that expectations seem to play in moral evaluation by modeling evaluations as the comparison of the state of the world before and after an event.
+According to this model, expectations influence evaluations through information-theoretic principles: less expected events do more to inform us about the state of the world than do more expected events.
+Consistent with this theory, in two preregistered studies we found that people’s judgments of morally significant events were affected by the prior probability of that event, so that people were more upset about an event that was unexpected (e.g., a robbery at a clothing store) than an event that was more more probable (e.g., a robbery at a convenience store).
+This bias threatens to impose a pernicious cycle: People living in poverty, geo-politically unstable regions, and other dangerous circumstances have the greatest need for assistance and concern.
+However, for these very reasons, it is often unsurprising when they are harmed.
+Consequently, the influence of expectations on moral evaluations threatens to reduce observer's concern for those people who are most in need.
+
+significance: |
+  People living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and thus are the very people in greatest need of assistance and concern from the world at-large.
+However, for these very reasons, it is often unsurprising when harm befalls them.
+Two experiments confirm this intuition: we found that people's expectations bias their reactions to morally-significant events--people are more upset by surprising harms than more expected harms.
+This bias threatens to reduce observer's concern for victims most in need, making people less likely to donate time or money to aid victims, or to take political action to prevent further harms.
+
+acknowledgements: |
+  We would like to acknowledge the help of Larisa Hussak, Keith Holyoak, Alan Fiske, Hongjing Lu, John Hummel, Andrei Cimpian, and Ellen Markman for their comments and support.
+
+keywords:
+  - evaluation
+  - moral judgment
+  - information theory
+  # - optional
+  # - optional
+
+## must be one of: pnasresearcharticle (usual two-column layout), pnasmathematics (one column layout), or pnasinvited (invited submissions only)
+pnas_type: pnasresearcharticle
+
+bibliography: wiw-biblio.bib
+csl: pnas.csl
+
+## change to true to add optional line numbering
+lineno: TRUE
+
+output: rticles::pnas_article
 ---
@@
 
-# Introduction
+```{r load data, include=FALSE}
+# analysis
 
-On the evening of November 13th, 2015, a terrorist attack in Paris left 130 people dead and injured over 300 more.
-In the aftermath, millions took to Twitter to express their shock, horror, and outrage at this tragedy under hashtags like `#parisattacks` and `#jesuisparis`.
-Yet, most of those mourning had little to say 15 hours earlier, when another tragic attack killed at least 43 people in Beirut.
-Several factors are surely at play in these different reactions  [e.g.
-group affiliations, @Brewer1999; construal-level theory, @Trope2010], yet one potentially fundamental factor has gone unmentioned: the fact that the Paris attack was more surprising than the attack in Beirut.
-In contrast to France, Lebanon had experienced dozens of terrorist bombings and attacks in recent years.
-Consequently, Beirut may seem to many like the sort of place where “these things happen,” whereas Paris is perceived as being stable and safe.
+library(tidyverse)
 
-We can see in everyday experience that people’s evaluations of events often depend on their expectations about those events.
-There is often disappointment when events fail to meet expectations, and there is a special thrill to having one’s expectations exceeded.
-Anecdotally, these forces seem to drive people’s tendencies to root for the underdog, hold surprise parties, and foreshadow bad news to ease its delivery [@Bell1985].
-Indeed, laboratory studies suggest that expectations play an important role in people’s evaluations of the utility of an event.
-For instance, Mellers and colleagues [-@Mellers1997] found that expectations influenced affective reactions during a gambling task: Given a gamble with a 10% chance to win \$30 and 90% chance to win \$0, participants felt little disappointment with the \$0 outcome, but were considerably more excited when they won \$30.
+d_raw <- read.csv("../data/WIW+Judgment+Task+Preregistered_March+30%2C+2018_11.17.csv")
+
+d_wide <- d_raw %>%
+  filter(DistributionChannel=="anonymous") %>%
+  rename(duration = Duration..in.seconds.) %>%
+  select(workerId, age, sex, attn, duration, contains("_trial"), -contains("_trial_")) %>%
+  filter(Check_trial == 0, attn == 1) %>%
+  rename(
+    Corruption_trial = corruption_trial,
+    Clerk_trial = clerk_trial,
+    Flattire_trial = flattire_trial,
+    Rape_trial = rape_trial) %>%
+  mutate(sex = if_else(sex == 1, "male", "female"))
+
+reverse_items <- c( # sadly, only 6 of 10 items were reversed! :(
+  "Cancer_trial",
+  "Heater_trial",
+  "Flattire_trial",
+  "Clerk_trial",
+  "Corruption_trial",
+  "Rape_trial"
+)
+
+d_tidy <- d_wide %>%
+  gather(trial, response, Earthquake_trial:BigDiff5_trial) %>%
+  mutate(trial_type = ifelse(grepl("[0-9]", trial), "filler", "experimental")) %>%
+  mutate(trial_type = ifelse(grepl("BigDiff",trial), "filler-diff", trial_type)) %>%
+  mutate(likely_2nd = ifelse((trial %in% reverse_items), 1, 0))
+
+d <- d_tidy %>%
+  filter(trial_type == "experimental", trial != "Check_trial") %>%
+  mutate(response = ifelse(likely_2nd==1, (1-response), response)) %>%
+  mutate(response = abs(response-1)) %>%
+  mutate(trial = gsub("_trial","",trial))
+
+d2_raw <- read.csv("../data/WIW+Judgment+Task+Preregistered+-+Likert_April+20%2C+2018_17.08.csv")
+
+d2_wide <- d2_raw %>%
+  filter(DistributionChannel=="anonymous") %>%
+  rename(duration = Duration..in.seconds.) %>%
+  select(workerId, age, sex, attn, duration, contains("_trial"), -contains("_trial_")) %>%
+  filter(
+    Check1_trial == -1, 
+    Check2_trial == -2, 
+    Check3_trial == 1, 
+    attn == 1,
+    !is.na(workerId)
+    ) %>%
+  rename(
+    Corruption_trial = corruption_trial,
+    Clerk_trial = clerk_trial,
+    Flattire_trial = flattire_trial,
+    Rape_trial = rape_trial) %>%
+  mutate(sex = if_else(sex == 1, "male", "female"))
+
+# normal = more likely is 1
+# reverse = less likely is 1
+
+reverse_items2 <- c(
+  "Cancer_trial",
+  "Heater_trial",
+  "Flattire_trial",
+  "Clerk_trial",
+  "Corruption_trial",
+  "Rape_trial",
+  "Earthquake_trial",
+  "Fastfood_trial",
+  "Police_trial"
+)
+
+d2_tidy <- d2_wide %>%
+  select(-Check1_trial, -Check2_trial, -Check3_trial) %>%
+  gather(trial, response, Tornado_trial:BigDiff5_trial) %>%
+  mutate(trial_type = ifelse(grepl("[0-9]", trial), "filler", "experimental")) %>%
+  mutate(trial_type = ifelse(grepl("BigDiff",trial), "filler-diff", trial_type)) %>%
+  mutate(likely_2nd = ifelse((trial %in% reverse_items2), 1, 0))
+
+d2 <- d2_tidy %>%
+  filter(trial_type == "experimental") %>%
+  mutate(response = ifelse(likely_2nd==1, response*-1, response)) %>%
+  mutate(response = -1*response) %>%
+  mutate(response = response + 3) %>%
+  mutate(trial = gsub("_trial","",trial))
+```
+
+```{r compute descriptives, include=FALSE}
+gender_exp1 <- d %>% 
+  group_by(workerId) %>% 
+  summarize(sex=first(sex)) %>% 
+  ungroup() %>% 
+  count(sex) %>% 
+  spread(sex, n)
+
+gender_exp2 <- d2 %>% 
+  group_by(workerId) %>% 
+  summarize(sex=first(sex)) %>% 
+  ungroup() %>% 
+  count(sex) %>% 
+  spread(sex, n)
+
+age_exp1 <- d %>% 
+  group_by(workerId) %>% 
+  summarize(age=first(age)) %>% 
+  ungroup() %>% 
+  summarize(age = median(age)) %>% 
+  .$age
+
+age_exp2 <- d2 %>% 
+  group_by(workerId) %>% 
+  summarize(age=first(age)) %>% 
+  ungroup() %>% 
+  summarize(age = median(age)) %>% 
+  .$age
+```
+
+```{r fit brms models, include=FALSE}
+library(brms)
+devtools::source_gist(id = "f1994c0f8325abbc5d300600744af39d", filename="cbrm.R")
+
+fit_exp1 <- cbrm(
+  response ~ 0 + intercept + trial + (1|workerId), # another model adds likely_2nd
+  data = d,
+  family = bernoulli(),
+  prior = prior(normal(0,3), class="b"),
+  control = list(adapt_delta = .80),
+  cores = parallel::detectCores(),
+  sample_prior = TRUE,
+  iter = 2000,
+  cached_file = "fit_exp1.rds"
+  )
+
+fit_exp2 <- cbrm(
+  response ~ trial + (1|workerId), # another model adds likely_2nd
+  data = d2,
+  family = cumulative(),
+  prior = prior(normal(0,3), class="Intercept"),
+  control = list(adapt_delta = .80),
+  cores = parallel::detectCores(),
+  sample_prior = TRUE,
+  iter = 2000,
+  cached_file = "fit_exp2.rds"
+  )
+
+fit_exp1rand <- cbrm(
+  response ~ 0 + intercept + (1|trial) + (1|workerId), # another model adds likely_2nd
+  data = d,
+  family = bernoulli(),
+  prior = prior(normal(0,3), class="b"),
+  control = list(adapt_delta = .80),
+  sample_prior = TRUE,
+  cores = parallel::detectCores(),
+  iter = 2000,
+  cached_file = "fit_exp1rand.rds"
+  )
+
+fit_exp2rand <- cbrm(
+  response ~ (1|trial) + (1|workerId), # another model adds likely_2nd
+  data = d2,
+  family = cumulative(),
+  prior = prior(normal(0,3), class="Intercept"),
+  control = list(adapt_delta = .80),
+  cores = parallel::detectCores(),
+  sample_prior = TRUE,
+  iter = 2000,
+  cached_file = "fit_exp2rand.rds"
+  )
+```
+
+```{r hypothesis tests, include=FALSE}
+print_bf <- function(hypothesis) {
+
+  bf <- hypothesis$hypothesis$Evid.Ratio
+
+  if (is.na(bf)) {
+    return("< .001")
+  }
+  else if (bf < .001) {
+    return("< .001")
+  }
+  else if (bf > 1000) {
+    return("> 1000")
+  }
+  else {
+    return(paste("=", as.character(bf)))
+  }
+}
+
+coefs_exp1 <- broom::tidy(fit_exp1rand) %>%
+  filter(term=="b_intercept")
+
+# coefs_exp2 <- broom::tidy(fit_exp2) %>%
+#   filter(term=="b_intercept")
+
+bf_exp1 <- hypothesis(fit_exp1rand, "intercept = 0")
+# bf_exp1 <- h1$hypothesis$Evid.Ratio
+
+bf_exp2 <- hypothesis(fit_exp2rand, "-Intercept[2] = Intercept[3]")
+# bf_exp2 <- h2$hypothesis$Evid.Ratio
+```
+
+```{r make figures, include=FALSE}
+fig1 <- plot(marginal_effects(fit_exp1, "trial"))
+fig_exp1 <- fig1$trial$data %>%
+  ggplot() +
+  aes(x = reorder(trial, -estimate__), y = estimate__, ymin = lower__, ymax = upper__) +
+  geom_errorbar(width=.25) +
+  geom_point(shape=17, size = 2) +
+  coord_cartesian(ylim=c(0,1)) +
+  geom_hline(yintercept = .5, linetype = "dashed", color = "grey") +
+  theme_bw() +
+  theme(
+    panel.grid.major = element_blank(),
+    panel.grid.minor = element_blank(),
+    aspect.ratio = .75,
+    axis.text.x = element_text(angle = 45, hjust = 1, vjust = .7)
+    ) +
+  labs(
+    x = "Trial",
+    y = "Probability of judging less\nexpected more upsetting"
+  ) 
+# library(gridExtra)
+# grid.arrange(ggplotGrob(fig_exp1))
+
+ggsave("fig1.eps",fig_exp1, height=8.7, width=8.7, unit="cm")
+```
+
+```{r fig2, include=FALSE}
+
+fig2_cont <- marginal_effects(fit_exp2, "trial", plot=FALSE)
+fig2 <- plot(marginal_effects(fit_exp2, "trial", ordinal=TRUE))
+
+plt_cont <- fig2_cont$trial %>%
+  ggplot() +
+  aes(x = reorder(trial, -estimate__), y = estimate__, ymin = lower__, ymax = upper__) +
+  geom_errorbar(width=0) +
+  geom_point(shape=17, size = 1) +
+  coord_cartesian(ylim=c(2,4)) +
+  geom_hline(yintercept = 3, linetype = "dashed", color = "grey") +
+  theme_bw() +
+  theme(
+    panel.grid.major = element_blank(),
+    panel.grid.minor = element_blank(),
+    aspect.ratio = 10/16,
+    axis.text.x = element_text(angle = 45, hjust = 1, vjust = .7)
+  ) +
+  labs(
+    x = "Trial",
+    y = '"Upsetting" Judgments'
+  )
+
+plt_heatmap <- fig2$trial$data %>%
+  left_join(
+    fig2_cont$trial %>% 
+              rename(mean_estimate = estimate__) %>%
+              select(trial, mean_estimate)
+            , by = "trial"
+    ) %>%
+  ggplot(aes(y = cats__, x = reorder(trial, -mean_estimate), fill = estimate__)) +
+  geom_tile(color="white") + 
+  scale_fill_viridis_c(
+    option="B", 
+    limits=c(0,1), 
+    breaks = c(0, .25, .5, .75, 1), 
+    labels= c(0, .25, .5, .75, 1),
+    guide = guide_colorbar(
+      title.position="top", 
+      barheight = unit(.5, "cm")
+      )
+    ) +
+  theme_minimal() +
+  theme(
+    panel.grid.major = element_blank(),
+    panel.grid.minor = element_blank(),
+    # aspect.ratio = 1,
+    legend.position = "top",
+    legend.key.width = unit(.75,"cm"),
+    axis.text.x = element_text(angle = 45, hjust = 1, vjust = .7)
+  ) +
+  labs(
+    x = "Trial",
+    y = "Response",
+    fill = "Probability"
+  ) +
+  coord_fixed(ratio=1)
+
+library(gtable)
+g2 <- ggplotGrob(plt_cont + theme(
+  axis.title.y = element_text(margin = margin(t=0,b=.5,l=0,r=.25,unit="cm")) # this is a kludge to get shared axis
+))
+g3 <- ggplotGrob(plt_heatmap + theme(
+  axis.title.x = element_blank(),
+  axis.title.y = element_blank(),
+  axis.text.x = element_blank()
+  ))
+
+g <- rbind(g3, g2, size = "last")
+# grid.arrange(g)
+
+library(gridExtra)
+
+ggsave("fig2.eps", grid.arrange(g), height=11.4, width=8.7, units = "cm")
+```
+
+When we learn of tragic events--a friend's loss of a loved-one, a violent killing, or a terrorist attack--we are mournful, concerned, angry, and driven to action.
+But our responses to these events differ depending on the context [e.g., @Trope2010].
+In particular, our expectations play an important role in our reactions to events.
+This seems true of both the mundane--when a friend's loved-one passes, we often ask, "was it sudden?"--the dramatic--a murder in a sleepy town prompts a very different reaction than a similar crime in a major city--and even the catastrophic: acts of terrorism seem to evoke more outrage, horror, and concern when they occurs in peaceful rather than unstable regions.
+For instance, the 2015 Paris attacks were felt around the world.
+Yet, most of those mourning had little to say 15 hours earlier, when another tragic attack killed at least 43 people in Beirut [@Graham2015].
+Likewise, stories of the Virginia Tech massacre, which killed 32 people, wildly overshadowed coverage of attacks killing nearly 200 people in Iraq--one of the bloodiest days since the 2003 invasion [@Greenslade2007].
+In each of these situations, the surprise elicited by an event appears to affect people’s reactions to the event.
+In some contexts, harm is tragic and in others, “these things happen.”
+
+Although this pattern seems so intuitive as to be obvious, no reasonable code of ethics would justify this pattern of judgment, nor do any moral psychological frameworks explicate the role of expectations in moral evaluations.
+Outside of the domain of morality, however, everyday experiences and controlled experiments show that people’s evaluations of events unambiguously depend on their expectations about those events.
+There is a special thrill to having one’s expectations exceeded and disappointment when events fail to meet expectations: we root for the underdog, hold surprise parties, and foreshadow bad news to ease its delivery [@Bell1985].
+Indeed, in a controlled study, Mellers and colleagues [-@Mellers1997] found that expectations influenced affective reactions during a gambling task: Given a gamble with a 10% chance to win \$30 and 90% chance to win \$0, participants felt little disappointment with the \$0 outcome, but were considerably more excited when they won \$30.
 Conversely, given a gamble with a 90% chance of winning \$60 and 10% chance of winning \$0, participants were disappointed with the \$0 outcome and showed more muted enjoyment of the \$60 outcome.
-In fact, in gambles similar to these, Mellers and colleagues [-@Mellers1997] found that people were happier with the smaller unexpected gain than with the larger but more expected gain [also see @Shepperd2002].
+In fact, in gambles similar to these, people were happier with the smaller unexpected gain than with the larger but more expected gain [also see @Shepperd2002].
 
-Just as expectations affect the utility of simple gambles, we suspect they are equally likely to shape how people react to morally harmful events, such as acts of terrorism (Waldmann Cite and Rai and Fiske cite --domain general peeps; us also).
-However, unlike in the context of gambles, in these contexts the effects of expectations on evaluations may have harmful consequences.
-When events are shocking, we predict that people will perceive them as more severe and consequently be roused to action.
-In contrast, when events harm victims who are generally considered to be at greater risk--the poor, sick, or those living in unstable regions of the world, we predicted that reactions may be more muted.
-However, prior research has yet to articulate whether these same effects emerge in the moral domain and the implications of this possibility.
-For instance, it is unclear if these same mechanisms can help account for the divergent reactions people have to terrorism in France and Beirut.
-Further, people’s moral evaluations are in fact impacted by expectations what is their source and what does that tell us about how to overcome the biases they may impose?
+Applied to moral issues, the influences of expectations on evaluation would be a dangerous bias.
+Tragedies are tragic, whether expected or not, and failure to recognize that could lead to tacit endorsement of the pain and suffering of others wherever pain and suffering are the status quo.
+Do expectations directly bias reactions to morally consequential events, as intuition and anecdote seem to suggest?
+To demonstrate the plausibility of this suggestion, we develop a theoretical account of how expectations influence evaluations, not as a strange quirk or uniquely human bias, but instead as a fundamental component of how people evaluate events.
 
-A number of researchers have sought to develop theories of disappointment--the psychological reactions that result when experiences fail to meet expectations--and its role in evaluation and decision-making [e.g., @Bell1985; @Gul1991; @Loomes1986].
-These researchers have generally argued that expectations modify people's reactions to events, so that they are jointly determined by the context-free utilities of options or events (e.g., economic utilities), and the contextually-dependent disappointment that individuals experience as a function of their expectations.
-Similarly, Mellers and colleagues (1997) interpreted the divergence they observed between the utility of an outcome (i.e., monetary gains or losses) and the felt experience of that outcome as a result of post-decision affect.
+Most events of consequence consist of changes--a transition from one state to another.
+For this reason, we construe event evaluation as a comparison between the states of the world before and after the event.
+Comparison plays a crucial role in many theories of decision-making, and especially in theories aimed at explaining the role of expectations in evaluation  [e.g., @Bell1985; @Gul1991; @Loomes1986; @Mellers1997].
+For instance, in Prospect Theory [@Kahneman1979; @Tversky1992], the utility of a prospect is evaluated by comparison to a current reference point.
+Our proposal is similar: we argue that expectations can set the reference points against which people compare future outcomes, so that assigning value to an event consists in comparing the state of the world following the event to the state of the world just prior to the event.
+As knowledge of the world is uncertain, our understanding of these states is probabilistic--that is, our model for the state of the world is just our expectations about what has happened, what is happening, and what will happen in the future.
+One direct implication of this proposal is that the evaluation of an event is linked to what is learned about the world as a result of that event’s occurrence.
+That is, we evaluate events positively or negatively to the extent that they inform our understanding of the world in positive or negative ways.
 
-Each of these theories has been rigorously defined and supported within the small-world confines they were meant to capture.
-However, we seek to craft a more general explanation of how and why expectations influence evaluations and the range of contexts in which they should be expected to do so.
-Several accounts of utility evaluation, including early theories like Prospect Theory [@Kahneman1979; @Tversky1992], have emphasized the role of relative comparisons in evaluations of utility.
-In a similar spirit, we argue that expectations can set the reference points against which people compare future outcomes.
-On this view, evaluating the utility of some event consists in comparing the state of the world now that the event has occurred to the state of the world just prior to the event.
-Further, we assume that our knowledge of the world is uncertain, so that it is really just our (probabilistic) expectations about what has happened, what is happening, and what will happen in the future.
-Thus, the evaluation of an event might be driven by what is learned about the state of the world as a result of that event’s occurrence.
+This account thereby provides an information-theoretic route to understanding how expectations directly influence evaluation.
+A fundamental insight of Information Theory is that the information carried by an event is a function of its prior probability; low probability events carry more information than high probability events [@Shannon1948].
+We propose that people learn more about the state of the world when their expectations are violated by shocking world events as compared to when they are affirmed by less surprising events.
+For example, if a bombing occurs in Paris--a less expected location--rather than Lebanon--a more expected location--we learn that the world is more dangerous than we had previously believed.
+This suggestion is novel in the moral domain, but these points are widely accepted elsewhere.
+For instance, expectations have long been recognized as fundamental to models of human and animal learning [e.g., @Rescorla1972].
+Similarly, science progresses most when new findings impugn widely-accepted theories or when a theory’s extreme predictions are validated.
 
-The account outlined above provides an information-theoretic route to understanding how expectations directly influence evaluation.
-A fundamental insight of Information Theory is that the information carried by an event is a function of its prior probability.
-This means that low probability events carry more information than high probability events [@Shannon1948].
-Along similar lines, violations of expectations have long been recognized as fundamental to associative and animal learning models; a larger delta means people learn more [e.g., @Rescorla1972].
-Science progresses most when new findings impugn widely-accepted theories or when a theory’s extreme predictions are validated.
-Likewise, people learn more about the state of the world when their expectations are violated by shocking world events as compared to when they are affirmed by less surprising events.
-For example, if a bombing occurs in Paris -- an unexpected location -- rather than Lebanon --a more expected location--we learn that the world is more dangerous than we had previously believed.
+A more formal treatment of this theory establishes the potential for its generality across domains.
+Let the state of the world prior to some event be represented by a random variable, $X$.
+We'll assume $X$ is discrete, and that every realization $x$ represents some potential specific state of the world in all relevant aspects under consideration.
+An agent has some utility function $u$, that applies over the different realizations of $X$.
+Thus the expected utility of the present state of the world is:
 
-A more formal treatment of these issues reveals the generality of these conclusions.
-Formally, before some binary event $x_i$ (which either occurs or does not occur), an evaluator has a mental model of the current state of affairs, S.
-This mental model includes expectations about potential future events and uncertainty over current and past events.
-Thus S represents a joint probability distribution over possible states of affairs $p(x_1, x_2, ..., x_n)$.
-The evaluator's subjective utility function can assign an expected utility to S--representing how positive or negatively the evaluator views the current state of affairs.
-The occurrence of $x_i$ brings about some new state of affairs, S', to which the evaluator can again apply her utility function.
-The evaluation of event  $x_i$ then consists in the comparison between these two states of affairs, those before the event (reflecting the agents' expectations), and the new state of affairs brought about by the event:
+\begin{equation}
+E_X[u(X)] = \sum_x u(x)p(x)
+\end{equation}
 
-$$ V(x_i) = EU(S') -EU(S) $$
+When we learn an event has occurred, we can represent this as an observation of a second, binary random variable, $S$.
+We can think of $S$ as a "signalling event" that informs us about about $X$ [following the formalism deployed by @Arrow1996].
+That is, we assume there is a dependency between $S$ and $X$, so that observing $S$ informs us about the state of the world, $X$.
+For instance, reading a news report of a terror attack in Brussels informs us about the state of the world, telling us about what realizations of $X$ are no longer possible (e.g., any wherein it was a peaceful day in Brussels) and which are more or less likely.
+Conditioned on $S$, the expected utility of the world is:
 
-We let S be represented by a random vector X, whose members are any number of discrete events $x_1, x_2, ..., x_n$, where $x_i$ represents the event of interest so that $x_i$ occurs in S with some probability and occurs in S' with probability 1.
-That is, S represents $X$ and S' represents $X|x_i=1$.
-We write this as:
+\begin{equation}
+E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
+\end{equation}
 
-$$
-\begin{aligned}
-V(x_i) &= EU(X|x_i) - EU(X) \\
-&= \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x)
-\end{aligned}$$
+Given our construal of event evaluation as the comparison between states of the world following and prior to the event, we define the value assigned to the signalling event, $V(S)$, as:
 
-Applying the chain rule of probability and some algebra (see appendix) we eventually obtain:
+\begin{equation}
+V(S) = E_{X|S}[u(X)] - E_X[u(X)]
+\end{equation}
 
-$$V(x_i) = \big(1 - p(x_i) \big) \sum_{x \in X} U(x)p(x|x_i)$$
+From this it can be shown that the value assigned to event $S$ is proportional to its prior probability (see Theoretical Formalization).
+Specifically, it is the difference in the expected utility under different values of $S$, weighted by the inverse of the prior probability of $S$..
+[^1]
 
-Thus, the value assigned to event $x_i$ is proportional to the prior probability of $x_i$.
-Further, this holds true regardless of the nature of the event in question and of the form of the utility function over events.
+\begin{equation}
+V(s=1)=(1-p(s=1))\sum_x u(x)p(x|s=1) - u(x)p(x|s=0)
+\end{equation}
 
-In the context of our reaction to surprise parties, it might seem desirable that expectations would influence our experience of events.
-Much joy is tied to the psychological marriage between surprise and evaluation.
-However, this basic cognitive process may have negative moral repercussions.
-When harms are expected, these dynamics may lead people to draw less extreme evaluations, feel less concern for victims, and be less driven to help.
-To test this hypothesis, we examined whether people’s evaluations of morally harmful events are affected by their expectations about those events.
-We asked people to compare pairs of simple events where a victim suffered an identical harm, but where the events differed in their prior probability.
-For each pair of events, participants were asked to judge which event was more upsetting.
-Two studies provided evidence for the predicted effect of expectations on moral evaluations: people tended to view unexpected negative events as more upsetting, even when the harm to victims was identical.
+This conclusion does not depend on the nature of the event in question nor on the utility function over events, but instead holds generally wherever we conceive of evaluation as a holistic comparison between uncertain states of the world.
+Consequently, this account predicts expectations will equally affect reactions to gambles with clearly specified values and probabilities [@Mellers1997] and real-world events with ill-defined values and where reasoners' expectations are defined by their existing beliefs and world knowledge.
 
-# General Methods
+Given the potentially fundamental role expectations play in event evaluation, and the potentially serious moral biases that may result, we sought to test empirically whether people’s evaluations of morally harmful events are affected by their expectations about those events.
+We conducted two preregistered studies examining the role of expectations in moral evaluations.
+In both studies, participants were presented with a series of trials where they read brief descriptions of two different events and were asked to indicate which of the two events seemed more upsetting.
+In these trials, the two events were highly similar, but were manipulated so that they differed in their perceived prior probabilities: one event was more expected and one more unexpected.
 
-Here we present two preregistered studies examining the role of expectations in moral evaluations.
-Study 1 used a forced-choice task to test the hypothesis that people would be more upset about moral outcomes that were unexpected than expected.
-In Study 2, we tested of our hypothesis using a more conservative judgment task to increase the generalizability of our results.
+The design of these studies was chosen with special consideration to two theoretical points.
+First, the information-theoretic account we have described makes a solely directional prediction--all else being equal, more surprising events will elicit stronger reactions than less surprising events.
+Asking participants to compare two similar events allowed us to test this directional prediction.
+Second, it is the informativeness of events, determined by one's prior expectations of the event, that we predict will shape evaluations.
+Thus, reasoners should not have to be induced to hold expectations experimentally [as in experiments by @Mellers1997], but instead their own prior beliefs will be sufficient to shape their expectations and affect their evaluations.
+For this reason, we manipulated the prior probability of events by describing two otherwise identical events occurring in different contexts.
+This allowed us to test the role of expectations based on world knowledge, as well as providing a more naturalistic set of items.
 
-## Materials
+![Probability with which participants chose the less expected event as more upsetting across trials in Study 1.
+As predicted, participants exhibited a robust bias toward choosing the less expected event as more upsetting across all items.
+Error bars represent 95% credible intervals calculated from the hierarchical logistic regression model.
+See Supporting Information for full description of items.
+<span data-label="fig:fig1"></span>](fig1.eps)
 
-In both studies, participants were presented with a series of trials where they read brief (one sentence) descriptions of two different events and were asked to indicate which of the two events seemed more upsetting.
-In “experimental” trials, the two events were highly similar, but differed in their prior probabilities: one event was more expected and one more unexpected.
-(The perceived likelihood of a given event was confirmed in prior norming studies).
-These prior expectations were manipulated by changing the context in which the events occurred.
+Consistent with our predictions, manipulating expectations consistently shaped evaluations across a range of morally-significant events.
+We found that people viewed unexpected harmful events as more upsetting than expected harmful events, even when the harm a victim suffered was the same.
+For instance, people judged that a robbery at a clothing store (unexpected) was more upsetting than a robbery at a convenience store (expected).
+Furthermore, these differences were large, systematic, and robust.
+Although there was variability in the situations that elicited the largest differences, our statistical approach treated trial as a random effect (see Methods and Results), allowing us to generalize conclusions to items we did not test [@Judd2012].
+
+![Participants’ responses by item in Study 2.
+Upper: heatmap plot displaying probability of choosing each response on the 1 to 5 scale across items.
+Lower: marginal expected response across items, with 95% credible interval calculated from the ordinal regression model.
+Note the scale has been constrained to show relevant region of results.
+In both plots, response of 3 indicates indifference, a response below 3 indicates a bias to choose the more expected event as more upsetting, and a response above 3 indicates a bias to choose the less expected event as more upsetting.
+Similar to Study 1, when participants judged that one of the events was more upsetting, they showed a robust bias toward choosing the less expected event as more upsetting.
+Trials are sorted by response to aid readability.
+<span data-label="fig:fig1"></span>](fig2.eps)
+
+We have argued that expectations shape our reactions to events because they determine how informative those events are in shaping our understanding of the world.
+Thus, we argue that well-grounded information theoretic principles can account for people’s divergent moral reactions to likely and unlikely events.
+
+There's a sense in which it is clearly adaptive that the violation of our expectations would direct our attention to the events and situations about which our understanding is most in need of revision.
+For instance, an act of terrorism in a peaceful nation might be less expected because of our beliefs about the effectiveness of security forces, the stability of national relationships, and positive relationships among people groups in that nation.
+Therefore, a surprising act of terrorism casts doubt on our deeper understanding of the geopolitical situation, and thereby leads us to suspect similar acts are more likely to occur in the future.
+
+Yet, however well-grounded the underlying principles, there’s a clear moral danger whenever expectations influence our moral concerns.
+For instance, "victim blaming" describes the tendency to shift blame away from aggressors and onto victims of crimes like sexual assault [@Koss2000] and police violence.
+Work by Janoff-Bulman and colleagues [-@Janoff-bulman1985] has shown that merely increasing the perceived probability of a victim suffering a sexual assault induced participants to blame the victim for her assault.
+People’s expectations may explain the flip side of victim blaming, whereby those who victim blame also tend to judge transgressors less harshly: When people perceive a victims’ actions as increasing the probability of their eventual assault, they may perceive an aggressor’s actions as producing a smaller change in utility, thus warranting weaker condemnation.
+
+The influence of expectations on moral evaluations poses perhaps the greatest threat when considering our reactions to humanitarian challenges brought on by poverty or regional instability.
+In these cases, the influence of expectations on moral evaluations threatens to impose vicious and morally pernicious cycles.
+For instance, people living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and are the very people in greatest need of assistance and concern from the world at-large.
+However, for these very reasons, it is often unsurprising when harm befalls people living in these circumstances.
+Thus the influence of expectations on moral evaluations threatens to reduce observer's concern for these victims most in need, making people less likely to donate time or money to aid victims, or to take political action to prevent further harms.
+The present research identifies this potentially pernicious moral bias and grounds it in information theoretic principles.
+Hopefully, by further understanding these cognitive dynamics, psychologists and policymakers can begin to find ways to reduce or counteract people's tendency to ignore the plight of those in most need of attention.
+
+# Materials and Methods {#materials-and-methods .unnumbered}
+
+We conducted two preregistered studies examining the role of expectations in moral evaluations using two different judgment tasks (https://osf.io/5qnf7/).
+Each study consisted of multiple trials where participants were presented brief descriptions of two events.
+Their task was to decide which of the two events was more upsetting.
+In Study 1, they made a two-alternative forced-choice, and in Study 2 they made their judgments using a five-point rating scale.
+Both studies consisted of 16 experimental trials, five “equivalent” filler trials, and five “non-equivalent” filler trials.
+
+In experimental trials, the two events described identical harms, but one event was more expected and one more unexpected, as determined in a prior norming study (see SI).
 For example, participants considered the following stimulus:
@@
 
-In each event, the harm to the victim is the same (here, death) but one event is more expected than the other, given the different likelihoods of earthquakes occurring in California versus Oklahoma.
+In each of these events, the harm to the victim is the same (death) but one event is more probable than the other, given the different likelihoods of earthquakes occurring in California versus Oklahoma.
+In addition to being more naturalistic, manipulating expectations through (otherwise irrelevant) contextual details helps ensure that the harm suffered by the victim is understood to be identical.
+In the context of gambles, it is possible to manipulate probabilities specifically and directly [@Mellers1997].
+However, in the context of more realistic moral harms, any attempt to directly manipulate probabilities risks also changing participants’ perceptions of the degree of harm.
+For instance, suppose we describe two cancer patients, one with a 5% chance of survival and one with a 50% chance of survival.
+Even if they both ultimately meet the same fate, participants might fairly assume the patient with the lower chance of survival had a more serious illness, was more debilitated, suffered more severely, and so forth.
+In contrast, changing the context of the two events explains away these likelihood differences, better equating the events in other respects.
 
-Each study contained between 16 experimental event-pairs that spanned a variety of different events and contexts.
-All experimental materials for these studies are available as supplemental online materials at [OSF LINK].
-
-Both studies included “equivalent” filler trials. In these trials, the two events differed in trivial contextual details that we did not expect would affect participants’ judgments. For example:
-
-* “A man in Connecticut starts a house fire.” [Equally expected]
-
-* “A man in New Hampshire starts a house fire.” [Equally expected]
-
+Both studies also included “equivalent” filler trials.
+In these trials, the two events differed in trivial contextual details that we did not expect would affect participants’ judgments.
 These filler trials were meant to prevent participants from becoming explicitly aware of the structure of the experimental trials.
 Finally, both studies included “non-equivalent” filler trials, the two events differed substantially in the degree of harm suffered by a victim, so that one event was expected to be seen as considerably more upsetting than the other.
-For example:
+These trials were included to allow participants a chance to use the extremes of the response scale and to reduce any task demands that might drive them to make artificially fine-grained distinctions when generating their responses.
+All experimental materials for these studies are available as Supporting Information at https://osf.io/5qnf7/.
 
-* “An 11-year-old child sets a doll on fire” [Less severe]
+## Exclusions {#exclusions .unnumbered}
 
-* “A 12-year-old child sets a cat on fire”  [More severe]
+In both studies, we excluded participants who failed to correctly answer attention-check questions or who indicated at the end of the study they had not taken their participation seriously.
 
-These trials were included to allow participants a chance to use the extremes of the response scale and to reduce any task demands that might drive them to make artificially fine-grained distinctions between the severity of events.
-
-## Exclusions
-
-In both studies, we excluded participants who failed to incorrectly answer attention-check questions.
-These questions asked participants to enter a particular response to ensure that they were paying attention and reading the items as they proceeded through the study.
-A final question asked participants if they had paid attention and taken the study seriously, encouraging them to be honest in their replies.
-
-## Data Analysis
+## Data Analysis {#data-analysis .unnumbered}
 
 We analyzed our data by performing Bayesian estimation using the probabilistic programming language Stan [@Carpenter2017].
-We tested our predictions by computing Bayes Factors (i.e.
-BF01) on the intercept term of our regression model using the hypothesis function in the R package brms.
+We tested our preregistered predictions by computing Bayes Factors (i.e.
+BF01) using the hypothesis function in the R package brms.
 Bayes Factors express the ratio of the probability of data under the null hypothesis to the probability of the data under an alternative hypothesis.
-Therefore, larger Bayes Factors indicate that the data are more likely under the null hypothesis (e.g., that the intercept is not different from zero) than the alternative hypothesis (e.g., that the intercept is different from zero), and vice versa.
-Bayes Factors can be influenced by prior choices (Gelman, Simpson, & Betancourt, 2017) so we also performed prior robustness checks to confirm that our conclusions were unchanged under different prior assumptions.
+Larger Bayes Factors indicate that the data are more likely under the null hypothesis (i.e., that the intercept is not different from zero) than the alternative hypothesis (i.e., that the intercept is different from zero), and vice versa.
+As Bayes Factors can be influenced by prior choices (Gelman, Simpson, & Betancourt, 2017), we performed prior robustness checks to confirm that our conclusions would not depend substantially on the specification of the priors.
 
-# Appendix
+To ensure the generality of our findings, we tested participants’ reactions to a variety of items and treated both participants and items as random effects throughout our analyses.
+The use of random effects over items allows us to consider these items as a sample from a larger population of possible items, and to generalize our findings to that wider population, much as we generalize from our sample of participants to the larger population [@Judd2012].
 
-$$ \begin{aligned}
-V(x_i) &= EU(X|x_i) - EU(X) \\
-&=  \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x) \\
-&= \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x|x_i)p(x_i) \\
-&= \sum_{x \in X} U(x)p(x|x_i) - p(x_i)\sum_{x \in X} U(x)p(x|x_i) \\
-&= \big(1 - p(x_i) \big) \sum_{x \in X} U(x)p(x|x_i) \\
-\end{aligned}$$
+## Study 1 {#study1 .unnumbered}
 
-The first step utilizes the Law of the Unconscious Statistician, the next comes from the chain rule of probability, and the final two steps are simple algebra.
+### Participants {#s1-participants .unnumbered}
 
-# References
+A total of `r n_distinct(d_raw$workerId)` participants were recruited from Amazon’s Mechanical Turk work distribution website (mTurk).
+Of these, `r n_distinct(d$workerId)` passed attention checks and were included in the final analyses (`r gender_exp1$male` male, `r gender_exp1$female` female, median age = `r age_exp1` years old).
+All participants were paid $0.60 for their participation.
 
-\setlength{\parindent}{-0.25in} 
-\setlength{\leftskip}{0.125in}
-\noindent
+### Materials and procedure {#s1-materials-procedure .unnumbered}
+
+Participants judged 16 experimental event-pairs and 10 equivalent filler event-pairs.
+On each trial (an example is shown above), participants were presented with the event-pair stimulus and had to judge which outcome was more upsetting in a two-alternative forced choice task.
+The two events were labeled “Outcome 1” and “Outcome 2” and the order, and whether Outcome 1 or 2 was the unexpected event, was randomized and counter-balanced.
+
+### Results and discussion {#s1-results .unnumbered}
+
+We predicted that people would judge that events where unexpected harm occurred were more upsetting than events where expected harm occurred.
+As indicated in Figure 1, we observed this trend.
+We estimated the magnitude of this effect by fitting a Bayesian logistic random effects model with participants’ responses as the dependent variable (1 = unexpected event is more upsetting; 0 = expected event is more upsetting) and random intercepts for item and subject, allowing us to generalize both to items and subjects we did not test.
+The intercept in this model represents the log-odds of selecting the unexpected event as being more upsetting.
+Thus, by examining the population-level intercept, we can test whether participants were biased toward selecting the unexpected event ($\beta$ > 0), the expected event ($\beta$ < 0), or were unbiased ($\beta$ = 0).
+We found that people were considerably more likely to think that unexpected events were more upsetting than events that were expected, Intercept = `r round(coefs_exp1$estimate, 3)`, 95% CI [`r round(coefs_exp1$lower, 3)`, `r round(coefs_exp1$upper, 3)`], BF01 `r print_bf(bf_exp1)`.
+Figure 1 shows participants' responses broken-down by individual items. Participants' bias toward selecting the unexpected event as more upsetting was consistent across the 16 experimental items.
+
+## Study 2 {#study2 .unnumbered}
+
+In Study 1 we found that people’s judgments of events were biased by their expectations about those events.
+When forced to choose between two events, participants decided that unexpected events were more upsetting than expected events.
+In Study 2, we sought to test our hypothesis using a more conservative method: we provided participants with a more expressive response scale so that if they viewed the events under consideration as equally harmful, their responses could reflect their attitude.
+
+### Participants {#s2-participants .unnumbered}
+
+A total of `r n_distinct(d2_raw$workerId)` participants were recruited from Amazon’s Mechanical Turk work distribution website (mTurk).
+Of these, `r n_distinct(d2$workerId)` passed attention checks and were included in the final analyses (`r gender_exp2$male` male, `r gender_exp2$female` female, median age = `r age_exp2` years old).
+All participants were paid $0.60 for their participation.
+
+### Materials and procedure {#s2-materials-procedure .unnumbered}
+
+The materials in Study 2 were the same as those in Study 1.
+As in Study 1, on each trial of the study, participants were presented with a pair of actions labeled “Outcome 1” and “Outcome 2” and were asked, “Which outcome seems more upsetting?” In Study 2 participants made their rating on a five-point scale (Outcome 1 seems more upsetting, Outcome 1 seems a little more upsetting, neither seems more upsetting than the other, Outcome 2 seems a little more upsetting, Outcome 2 seems more upsetting).
+
+### Results and discussion {#s2-results .unnumbered}
+
+We predicted that participants would evaluate unexpected events as more upsetting than expected events.
+We estimated the magnitude of this effect by fitting a cumulative (ordinal) random effects model with participants’ scale responses as the dependent variable (1 to 5) and random intercepts for item and subject.
+This model produces four intercept coefficients, representing the cumulative log-odds of responses at each scale point or lower.
+For instance, the second coefficient represents the log-odds participants chose a 2 ("outcome 2 seems slightly more upsetting") or lower on the scale.
+Similarly, the third intercept coefficient represents the log-odds participants chose a 3 or lower on the scale.
+To examine the predicted effect, we compared the third intercept coefficient (representing the log-odds of choosing 4 or higher) to the inverse of the second intercept coefficient (thereby representing the log-odds *not* choosing a 3 or higher--i.e., choosing a 1 or 2), allowing us to test whether participants were more likely to choose the expected or unexpected event as being more upsetting in cases where they did not choose the neither option.
+This analysis indicated that people were more likely to think that events that were unexpected were more upsetting than events that were expected, BF01 `r print_bf(bf_exp2)` (see supporting information for full model results).
+We observed this difference consistently across nearly every item (Figure 2).
+
+We note that the most common response was to view both events, which involve identical harms, as equivalently upsetting.
+However, when participants did view one of the events as more upsetting, they showed a clear bias towards choosing unexpected events as more upsetting.
+Thus, Study 2 replicates the tendency to view unexpected moral events as more upsetting using a more conservative judgment task.
+
+# Theoretical Formalization {#theory .unnumbered}
+
+Evaluation (e.g., of utility) is often examined in the context of decision-making.
+However, people often generate post hoc evaluations of events for their own sake.
+For instance, to reflect on the results of our choices and to evaluate others--to judge the wisdom of their decisions or the morality of their actions.
+
+We construe event evaluation as a comparison between the state of affairs prior to an event and the state of affairs following the event.
+We represent the state of the world as a random variable, $X$.
+We assume $X$ is discrete, and that every realization $x$ represents some potential specific state of the world in all relevant aspects under consideration.
+An agent has some utility function $u$, that applies over the different realizations of $X$.
+Thus the expected utility of the present state of the world is:
+
+\begin{equation}
+E_X[u(X)] = \sum_x u(x)p(x)
+\end{equation}
+
+When we learn an event has occurred, we can represent this as an observation of another random variable, $S$.
+We can think of $S$ as a "signalling event" because it informs (signals) to us something about $X$ [this formalism follows @Arrow1996].
+For simplicity, we treat $S$ as binary, either 1 or 0.
+Importantly, we assume there is a dependency between $S$ and $X$, so that observing $S$ informs our knowledge of $X$.
+For instance, reading a news report of a terror attack in Brussels informs us about the state of the world, telling us about what realizations of X are no longer possible (e.g., any wherein it was a peaceful day in Brussels) and which are more or less likely.
+Conditioned on $S$, the expected utility of the world is:
+
+\begin{equation}
+E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
+\end{equation}
+
+We define the value assigned to the signalling event, $V(S)$, as:
+
+\begin{equation}
+V(S) = E_{X|S}[u(X)] - E_X[u(X)]
+\end{equation}
+
+Which we write more specifically as:
+
+\begin{equation}
+V(s=1) = \sum_x u(x)p(x|s=1) - \sum_x u(x)p(x)
+\end{equation}
+
+Applying the expectation of conditional expectation, we can write:
+
+\begin{equation}
+V(s=1) = \sum_x u(x)p(x|s=1) - \sum_x \sum_s u(x)p(x|s)p(s)
+\end{equation}
+
+We then perform some algebra.
+We break out the summation over $S$ (6), pull the multipliers out of the last two sums (7), rewrite $p(s=0)$ in terms of $p(s=1)$ (8), and factor to combine the first two sums (9).
+
+\begin{align}
+V(s=1) &= \sum_x u(x)p(x|s=1) - \sum_x u(x)p(x|s=1)p(s=1) - \sum_x u(x)p(x|s=0)p(s=0) \\
+ &= \sum_x u(x)p(x|s=1) - p(s=1)\sum_x u(x)p(x|s=1) - p(s=0) \sum_x u(x)p(x|s=0) \\
+ &= \sum_x u(x)p(x|s=1) - p(s=1)\sum_x u(x)p(x|s=1) - (1-p(s=1)) \sum_x u(x)p(x|s=0) \\
+ &=(1-p(s=1))\sum_x u(x)p(x|s=1) - (1-p(s=1)) \sum_x u(x)p(x|s=0)
+\end{align}
+
+Finally, combining both sums by their common factor yields (10):
+
+\begin{equation}
+V(s=1)=(1-p(s=1))\sum_x u(x)p(x|s=1) - u(x)p(x|s=0)
+\end{equation}
+
+Thus, it can be seen under this formalization that the role of expectations in the evaluation of events is fundamental--it does not, for instance, depend on any specifics concerning the utility function over states of the world.
+Instead, this conclusion depends on just two assumptions: first, that events are evaluated by comparing the utilities of the states of the world prior to and following the event; and second, that knowledge of those states of the world is uncertain.
+
+We have appealed to the informativeness of an event to illustrate intuitively how expectations should influence evaluations.
+More expected events carry less information, and thereby engender weaker reactions.
+Though our model of event evaluation is not an information-theoretic model in a direct sense, this interpretation still applies: the information carried by observing $S$ is a function of that observation’s prior probability, which we have in turn shown determines its evaluation.
+Also relevant, although not addressed here, is the mutual information between $X$ and $S$--the extent to which $S$ informs $X$.
+
+[^1]: It should be recognized that the result we obtain here bears a similarity to the “disappointment function” included in Mellers and colleagues’ [-@Mellers1997] Decision Affect Theory.
+
+<!-- Leave these lines as they are at the end of your .Rmd file to ensure placement of methods & acknowledgements sections before the references-->
+\showmatmethods
+\showacknow
+\pnasbreak

### 2018-06-29 paper/paper.tex
added norming data

--- 2018-06-02 paper/paper.tex
@@
-\documentclass[]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
+
+%% Some pieces required from the pandoc template
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% Use the lineno option to display guide line numbers if required.
+% Note that the use of elements such as single-column equations
+% may affect the guide line number alignment.
+
 \usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage[margin=1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdftitle={Expectations bias moral evaluations},
-            pdfauthor={Derek Powell and Zachary Horne},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
 
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
+\templatetype{pnasresearcharticle}  % Choose template
 
 \title{Expectations bias moral evaluations}
-    \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
-    \author{Derek Powell and Zachary Horne}
-    \preauthor{\centering\large\emph}
-  \postauthor{\par}
-    \date{}
-    \predate{}\postdate{}
+
+\author[a,1]{Derek Powell}
+\author[b]{Zachary Horne}
+
+  \affil[a]{Stanford University, Department of Psychology, 450 Serra Mall, Stanford,
+CA, 94305}
+  \affil[b]{Arizona State University, School of Social and Behavioral Sciences,
+13591 N 47th Ave, Phoenix, AZ 85051}
+
+% Please give the surname of the lead author for the running footer
+\leadauthor{Powell}
+
+% Please add here a significance statement to explain the relevance of your work
+\significancestatement{People living in geo-politically unstable regions or in the developing
+world are often those who are most affected by terrorism, famine, and
+natural disasters, and thus are the very people in greatest need of
+assistance and concern from the world at-large. However, for these very
+reasons, it is often unsurprising when harm befalls them. Two
+experiments confirm this intuition: we found that people's expectations
+bias their reactions to morally-significant events--people are more
+upset by surprising harms than more expected harms. This bias threatens
+to reduce observer's concern for victims most in need, making people
+less likely to donate time or money to aid victims, or to take political
+action to prevent further harms.}
+
+\authorcontributions{Please provide details of author contributions here.}
+
+\authordeclaration{Please declare any conflict of interest here.}
+
+\correspondingauthor{\textsuperscript{} }
+
+% Keywords are not mandatory, but authors are strongly encouraged to provide them. If provided, please include two to five keywords, separated by the pipe symbol, e.g:
+ \keywords{  evaluation |  moral judgment |  information theory   }
+
+\begin{abstract}
+When we learn of tragic events--a violent killing or a terrorist
+attack--we are mournful and often driven to action. But our responses to
+these events can differ substantially depending on the context of the
+event. When tragedy takes us by surprise, we are shocked and outraged,
+but when it strikes in unstable regions or rough neighborhoods, we are
+often resigned to feel that ``these things happen.'' Here we explain the
+role that expectations seem to play in moral evaluation by modeling
+evaluations as the comparison of the state of the world before and after
+an event. According to this model, expectations influence evaluations
+through information-theoretic principles: less expected events do more
+to inform us about the state of the world than do more expected events.
+Consistent with this theory, in two preregistered studies we found that
+people's judgments of morally significant events were affected by the
+prior probability of that event, so that people were more upset about an
+event that was unexpected (e.g., a robbery at a clothing store) than an
+event that was more more probable (e.g., a robbery at a convenience
+store). This bias threatens to impose a pernicious cycle: People living
+in poverty, geo-politically unstable regions, and other dangerous
+circumstances have the greatest need for assistance and concern.
+However, for these very reasons, it is often unsurprising when they are
+harmed. Consequently, the influence of expectations on moral evaluations
+threatens to reduce observer's concern for those people who are most in
+need.
+\end{abstract}
+
+\dates{This manuscript was compiled on \today}
+\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}
 
 \begin{document}
+
+% Optional adjustment to line up main text (after abstract) of first page with line numbers, when using both lineno and twocolumn options.
+% You should only change this length when you've finalised the article contents.
+\verticaladjustment{-2pt}
+
 \maketitle
+\thispagestyle{firststyle}
+\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
 
-\section{Introduction}\label{introduction}
+% If your first paragraph (i.e.
+with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation.
+If this is the case, add \parshape=0 to the end of the list environment.
 
-On the evening of November 13th, 2015, a terrorist attack in Paris left
-130 people dead and injured over 300 more. In the aftermath, millions
-took to Twitter to express their shock, horror, and outrage at this
-tragedy under hashtags like \texttt{\#parisattacks} and
-\texttt{\#jesuisparis}. Yet, most of those mourning had little to say 15
-hours earlier, when another tragic attack killed at least 43 people in
-Beirut. Several factors are surely at play in these different reactions
-(construal-level theory, {\textbf{???}}; e.g.~group affiliations,
-Brewer, 1999), yet one potentially fundamental factor has gone
-unmentioned: the fact that the Paris attack was more surprising than the
-attack in Beirut. In contrast to France, Lebanon had experienced dozens
-of terrorist bombings and attacks in recent years. Consequently, Beirut
-may seem to many like the sort of place where ``these things happen,''
-whereas Paris is perceived as being stable and safe.
+\acknow{We would like to acknowledge the help of Larisa Hussak, Keith Holyoak,
+Alan Fiske, Hongjing Lu, John Hummel, Andrei Cimpian, and Ellen Markman
+for their comments and support.}
 
-We can see in everyday experience that people's evaluations of events
-often depend on their expectations about those events. There is often
-disappointment when events fail to meet expectations, and there is a
-special thrill to having one's expectations exceeded. Anecdotally, these
-forces seem to drive people's tendencies to root for the underdog, hold
-surprise parties, and foreshadow bad news to ease its delivery (Bell,
-1985). Indeed, laboratory studies suggest that expectations play an
-important role in people's evaluations of the utility of an event. For
-instance, Mellers and colleagues (1997) found that expectations
-influenced affective reactions during a gambling task: Given a gamble
-with a 10\% chance to win \$30 and 90\% chance to win \$0, participants
-felt little disappointment with the \$0 outcome, but were considerably
-more excited when they won \$30. Conversely, given a gamble with a 90\%
-chance of winning \$60 and 10\% chance of winning \$0, participants were
-disappointed with the \$0 outcome and showed more muted enjoyment of the
-\$60 outcome. In fact, in gambles similar to these, Mellers and
-colleagues (1997) found that people were happier with the smaller
-unexpected gain than with the larger but more expected gain (also see
-Shepperd \& Mcnulty, 2002).
+When we learn of tragic events--a friend's loss of a loved-one, a
+violent killing, or a terrorist attack--we are mournful, concerned,
+angry, and driven to action. But our responses to these events differ
+depending on the context (1). In particular, our expectations play an
+important role in our reactions to events. This seems true of both the
+mundane--when a friend's loved-one passes, we often ask, ``was it
+sudden?''--the dramatic--a murder in a sleepy town prompts a very
+different reaction than a similar crime in a major city--and even the
+catastrophic: acts of terrorism seem to evoke more outrage, horror, and
+concern when they occurs in peaceful rather than unstable regions. For
+instance, the 2015 Paris attacks were felt around the world. Yet, most
+of those mourning had little to say 15 hours earlier, when another
+tragic attack killed at least 43 people in Beirut (2). Likewise, stories
+of the Virginia Tech massacre, which killed 32 people, wildly
+overshadowed coverage of attacks killing nearly 200 people in Iraq--one
+of the bloodiest days since the 2003 invasion (3). In each of these
+situations, the surprise elicited by an event appears to affect people's
+reactions to the event. In some contexts, harm is tragic and in others,
+``these things happen.''
 
-Just as expectations affect the utility of simple gambles, we suspect
-they are equally likely to shape how people react to morally harmful
-events, such as acts of terrorism (Waldmann Cite and Rai and Fiske cite
---domain general peeps; us also). However, unlike in the context of
-gambles, in these contexts the effects of expectations on evaluations
-may have harmful consequences. When events are shocking, we predict that
-people will perceive them as more severe and consequently be roused to
-action. In contrast, when events harm victims who are generally
-considered to be at greater risk--the poor, sick, or those living in
-unstable regions of the world, we predicted that reactions may be more
-muted. However, prior research has yet to articulate whether these same
-effects emerge in the moral domain and the implications of this
-possibility. For instance, it is unclear if these same mechanisms can
-help account for the divergent reactions people have to terrorism in
-France and Beirut. Further, people's moral evaluations are in fact
-impacted by expectations what is their source and what does that tell us
-about how to overcome the biases they may impose?
+Although this pattern seems so intuitive as to be obvious, no reasonable
+code of ethics would justify this pattern of judgment, nor do any moral
+psychological frameworks explicate the role of expectations in moral
+evaluations. Outside of the domain of morality, however, everyday
+experiences and controlled experiments show that people's evaluations of
+events unambiguously depend on their expectations about those events.
+There is a special thrill to having one's expectations exceeded and
+disappointment when events fail to meet expectations: we root for the
+underdog, hold surprise parties, and foreshadow bad news to ease its
+delivery (4). Indeed, in a controlled study, Mellers and colleagues (5)
+found that expectations influenced affective reactions during a gambling
+task: Given a gamble with a 10\% chance to win \$30 and 90\% chance to
+win \$0, participants felt little disappointment with the \$0 outcome,
+but were considerably more excited when they won \$30. Conversely, given
+a gamble with a 90\% chance of winning \$60 and 10\% chance of winning
+\$0, participants were disappointed with the \$0 outcome and showed more
+muted enjoyment of the \$60 outcome. In fact, in gambles similar to
+these, people were happier with the smaller unexpected gain than with
+the larger but more expected gain (6).
 
-A number of researchers have sought to develop theories of
-disappointment--the psychological reactions that result when experiences
-fail to meet expectations--and its role in evaluation and
-decision-making (e.g., Bell, 1985; Gul, 1991; Loomes \& Sugden, 1986).
-These researchers have generally argued that expectations modify
-people's reactions to events, so that they are jointly determined by the
-context-free utilities of options or events (e.g., economic utilities),
-and the contextually-dependent disappointment that individuals
-experience as a function of their expectations. Similarly, Mellers and
-colleagues (1997) interpreted the divergence they observed between the
-utility of an outcome (i.e., monetary gains or losses) and the felt
-experience of that outcome as a result of post-decision affect.
+Applied to moral issues, the influences of expectations on evaluation
+would be a dangerous bias. Tragedies are tragic, whether expected or
+not, and failure to recognize that could lead to tacit endorsement of
+the pain and suffering of others wherever pain and suffering are the
+status quo. Do expectations directly bias reactions to morally
+consequential events, as intuition and anecdote seem to suggest? To
+demonstrate the plausibility of this suggestion, we develop a
+theoretical account of how expectations influence evaluations, not as a
+strange quirk or uniquely human bias, but instead as a fundamental
+component of how people evaluate events.
 
-Each of these theories has been rigorously defined and supported within
-the small-world confines they were meant to capture. However, we seek to
-craft a more general explanation of how and why expectations influence
-evaluations and the range of contexts in which they should be expected
-to do so. Several accounts of utility evaluation, including early
-theories like Prospect Theory (Kahneman \& Tversky, 1979; Tversky \&
-Kahneman, 1992), have emphasized the role of relative comparisons in
-evaluations of utility. In a similar spirit, we argue that expectations
-can set the reference points against which people compare future
-outcomes. On this view, evaluating the utility of some event consists in
-comparing the state of the world now that the event has occurred to the
-state of the world just prior to the event. Further, we assume that our
-knowledge of the world is uncertain, so that it is really just our
-(probabilistic) expectations about what has happened, what is happening,
-and what will happen in the future. Thus, the evaluation of an event
-might be driven by what is learned about the state of the world as a
-result of that event's occurrence.
+Most events of consequence consist of changes--a transition from one
+state to another. For this reason, we construe event evaluation as a
+comparison between the states of the world before and after the event.
+Comparison plays a crucial role in many theories of decision-making, and
+especially in theories aimed at explaining the role of expectations in
+evaluation (4, 5, 7, 8). For instance, in Prospect Theory (9, 10), the
+utility of a prospect is evaluated by comparison to a current reference
+point. Our proposal is similar: we argue that expectations can set the
+reference points against which people compare future outcomes, so that
+assigning value to an event consists in comparing the state of the world
+following the event to the state of the world just prior to the event.
+As knowledge of the world is uncertain, our understanding of these
+states is probabilistic--that is, our model for the state of the world
+is just our expectations about what has happened, what is happening, and
+what will happen in the future. One direct implication of this proposal
+is that the evaluation of an event is linked to what is learned about
+the world as a result of that event's occurrence. That is, we evaluate
+events positively or negatively to the extent that they inform our
+understanding of the world in positive or negative ways.
 
-The account outlined above provides an information-theoretic route to
+This account thereby provides an information-theoretic route to
 understanding how expectations directly influence evaluation. A
 fundamental insight of Information Theory is that the information
-carried by an event is a function of its prior probability. This means
-that low probability events carry more information than high probability
-events (Shannon, 1948). Along similar lines, violations of expectations
-have long been recognized as fundamental to associative and animal
-learning models; a larger delta means people learn more (e.g., Rescorla
-\& Wagner, 1972). Science progresses most when new findings impugn
-widely-accepted theories or when a theory's extreme predictions are
-validated. Likewise, people learn more about the state of the world when
-their expectations are violated by shocking world events as compared to
-when they are affirmed by less surprising events. For example, if a
-bombing occurs in Paris -- an unexpected location -- rather than Lebanon
---a more expected location--we learn that the world is more dangerous
-than we had previously believed.
+carried by an event is a function of its prior probability; low
+probability events carry more information than high probability events
+(11). We propose that people learn more about the state of the world
+when their expectations are violated by shocking world events as
+compared to when they are affirmed by less surprising events. For
+example, if a bombing occurs in Paris--a less expected location--rather
+than Lebanon--a more expected location--we learn that the world is more
+dangerous than we had previously believed. This suggestion is novel in
+the moral domain, but these points are widely accepted elsewhere. For
+instance, expectations have long been recognized as fundamental to
+models of human and animal learning (12). Similarly, science progresses
+most when new findings impugn widely-accepted theories or when a
+theory's extreme predictions are validated.
 
-A more formal treatment of these issues reveals the generality of these
-conclusions. Formally, before some binary event \(x_i\) (which either
-occurs or does not occur), an evaluator has a mental model of the
-current state of affairs, S. This mental model includes expectations
-about potential future events and uncertainty over current and past
-events. Thus S represents a joint probability distribution over possible
-states of affairs \(p(x_1, x_2, ..., x_n)\). The evaluator's subjective
-utility function can assign an expected utility to S--representing how
-positive or negatively the evaluator views the current state of affairs.
-The occurrence of \(x_i\) brings about some new state of affairs, S', to
-which the evaluator can again apply her utility function. The evaluation
-of event \(x_i\) then consists in the comparison between these two
-states of affairs, those before the event (reflecting the agents'
-expectations), and the new state of affairs brought about by the event:
+A more formal treatment of this theory establishes the potential for its
+generality across domains. Let the state of the world prior to some
+event be represented by a random variable, \(X\). We'll assume \(X\) is
+discrete, and that every realization \(x\) represents some potential
+specific state of the world in all relevant aspects under consideration.
+An agent has some utility function \(u\), that applies over the
+different realizations of \(X\). Thus the expected utility of the
+present state of the world is:
 
-\[ V(x_i) = EU(S') -EU(S) \]
+\begin{equation}
+E_X[u(X)] = \sum_x u(x)p(x)
+\end{equation}
 
-We let S be represented by a random vector X, whose members are any
-number of discrete events \(x_1, x_2, ..., x_n\), where \(x_i\)
-represents the event of interest so that \(x_i\) occurs in S with some
-probability and occurs in S' with probability 1. That is, S represents
-\(X\) and S' represents \(X|x_i=1\). We write this as:
+When we learn an event has occurred, we can represent this as an
+observation of a second, binary random variable, \(S\). We can think of
+\(S\) as a ``signalling event'' that informs us about about \(X\) (13).
+That is, we assume there is a dependency between \(S\) and \(X\), so
+that observing \(S\) informs us about the state of the world, \(X\). For
+instance, reading a news report of a terror attack in Brussels informs
+us about the state of the world, telling us about what realizations of
+\(X\) are no longer possible (e.g., any wherein it was a peaceful day in
+Brussels) and which are more or less likely. Conditioned on \(S\), the
+expected utility of the world is:
 
-\[
-\begin{aligned}
-V(x_i) &= EU(X|x_i) - EU(X) \\
-&= \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x)
-\end{aligned}\]
+\begin{equation}
+E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
+\end{equation}
 
-Applying the chain rule of probability and some algebra (see appendix)
-we eventually obtain:
+Given our construal of event evaluation as the comparison between states
+of the world following and prior to the event, we define the value
+assigned to the signalling event, \(V(S)\), as:
 
-\[V(x_i) = \big(1 - p(x_i) \big) \sum_{x \in X} U(x)p(x|x_i)\]
+\begin{equation}
+V(S) = E_{X|S}[u(X)] - E_X[u(X)]
+\end{equation}
 
-Thus, the value assigned to event \(x_i\) is proportional to the prior
-probability of \(x_i\). Further, this holds true regardless of the
-nature of the event in question and of the form of the utility function
-over events.
+From this it can be shown that the value assigned to event \(S\) is
+proportional to its prior probability (see Theoretical Formalization).
+Specifically, it is the difference in the expected utility under
+different values of \(S\), weighted by the inverse of the prior
+probability of \(S\)..\footnote{It should be recognized that the result
+  we obtain here bears a similarity to the ``disappointment function''
+  included in Mellers and colleagues' (5) Decision Affect Theory.}
 
-In the context of our reaction to surprise parties, it might seem
-desirable that expectations would influence our experience of events.
-Much joy is tied to the psychological marriage between surprise and
-evaluation. However, this basic cognitive process may have negative
-moral repercussions. When harms are expected, these dynamics may lead
-people to draw less extreme evaluations, feel less concern for victims,
-and be less driven to help. To test this hypothesis, we examined whether
-people's evaluations of morally harmful events are affected by their
-expectations about those events. We asked people to compare pairs of
-simple events where a victim suffered an identical harm, but where the
-events differed in their prior probability. For each pair of events,
-participants were asked to judge which event was more upsetting. Two
-studies provided evidence for the predicted effect of expectations on
-moral evaluations: people tended to view unexpected negative events as
-more upsetting, even when the harm to victims was identical.
+\begin{equation}
+V(s=1)=(1-p(s=1))\sum_x u(x)p(x|s=1) - u(x)p(x|s=0)
+\end{equation}
 
-\section{General Methods}\label{general-methods}
+This conclusion does not depend on the nature of the event in question
+nor on the utility function over events, but instead holds generally
+wherever we conceive of evaluation as a holistic comparison between
+uncertain states of the world. Consequently, this account predicts
+expectations will equally affect reactions to gambles with clearly
+specified values and probabilities (5) and real-world events with
+ill-defined values and where reasoners' expectations are defined by
+their existing beliefs and world knowledge.
 
-Here we present two preregistered studies examining the role of
-expectations in moral evaluations. Study 1 used a forced-choice task to
-test the hypothesis that people would be more upset about moral outcomes
-that were unexpected than expected. In Study 2, we tested of our
-hypothesis using a more conservative judgment task to increase the
-generalizability of our results.
+Given the potentially fundamental role expectations play in event
+evaluation, and the potentially serious moral biases that may result, we
+sought to test empirically whether people's evaluations of morally
+harmful events are affected by their expectations about those events. We
+conducted two preregistered studies examining the role of expectations
+in moral evaluations. In both studies, participants were presented with
+a series of trials where they read brief descriptions of two different
+events and were asked to indicate which of the two events seemed more
+upsetting. In these trials, the two events were highly similar, but were
+manipulated so that they differed in their perceived prior
+probabilities: one event was more expected and one more unexpected.
 
-\subsection{Materials}\label{materials}
+The design of these studies was chosen with special consideration to two
+theoretical points. First, the information-theoretic account we have
+described makes a solely directional prediction--all else being equal,
+more surprising events will elicit stronger reactions than less
+surprising events. Asking participants to compare two similar events
+allowed us to test this directional prediction. Second, it is the
+informativeness of events, determined by one's prior expectations of the
+event, that we predict will shape evaluations. Thus, reasoners should
+not have to be induced to hold expectations experimentally (5), but
+instead their own prior beliefs will be sufficient to shape their
+expectations and affect their evaluations. For this reason, we
+manipulated the prior probability of events by describing two otherwise
+identical events occurring in different contexts. This allowed us to
+test the role of expectations based on world knowledge, as well as
+providing a more naturalistic set of items.
 
-In both studies, participants were presented with a series of trials
-where they read brief (one sentence) descriptions of two different
-events and were asked to indicate which of the two events seemed more
-upsetting. In ``experimental'' trials, the two events were highly
-similar, but differed in their prior probabilities: one event was more
-expected and one more unexpected. (The perceived likelihood of a given
-event was confirmed in prior norming studies). These prior expectations
-were manipulated by changing the context in which the events occurred.
-For example, participants considered the following stimulus:
+\begin{figure}
+\centering
+\includegraphics{fig1.eps}
+\caption{Probability with which participants chose the less expected
+event as more upsetting across trials in Study 1. As predicted,
+participants exhibited a robust bias toward choosing the less expected
+event as more upsetting across all items. Error bars represent 95\%
+credible intervals calculated from the hierarchical logistic regression
+model. See Supporting Information for full description of items. {}}
+\end{figure}
+
+Consistent with our predictions, manipulating expectations consistently
+shaped evaluations across a range of morally-significant events. We
+found that people viewed unexpected harmful events as more upsetting
+than expected harmful events, even when the harm a victim suffered was
+the same. For instance, people judged that a robbery at a clothing store
+(unexpected) was more upsetting than a robbery at a convenience store
+(expected). Furthermore, these differences were large, systematic, and
+robust. Although there was variability in the situations that elicited
+the largest differences, our statistical approach treated trial as a
+random effect (see Methods and Results), allowing us to generalize
+conclusions to items we did not test (14).
+
+\begin{figure}
+\centering
+\includegraphics{fig2.eps}
+\caption{Participants' responses by item in Study 2. Upper: heatmap plot
+displaying probability of choosing each response on the 1 to 5 scale
+across items. Lower: marginal expected response across items, with 95\%
+credible interval calculated from the ordinal regression model. Note the
+scale has been constrained to show relevant region of results. In both
+plots, response of 3 indicates indifference, a response below 3
+indicates a bias to choose the more expected event as more upsetting,
+and a response above 3 indicates a bias to choose the less expected
+event as more upsetting. Similar to Study 1, when participants judged
+that one of the events was more upsetting, they showed a robust bias
+toward choosing the less expected event as more upsetting. Trials are
+sorted by response to aid readability. {}}
+\end{figure}
+
+We have argued that expectations shape our reactions to events because
+they determine how informative those events are in shaping our
+understanding of the world. Thus, we argue that well-grounded
+information theoretic principles can account for people's divergent
+moral reactions to likely and unlikely events.
+
+There's a sense in which it is clearly adaptive that the violation of
+our expectations would direct our attention to the events and situations
+about which our understanding is most in need of revision. For instance,
+an act of terrorism in a peaceful nation might be less expected because
+of our beliefs about the effectiveness of security forces, the stability
+of national relationships, and positive relationships among people
+groups in that nation. Therefore, a surprising act of terrorism casts
+doubt on our deeper understanding of the geopolitical situation, and
+thereby leads us to suspect similar acts are more likely to occur in the
+future.
+
+Yet, however well-grounded the underlying principles, there's a clear
+moral danger whenever expectations influence our moral concerns. For
+instance, ``victim blaming'' describes the tendency to shift blame away
+from aggressors and onto victims of crimes like sexual assault (15) and
+police violence. Work by Janoff-Bulman and colleagues (16) has shown
+that merely increasing the perceived probability of a victim suffering a
+sexual assault induced participants to blame the victim for her assault.
+People's expectations may explain the flip side of victim blaming,
+whereby those who victim blame also tend to judge transgressors less
+harshly: When people perceive a victims' actions as increasing the
+probability of their eventual assault, they may perceive an aggressor's
+actions as producing a smaller change in utility, thus warranting weaker
+condemnation.
+
+The influence of expectations on moral evaluations poses perhaps the
+greatest threat when considering our reactions to humanitarian
+challenges brought on by poverty or regional instability. In these
+cases, the influence of expectations on moral evaluations threatens to
+impose vicious and morally pernicious cycles. For instance, people
+living in geo-politically unstable regions or in the developing world
+are often those who are most affected by terrorism, famine, and natural
+disasters, and are the very people in greatest need of assistance and
+concern from the world at-large. However, for these very reasons, it is
+often unsurprising when harm befalls people living in these
+circumstances. Thus the influence of expectations on moral evaluations
+threatens to reduce observer's concern for these victims most in need,
+making people less likely to donate time or money to aid victims, or to
+take political action to prevent further harms. The present research
+identifies this potentially pernicious moral bias and grounds it in
+information theoretic principles. Hopefully, by further understanding
+these cognitive dynamics, psychologists and policymakers can begin to
+find ways to reduce or counteract people's tendency to ignore the plight
+of those in most need of attention.
+
+\section*{Materials and Methods}\label{materials-and-methods}
+\addcontentsline{toc}{section}{Materials and Methods}
+
+We conducted two preregistered studies examining the role of
+expectations in moral evaluations using two different judgment tasks
+(\url{https://osf.io/5qnf7/}). Each study consisted of multiple trials
+where participants were presented brief descriptions of two events.
+Their task was to decide which of the two events was more upsetting. In
+Study 1, they made a two-alternative forced-choice, and in Study 2 they
+made their judgments using a five-point rating scale. Both studies
+consisted of 16 experimental trials, five ``equivalent'' filler trials,
+and five ``non-equivalent'' filler trials.
+
+In experimental trials, the two events described identical harms, but
+one event was more expected and one more unexpected, as determined in a
+prior norming study (see SI). For example, participants considered the
+following stimulus:
 
@@
 
-In each event, the harm to the victim is the same (here, death) but one
-event is more expected than the other, given the different likelihoods
-of earthquakes occurring in California versus Oklahoma.
+In each of these events, the harm to the victim is the same (death) but
+one event is more probable than the other, given the different
+likelihoods of earthquakes occurring in California versus Oklahoma. In
+addition to being more naturalistic, manipulating expectations through
+(otherwise irrelevant) contextual details helps ensure that the harm
+suffered by the victim is understood to be identical. In the context of
+gambles, it is possible to manipulate probabilities specifically and
+directly (5). However, in the context of more realistic moral harms, any
+attempt to directly manipulate probabilities risks also changing
+participants' perceptions of the degree of harm. For instance, suppose
+we describe two cancer patients, one with a 5\% chance of survival and
+one with a 50\% chance of survival. Even if they both ultimately meet
+the same fate, participants might fairly assume the patient with the
+lower chance of survival had a more serious illness, was more
+debilitated, suffered more severely, and so forth. In contrast, changing
+the context of the two events explains away these likelihood
+differences, better equating the events in other respects.
 
-Each study contained between 16 experimental event-pairs that spanned a
-variety of different events and contexts. All experimental materials for
-these studies are available as supplemental online materials at {[}OSF
-LINK{]}.
+Both studies also included ``equivalent'' filler trials. In these
+trials, the two events differed in trivial contextual details that we
+did not expect would affect participants' judgments. These filler trials
+were meant to prevent participants from becoming explicitly aware of the
+structure of the experimental trials. Finally, both studies included
+``non-equivalent'' filler trials, the two events differed substantially
+in the degree of harm suffered by a victim, so that one event was
+expected to be seen as considerably more upsetting than the other. These
+trials were included to allow participants a chance to use the extremes
+of the response scale and to reduce any task demands that might drive
+them to make artificially fine-grained distinctions when generating
+their responses. All experimental materials for these studies are
+available as Supporting Information at \url{https://osf.io/5qnf7/}.
 
-Both studies included ``equivalent'' filler trials. In these trials, the
-two events differed in trivial contextual details that we did not expect
-would affect participants' judgments. For example:
+\subsection*{Exclusions}\label{exclusions}
+\addcontentsline{toc}{subsection}{Exclusions}
 
-\begin{itemize}
-\item
-  ``A man in Connecticut starts a house fire.'' {[}Equally expected{]}
-\item
-  ``A man in New Hampshire starts a house fire.'' {[}Equally expected{]}
-\end{itemize}
+In both studies, we excluded participants who failed to correctly answer
+attention-check questions or who indicated at the end of the study they
+had not taken their participation seriously.
 
-These filler trials were meant to prevent participants from becoming
-explicitly aware of the structure of the experimental trials. Finally,
-both studies included ``non-equivalent'' filler trials, the two events
-differed substantially in the degree of harm suffered by a victim, so
-that one event was expected to be seen as considerably more upsetting
-than the other. For example:
-
-\begin{itemize}
-\item
-  ``An 11-year-old child sets a doll on fire'' {[}Less severe{]}
-\item
-  ``A 12-year-old child sets a cat on fire'' {[}More severe{]}
-\end{itemize}
-
-These trials were included to allow participants a chance to use the
-extremes of the response scale and to reduce any task demands that might
-drive them to make artificially fine-grained distinctions between the
-severity of events.
-
-\subsection{Exclusions}\label{exclusions}
-
-In both studies, we excluded participants who failed to incorrectly
-answer attention-check questions. These questions asked participants to
-enter a particular response to ensure that they were paying attention
-and reading the items as they proceeded through the study. A final
-question asked participants if they had paid attention and taken the
-study seriously, encouraging them to be honest in their replies.
-
-\subsection{Data Analysis}\label{data-analysis}
+\subsection*{Data Analysis}\label{data-analysis}
+\addcontentsline{toc}{subsection}{Data Analysis}
 
 We analyzed our data by performing Bayesian estimation using the
-probabilistic programming language Stan (Carpenter et al., 2017). We
-tested our predictions by computing Bayes Factors (i.e.~BF01) on the
-intercept term of our regression model using the hypothesis function in
-the R package brms. Bayes Factors express the ratio of the probability
-of data under the null hypothesis to the probability of the data under
-an alternative hypothesis. Therefore, larger Bayes Factors indicate that
-the data are more likely under the null hypothesis (e.g., that the
-intercept is not different from zero) than the alternative hypothesis
-(e.g., that the intercept is different from zero), and vice versa. Bayes
-Factors can be influenced by prior choices (Gelman, Simpson, \&
-Betancourt, 2017) so we also performed prior robustness checks to
-confirm that our conclusions were unchanged under different prior
-assumptions.
+probabilistic programming language Stan (17). We tested our
+preregistered predictions by computing Bayes Factors (i.e.~BF01) using
+the hypothesis function in the R package brms. Bayes Factors express the
+ratio of the probability of data under the null hypothesis to the
+probability of the data under an alternative hypothesis. Larger Bayes
+Factors indicate that the data are more likely under the null hypothesis
+(i.e., that the intercept is not different from zero) than the
+alternative hypothesis (i.e., that the intercept is different from
+zero), and vice versa. As Bayes Factors can be influenced by prior
+choices (Gelman, Simpson, \& Betancourt, 2017), we performed prior
+robustness checks to confirm that our conclusions would not depend
+substantially on the specification of the priors.
 
-\section{Appendix}\label{appendix}
+To ensure the generality of our findings, we tested participants'
+reactions to a variety of items and treated both participants and items
+as random effects throughout our analyses. The use of random effects
+over items allows us to consider these items as a sample from a larger
+population of possible items, and to generalize our findings to that
+wider population, much as we generalize from our sample of participants
+to the larger population (14).
 
-\[ \begin{aligned}
-V(x_i) &= EU(X|x_i) - EU(X) \\
-&=  \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x) \\
-&= \sum_{x \in X} U(x)p(x|x_i) - \sum_{x \in X} U(x)p(x|x_i)p(x_i) \\
-&= \sum_{x \in X} U(x)p(x|x_i) - p(x_i)\sum_{x \in X} U(x)p(x|x_i) \\
-&= \big(1 - p(x_i) \big) \sum_{x \in X} U(x)p(x|x_i) \\
-\end{aligned}\]
+\subsection*{Study 1}\label{study1}
+\addcontentsline{toc}{subsection}{Study 1}
 
-The first step utilizes the Law of the Unconscious Statistician, the
-next comes from the chain rule of probability, and the final two steps
-are simple algebra.
+\subsubsection*{Participants}\label{s1-participants}
+\addcontentsline{toc}{subsubsection}{Participants}
 
-\section{References}\label{references}
+A total of 255 participants were recruited from Amazon's Mechanical Turk
+work distribution website (mTurk). Of these, 230 passed attention checks
+and were included in the final analyses (105 male, 125 female, median
+age = 33 years old). All participants were paid \$0.60 for their
+participation.
 
-\setlength{\parindent}{-0.25in} \setlength{\leftskip}{0.125in} \noindent
+\subsubsection*{Materials and procedure}\label{s1-materials-procedure}
+\addcontentsline{toc}{subsubsection}{Materials and procedure}
+
+Participants judged 16 experimental event-pairs and 10 equivalent filler
+event-pairs. On each trial (an example is shown above), participants
+were presented with the event-pair stimulus and had to judge which
+outcome was more upsetting in a two-alternative forced choice task. The
+two events were labeled ``Outcome 1'' and ``Outcome 2'' and the order,
+and whether Outcome 1 or 2 was the unexpected event, was randomized and
+counter-balanced.
+
+\subsubsection*{Results and discussion}\label{s1-results}
+\addcontentsline{toc}{subsubsection}{Results and discussion}
+
+We predicted that people would judge that events where unexpected harm
+occurred were more upsetting than events where expected harm occurred.
+As indicated in Figure 1, we observed this trend. We estimated the
+magnitude of this effect by fitting a Bayesian logistic random effects
+model with participants' responses as the dependent variable (1 =
+unexpected event is more upsetting; 0 = expected event is more
+upsetting) and random intercepts for item and subject, allowing us to
+generalize both to items and subjects we did not test. The intercept in
+this model represents the log-odds of selecting the unexpected event as
+being more upsetting. Thus, by examining the population-level intercept,
+we can test whether participants were biased toward selecting the
+unexpected event (\(\beta\) \textgreater{} 0), the expected event
+(\(\beta\) \textless{} 0), or were unbiased (\(\beta\) = 0). We found
+that people were considerably more likely to think that unexpected
+events were more upsetting than events that were expected, Intercept =
+1.424, 95\% CI {[}1.147, 1.707{]}, BF01 \textless{} .001. Figure 1 shows
+participants' responses broken-down by individual items. Participants'
+bias toward selecting the unexpected event as more upsetting was
+consistent across the 16 experimental items.
+
+\subsection*{Study 2}\label{study2}
+\addcontentsline{toc}{subsection}{Study 2}
+
+In Study 1 we found that people's judgments of events were biased by
+their expectations about those events. When forced to choose between two
+events, participants decided that unexpected events were more upsetting
+than expected events. In Study 2, we sought to test our hypothesis using
+a more conservative method: we provided participants with a more
+expressive response scale so that if they viewed the events under
+consideration as equally harmful, their responses could reflect their
+attitude.
+
+\subsubsection*{Participants}\label{s2-participants}
+\addcontentsline{toc}{subsubsection}{Participants}
+
+A total of 504 participants were recruited from Amazon's Mechanical Turk
+work distribution website (mTurk). Of these, 372 passed attention checks
+and were included in the final analyses (174 male, 198 female, median
+age = 34 years old). All participants were paid \$0.60 for their
+participation.
+
+\subsubsection*{Materials and procedure}\label{s2-materials-procedure}
+\addcontentsline{toc}{subsubsection}{Materials and procedure}
+
+The materials in Study 2 were the same as those in Study 1. As in Study
+1, on each trial of the study, participants were presented with a pair
+of actions labeled ``Outcome 1'' and ``Outcome 2'' and were asked,
+``Which outcome seems more upsetting?'' In Study 2 participants made
+their rating on a five-point scale (Outcome 1 seems more upsetting,
+Outcome 1 seems a little more upsetting, neither seems more upsetting
+than the other, Outcome 2 seems a little more upsetting, Outcome 2 seems
+more upsetting).
+
+\subsubsection*{Results and discussion}\label{s2-results}
+\addcontentsline{toc}{subsubsection}{Results and discussion}
+
+We predicted that participants would evaluate unexpected events as more
+upsetting than expected events. We estimated the magnitude of this
+effect by fitting a cumulative (ordinal) random effects model with
+participants' scale responses as the dependent variable (1 to 5) and
+random intercepts for item and subject. This model produces four
+intercept coefficients, representing the cumulative log-odds of
+responses at each scale point or lower. For instance, the second
+coefficient represents the log-odds participants chose a 2 (``outcome 2
+seems slightly more upsetting'') or lower on the scale. Similarly, the
+third intercept coefficient represents the log-odds participants chose a
+3 or lower on the scale. To examine the predicted effect, we compared
+the third intercept coefficient (representing the log-odds of choosing 4
+or higher) to the inverse of the second intercept coefficient (thereby
+representing the log-odds \emph{not} choosing a 3 or higher--i.e.,
+choosing a 1 or 2), allowing us to test whether participants were more
+likely to choose the expected or unexpected event as being more
+upsetting in cases where they did not choose the neither option. This
+analysis indicated that people were more likely to think that events
+that were unexpected were more upsetting than events that were expected,
+BF01 \textless{} .001 (see supporting information for full model
+results). We observed this difference consistently across nearly every
+item (Figure 2).
+
+We note that the most common response was to view both events, which
+involve identical harms, as equivalently upsetting. However, when
+participants did view one of the events as more upsetting, they showed a
+clear bias towards choosing unexpected events as more upsetting. Thus,
+Study 2 replicates the tendency to view unexpected moral events as more
+upsetting using a more conservative judgment task.
+
+\section*{Theoretical Formalization}\label{theory}
+\addcontentsline{toc}{section}{Theoretical Formalization}
+
+Evaluation (e.g., of utility) is often examined in the context of
+decision-making. However, people often generate post hoc evaluations of
+events for their own sake. For instance, to reflect on the results of
+our choices and to evaluate others--to judge the wisdom of their
+decisions or the morality of their actions.
+
+We construe event evaluation as a comparison between the state of
+affairs prior to an event and the state of affairs following the event.
+We represent the state of the world as a random variable, \(X\). We
+assume \(X\) is discrete, and that every realization \(x\) represents
+some potential specific state of the world in all relevant aspects under
+consideration. An agent has some utility function \(u\), that applies
+over the different realizations of \(X\). Thus the expected utility of
+the present state of the world is:
+
+\begin{equation}
+E_X[u(X)] = \sum_x u(x)p(x)
+\end{equation}
+
+When we learn an event has occurred, we can represent this as an
+observation of another random variable, \(S\). We can think of \(S\) as
+a ``signalling event'' because it informs (signals) to us something
+about \(X\) (13). For simplicity, we treat \(S\) as binary, either 1 or
+0. Importantly, we assume there is a dependency between \(S\) and \(X\),
+so that observing \(S\) informs our knowledge of \(X\). For instance,
+reading a news report of a terror attack in Brussels informs us about
+the state of the world, telling us about what realizations of X are no
+longer possible (e.g., any wherein it was a peaceful day in Brussels)
+and which are more or less likely. Conditioned on \(S\), the expected
+utility of the world is:
+
+\begin{equation}
+E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
+\end{equation}
+
+We define the value assigned to the signalling event, \(V(S)\), as:
+
+\begin{equation}
+V(S) = E_{X|S}[u(X)] - E_X[u(X)]
+\end{equation}
+
+Which we write more specifically as:
+
+\begin{equation}
+V(s=1) = \sum_x u(x)p(x|s=1) - \sum_x u(x)p(x)
+\end{equation}
+
+Applying the expectation of conditional expectation, we can write:
+
+\begin{equation}
+V(s=1) = \sum_x u(x)p(x|s=1) - \sum_x \sum_s u(x)p(x|s)p(s)
+\end{equation}
+
+We then perform some algebra. We break out the summation over \(S\) (6),
+pull the multipliers out of the last two sums (7), rewrite \(p(s=0)\) in
+terms of \(p(s=1)\) (8), and factor to combine the first two sums (9).
+
+\begin{align}
+V(s=1) &= \sum_x u(x)p(x|s=1) - \sum_x u(x)p(x|s=1)p(s=1) - \sum_x u(x)p(x|s=0)p(s=0) \\
+ &= \sum_x u(x)p(x|s=1) - p(s=1)\sum_x u(x)p(x|s=1) - p(s=0) \sum_x u(x)p(x|s=0) \\
+ &= \sum_x u(x)p(x|s=1) - p(s=1)\sum_x u(x)p(x|s=1) - (1-p(s=1)) \sum_x u(x)p(x|s=0) \\
+ &=(1-p(s=1))\sum_x u(x)p(x|s=1) - (1-p(s=1)) \sum_x u(x)p(x|s=0)
+\end{align}
+
+Finally, combining both sums by their common factor yields (10):
+
+\begin{equation}
+V(s=1)=(1-p(s=1))\sum_x u(x)p(x|s=1) - u(x)p(x|s=0)
+\end{equation}
+
+Thus, it can be seen under this formalization that the role of
+expectations in the evaluation of events is fundamental--it does not,
+for instance, depend on any specifics concerning the utility function
+over states of the world. Instead, this conclusion depends on just two
+assumptions: first, that events are evaluated by comparing the utilities
+of the states of the world prior to and following the event; and second,
+that knowledge of those states of the world is uncertain.
+
+We have appealed to the informativeness of an event to illustrate
+intuitively how expectations should influence evaluations. More expected
+events carry less information, and thereby engender weaker reactions.
+Though our model of event evaluation is not an information-theoretic
+model in a direct sense, this interpretation still applies: the
+information carried by observing \(S\) is a function of that
+observation's prior probability, which we have in turn shown determines
+its evaluation. Also relevant, although not addressed here, is the
+mutual information between \(X\) and \(S\)--the extent to which \(S\)
+informs \(X\).
+
+\showmatmethods
+\showacknow
+\pnasbreak
 
 \hypertarget{refs}{}
+\hypertarget{ref-Trope2010}{}
+1. Trope Y, Liberman N (2010) Construal-level theory of psychological
+distance. \emph{Psychological review} 117(2):440--63.
+
+\hypertarget{ref-Graham2015}{}
+2. Graham DA (2015) The empathy gap between paris and beirut.
+
+\hypertarget{ref-Greenslade2007}{}
+3. Greenslade R (2007) A hierarchy of death.
+
 \hypertarget{ref-Bell1985}{}
-Bell, D. (1985). Disappointment In Decision Making Under Uncertainty.
-\emph{Operations Research}, \emph{33}(1), 1--27.
-\url{http://doi.org/10.1287/opre.33.1.1}
+4. Bell D (1985) Disappointment In Decision Making Under Uncertainty.
+\emph{Operations Research} 33(1):1--27.
 
-\hypertarget{ref-Brewer1999}{}
-Brewer, M. B. (1999). The psychology of prejudice: Ingroup love or
-outgroup hate? \emph{Journal of Social Issues}, \emph{55}(3), 429--444.
-\url{http://doi.org/10.1111/0022-4537.00126}
+\hypertarget{ref-Mellers1997}{}
+5. Mellers B a, Schwartz a, Ho K, Ritov I (1997) Decision Affect Theory:
+Emotional Reactions to the Outcomes of Risky Options.
+\emph{Psychological Science} 8(6):423--429.
+
+\hypertarget{ref-Shepperd2002}{}
+6. Shepperd J a, Mcnulty JK (2002) The affective consequences of
+expected and unexpected outcomes. \emph{Psychological science}
+13(1):85--88.
+
+\hypertarget{ref-Gul1991}{}
+7. Gul F (1991) A theory of disappointment aversion. \emph{Econometrica}
+59(3):667--686.
+
+\hypertarget{ref-Loomes1986}{}
+8. Loomes G, Sugden R (1986) Disappointment and Dynamic Consistency in
+Choice under Uncertainty. \emph{Review of Economic Studies}
+53(2):271--282.
+
+\hypertarget{ref-Kahneman1979}{}
+9. Kahneman D, Tversky A (1979) Prospect Theory: An Analysis of Decision
+under Risk. \emph{Econometrica} 47(2):263--292.
+
+\hypertarget{ref-Tversky1992}{}
+10. Tversky A, Kahneman D (1992) Advances in prospect theory: Cumulative
+representation of uncertainty. \emph{Journal of Risk and uncertainty}
+5:297--323.
+
+\hypertarget{ref-Shannon1948}{}
+11. Shannon CE (1948) A Mathematical Theory of Communication. \emph{Bell
+System Technical Journal} 27(3):379--423.
+
+\hypertarget{ref-Rescorla1972}{}
+12. Rescorla RA, Wagner AR (1972) A theory of Pavlovian conditioning:
+Variations in the effectiveness of reinforcement and nonreinforcement.
+\emph{Classical Conditioning II Current Research and Theory}, pp 64--99.
+
+\hypertarget{ref-Arrow1996}{}
+13. Arrow KJ (1996) The economics of information: An exposition.
+\emph{Empirica} 23(2):119--128.
+
+\hypertarget{ref-Judd2012}{}
+14. Judd CM, Westfall J, Kenny DA (2012) Treating stimuli as a random
+factor in social psychology: A new and comprehensive solution to a
+pervasive but largely ignored problem. \emph{Journal of Personality and
+Social Psychology} 103(1):54--69.
+
+\hypertarget{ref-Koss2000}{}
+15. Koss MP (2000) Blame, shame, and community: Justice responses to
+violence against women. \emph{American Psychologist} 55(11):1332--1343.
+
+\hypertarget{ref-Janoff-bulman1985}{}
+16. Janoff-Bulman R, Timko C, Carli LL (1985) Cognitive biases in
+blaming the victim. \emph{Journal of Experimental Social Psychology}
+21:161--177.
 
 \hypertarget{ref-Carpenter2017}{}
-Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B.,
-Betancourt, M., \ldots{} Riddell, A. (2017). Stan: A probabilistic
-programming language. \emph{Journal of Statistical Software},
-\emph{76}(1). \url{http://doi.org/10.18637/jss.v076.i01}
+17. Carpenter B, et al. (2017) Stan : A Probabilistic Programming
+Language. \emph{Journal of Statistical Software} 76(1).
+doi:\href{https://doi.org/10.18637/jss.v076.i01}{10.18637/jss.v076.i01}.
 
-\hypertarget{ref-Gul1991}{}
-Gul, F. (1991). A theory of disappointment aversion.
-\emph{Econometrica}, \emph{59}(3), 667--686.
-\url{http://doi.org/10.2307/2938223}
-
-\hypertarget{ref-Kahneman1979}{}
-Kahneman, D., \& Tversky, A. (1979). Prospect Theory: An Analysis of
-Decision under Risk. \emph{Econometrica}, \emph{47}(2), 263--292.
-
-\hypertarget{ref-Loomes1986}{}
-Loomes, G., \& Sugden, R. (1986). Disappointment and Dynamic Consistency
-in Choice under Uncertainty. \emph{Review of Economic Studies},
-\emph{53}(2), 271--282. \url{http://doi.org/10.2307/2297651}
-
-\hypertarget{ref-Mellers1997}{}
-Mellers, B. a, Schwartz, a., Ho, K., \& Ritov, I. (1997). Decision
-Affect Theory: Emotional Reactions to the Outcomes of Risky Options.
-\emph{Psychological Science}, \emph{8}(6), 423--429.
-\url{http://doi.org/10.1111/j.1467-9280.1997.tb00455.x}
-
-\hypertarget{ref-Rescorla1972}{}
-Rescorla, R. A., \& Wagner, A. R. (1972). A theory of Pavlovian
-conditioning: Variations in the effectiveness of reinforcement and
-nonreinforcement. In \emph{Classical conditioning ii current research
-and theory} (Vol. 21, pp. 64--99).
-\url{http://doi.org/10.1101/gr.110528.110}
-
-\hypertarget{ref-Shannon1948}{}
-Shannon, C. E. (1948). A Mathematical Theory of Communication.
-\emph{Bell System Technical Journal}, \emph{27}(3), 379--423.
-\url{http://doi.org/10.1002/j.1538-7305.1948.tb01338.x}
-
-\hypertarget{ref-Shepperd2002}{}
-Shepperd, J. a, \& Mcnulty, J. K. (2002). The affective consequences of
-expected and unexpected outcomes. \emph{Psychological Science},
-\emph{13}(1), 85--88. \url{http://doi.org/10.1111/1467-9280.00416}
-
-\hypertarget{ref-Tversky1992}{}
-Tversky, A., \& Kahneman, D. (1992). Advances in prospect theory:
-Cumulative representation of uncertainty. \emph{Journal of Risk and
-Uncertainty}, \emph{5}, 297--323.
+% Bibliography
+% \bibliography{pnas-sample}
 

### 2018-07-03 SuppInfo.Rmd
updating pre-submission

---
title: "Supporting Information"
output:
  pdf_document: default
  html_notebook: default
---

```{r, include=FALSE}
library(tidyverse)

df_raw <- read.csv("data/WIW_Norming_for_preregistered_study.csv")

reverse_items <- c(
  "Earthquake",
  "Tornado",
  "Fastfood",
  "Marine",
  "Police",
  "Mugging",
  "Crushed",
  "Dysentery",
  "Drunk",
  "Football"
)

df_wide <- df_raw %>%
  filter(Status==0) %>%
  filter(Check_trial == 5) %>%
  select(-Check_trial) %>%
  mutate(sex = if_else(sex == 1, "male", "female"))

df <- df_wide %>%
  gather(trial, response, Earthquake:Rape) %>%
  mutate(response = ifelse(trial %in% reverse_items, 6-response, response))
```
# Item Norming

We created and normed 20 items, each consisting of a pair of events that differed in their prior likelihood. The two events (outcomes) were manipulated so what one would see more expected than the other. Participants (_N_ = `r nrow(df_wide)`) were asked to choose which event seemed more likely using a 5-point scale (1 = Event 1 seems a lot more likely, 2 = Event 1 seems a little more likely, 3 = Neither event seems more likely, 4 = Event 2 seems a little more likely, 5 = Event 2 seems much more likely). As predicted, for all items the "more expected" outcome was judged to be more likely. Plot below shows mean likelihood rating by item along with approximate 95% confidence intervals.

```{r, echo=FALSE}
# standard error function
stderr <- function(x) {
          sqrt(var(x[!is.na(x)]) / length(x[!is.na(x)]))
}

df %>%
  group_by(trial) %>%
  summarize(
    Mean = mean(response),
    UL = mean(response) + 1.96*stderr(response),
    LL = mean(response) - 1.96*stderr(response)
    ) %>%
  ggplot() +
  aes(x = reorder(trial, -Mean), y = Mean, ymin = LL, ymax=UL) +
  geom_pointrange() +
  geom_hline(yintercept = 3, linetype="dashed") +
  coord_cartesian(ylim=c(1,5)) +
  theme_bw()+
  labs(x = "Trial", y="Mean Likelihoods Judgment") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

From among these 20 items we selected 16 items for use in our experiments.

\newpage
## Items

Below we provide the full text of items as presented to participants.

### Experimental Trials

| Trial      | More Expected                                                | Less Expected                                                |
| ---------- | ------------------------------------------------------------ | ------------------------------------------------------------ |
| Tornado    | A family in Nebraska has their home destroyed by a tornado.  | A family in Oregon has their home destroyed by a tornado.    |
| Marine     | A young US marine is shot in the head and killed by friendly fire on patrol outside Baghdad. | A young US marine is shot in the head and killed by friendly fire during a training exercise. |
| Mugging    | A man in downtown Detroit is mugged.                         | A man in downtown Denver is mugged.                          |
| Crushed    | A worker is crushed to death due to an accident at a factory in India. | A worker is crushed to death due to an accident at a factory in Japan. |
| Dysentery  | A pair of travelers contract dysentery after drinking water in Mexico. | A pair of travelers contract dysentery after drinking water in Arizona. |
| Cancer     | A 60 year old woman dies from breast cancer.                 | A 60 year old man dies from breast cancer.                   |
| Drunk      | A pedestrian is killed by a drunk driver on a Friday night.  | A pedestrian is killed by a drunk driver on a Saturday morning. |
| Football   | A 16 year old sustains a major concussion while playing in a high school football game. | A 16 year old sustains a major concussion while playing in a high school basketball game. |
| Heater     | A woman on disability dies when her 15 year-old heater fails during winter. | A woman on disability dies when her 3 year-old heater fails during winter. |
| Flattire   | A man repairing a flat tire on the side of the highway is struck by a car going 50 mph. | A man repairing a flat tire on the side of a suburban street is struck by a car going 50 mph. |
| Clerk      | A clerk is held up at gunpoint in a convenience store.       | A clerk is held up at gunpoint in a clothing store.          |
| Corruption | A corrupt government official in Mexico takes a bribe that allows a company to pollute the environment. | A corrupt government official in Canada takes a bribe that allows a company to pollute the environment. |
| Rape       | A woman is raped leaving a bar.                              | A woman is raped leaving a coffee shop.                      |
| Earthquake | A 30 year old man in California dies in an earthquake.       | A 30 year old man in Oklahoma dies in an earthquake.         |
| Fastfood   | A 32 year old woman gets food poisoning after eating a hamburger at a fast food restaurant. | A 32 year old woman gets food poisoning after eating a hamburger at a four star restaurant. |
| Police     | A criminal shoots and wounds a police officer while the officer is responding to a 911 call reporting a robbery in progress. | A criminal shoots and wounds a police officer while the officer is responding to a 911 call reporting vandalism. |

\newpage

### Filler Trials

| Trial    | Outcome 1                                                    | Outcome 2                                                    |
| -------- | ------------------------------------------------------------ | ------------------------------------------------------------ |
| Check1   | This item tests if you are paying attention                  | Please choose event 2 seems a little more upsetting          |
| Check2   | Someone who is paying attention will select event two        | Please select event two seems a lot more upsetting.          |
| Check3   | A participant who is paying attention will select event one  | Please select event one seems a little more upsetting.       |
| NoDiff1  | A man in Connecticut dies in a house fire.                   | A man in New Hampshire dies in a house fire.                 |
| NoDiff2  | A family in Florida is evicted from their home after they run into financial troubles. | A family in Louisiana is evicted from their home after they run into financial troubles. |
| NoDiff3  | After an accident, a machinist has to have his foot amputated. | After an accident, a metalworker has to have his foot amputated. |
| NoDiff4  | A girl skins her knees skateboarding.                        | A girl skins her knees rollerblading.                        |
| NoDiff5  | A large tree branch falls and injures a couple walking home from a baby shower. | A large tree branch falls and injures a couple walking home from their church. |
| BigDiff1 | A cat hurts its leg jumping off the refrigerator.            | A cat dies after eating spoiled food.                        |
| BigDiff2 | A parked car is damaged in a parking lot in Oregon.          | A car is totaled on the highway in Washington.               |
| BigDiff3* | \$1000 worth of products from an electronics store are destroyed in a shipping accident. | \$200 worth of products from a clothing store are burned in a shipping accident. |
| BigDiff4 | A knife is found in a high school student's locker.          | A gun is found in a high school student's gym bag.           |
| BigDiff5 | A 30 year-old man is killed in a car crash.                  | A 35 year-old man is injured in a cycling accident.          |

_* As presented in Study 2. In Study 1, this concerned a construction worker having his left versus right foot amputated, but participants judged having one's right foot amputated was somewhat worse than having one's left foot amptuated._

\newpage

## Discarded items

These items were included in the norming study but were not ultimately selected for the final study. "Heroin" and "theft" showed the weakest effects on likelihood judgments. We decided that the damage to property in "Scratched" was not sufficiently moral in nature. Finally, "brokeninto" was discarded for being too similar to "flattire" which we ultimately decided had more significant moral content.

| Trial      | Event 1                                                      | Event 2                                                      |
| ---------- | ------------------------------------------------------------ | ------------------------------------------------------------ |
| heroin     | A suburban teenage boy dies of a heroin overdose.            | A homeless teenage boy dies of a heroin overdose.            |
| theft      | A man in training steals \$50,000 from a small business he recently began working for. | A man on probation steals $50,000 from a small business he recently began working for. |
| scratched  | A woman's car is scratched in a quiet parking lot.           | A woman's car is scratched in a busy parking lot.            |
| brokeninto | Three cars on the same street are broken into while parked in the suburbs. | Three cars on the same street are broken into while parked downtown. |

### 2018-07-03 paper/paper.Rmd
updating pre-submission

@@
 significance: |
-  People living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and thus are the very people in greatest need of assistance and concern from the world at-large.
-However, for these very reasons, it is often unsurprising when harm befalls them.
-Two experiments confirm this intuition: we found that people's expectations bias their reactions to morally-significant events--people are more upset by surprising harms than more expected harms.
-This bias threatens to reduce observer's concern for victims most in need, making people less likely to donate time or money to aid victims, or to take political action to prevent further harms.
+  In two behavioral studies, we found that people's expectations biased their reactions to morally-significant events--people are more upset by surprising harms than more expected harms.
+This bias appears deeply pernicious: because it is often unsurprising when people living in poverty, dangerous neighborhoods, or unstable regions become victims of crime, natural disasters, or terrorism, observers are likely to experience reduced concern for these victims.
+Yet people living in disadvantaged circumstances are those most in need of assistance and concern from the world at-large.
+By blunting people's concern for those victims who are most in need, the bias we have uncovered threatens to increase their suffering and further sustain inequality.
 
@@
 ## change to true to add optional line numbering
-lineno: TRUE
+lineno: FALSE
 
@@
 When we learn of tragic events--a friend's loss of a loved-one, a violent killing, or a terrorist attack--we are mournful, concerned, angry, and driven to action.
-But our responses to these events differ depending on the context [e.g., @Trope2010].
+But our responses to these events differ depending on the context in which these events occur [e.g., @Trope2010].
 In particular, our expectations play an important role in our reactions to events.
@@
 Although this pattern seems so intuitive as to be obvious, no reasonable code of ethics would justify this pattern of judgment, nor do any moral psychological frameworks explicate the role of expectations in moral evaluations.
-Outside of the domain of morality, however, everyday experiences and controlled experiments show that people’s evaluations of events unambiguously depend on their expectations about those events.
+Outside of the domain of morality, everyday experiences and controlled experiments show that people’s evaluations of events unambiguously depend on their expectations about those events.
 There is a special thrill to having one’s expectations exceeded and disappointment when events fail to meet expectations: we root for the underdog, hold surprise parties, and foreshadow bad news to ease its delivery [@Bell1985].
@@
 
-Applied to moral issues, the influences of expectations on evaluation would be a dangerous bias.
+The influence of expectations on moral evaluations threatens to produce negative social consequences.
 Tragedies are tragic, whether expected or not, and failure to recognize that could lead to tacit endorsement of the pain and suffering of others wherever pain and suffering are the status quo.
@@
 Our proposal is similar: we argue that expectations can set the reference points against which people compare future outcomes, so that assigning value to an event consists in comparing the state of the world following the event to the state of the world just prior to the event.
-As knowledge of the world is uncertain, our understanding of these states is probabilistic--that is, our model for the state of the world is just our expectations about what has happened, what is happening, and what will happen in the future.
+We further assume that people’s knowledge of the world is uncertain, so we represent the states of the world probabilistically.
+This means that a person’s model for the state of the world is just their expectations about what has happened, what is happening, and what will happen in the future.
 One direct implication of this proposal is that the evaluation of an event is linked to what is learned about the world as a result of that event’s occurrence.
@@
 
-This account thereby provides an information-theoretic route to understanding how expectations directly influence evaluation.
+One striking implication of this account is that it provides an information-theoretic route to understanding how expectations directly influence evaluation.
 A fundamental insight of Information Theory is that the information carried by an event is a function of its prior probability; low probability events carry more information than high probability events [@Shannon1948].
@@
 For instance, expectations have long been recognized as fundamental to models of human and animal learning [e.g., @Rescorla1972].
-Similarly, science progresses most when new findings impugn widely-accepted theories or when a theory’s extreme predictions are validated.
 
@@
 
-\begin{equation}
+\begin{equation*}
 E_X[u(X)] = \sum_x u(x)p(x)
-\end{equation}
+\end{equation*}
 
@@
 
-\begin{equation}
+\begin{equation*}
 E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
-\end{equation}
+\end{equation*}
 
-Given our construal of event evaluation as the comparison between states of the world following and prior to the event, we define the value assigned to the signalling event, $V(S)$, as:
+Our construal of event evaluation is as a comparison between states of the world following and prior to an event. Consequently, we define the value assigned to the signalling event, $V(S)$, as:
 
-\begin{equation}
+\begin{equation*}
 V(S) = E_{X|S}[u(X)] - E_X[u(X)]
-\end{equation}
+\end{equation*}
 
@@
 
-\begin{equation}
+\begin{equation*}
 V(s=1)=(1-p(s=1))\sum_x u(x)p(x|s=1) - u(x)p(x|s=0)
-\end{equation}
+\end{equation*}
 
@@
 For this reason, we manipulated the prior probability of events by describing two otherwise identical events occurring in different contexts.
+We explicitly chose contexts that differed in ways that could not be accounted for by construal-level explanations (@Trope2010) or explanations based on biases towards one’s ingroup (@Brewer1979).
 This allowed us to test the role of expectations based on world knowledge, as well as providing a more naturalistic set of items.
@@
 Furthermore, these differences were large, systematic, and robust.
-Although there was variability in the situations that elicited the largest differences, our statistical approach treated trial as a random effect (see Methods and Results), allowing us to generalize conclusions to items we did not test [@Judd2012].
+Although there was variability in the situations that elicited the largest differences, our statistical approach treated trial as a random effect (see Methods and Results), allowing us to generalize our conclusions to items we did not test [@Judd2012].
 
@@
 
-Yet, however well-grounded the underlying principles, there’s a clear moral danger whenever expectations influence our moral concerns.
-For instance, "victim blaming" describes the tendency to shift blame away from aggressors and onto victims of crimes like sexual assault [@Koss2000] and police violence.
-Work by Janoff-Bulman and colleagues [-@Janoff-bulman1985] has shown that merely increasing the perceived probability of a victim suffering a sexual assault induced participants to blame the victim for her assault.
-People’s expectations may explain the flip side of victim blaming, whereby those who victim blame also tend to judge transgressors less harshly: When people perceive a victims’ actions as increasing the probability of their eventual assault, they may perceive an aggressor’s actions as producing a smaller change in utility, thus warranting weaker condemnation.
-
-The influence of expectations on moral evaluations poses perhaps the greatest threat when considering our reactions to humanitarian challenges brought on by poverty or regional instability.
+Yet, however well-grounded the underlying principles, there’s a clear moral danger whenever expectations influence our moral concerns [@Koss2000; -@Janoff-bulman1985].
+These effects pose a significant threat when considering our reactions to humanitarian challenges brought on by poverty or regional instability.
 In these cases, the influence of expectations on moral evaluations threatens to impose vicious and morally pernicious cycles.
@@
 
-We conducted two preregistered studies examining the role of expectations in moral evaluations using two different judgment tasks (https://osf.io/5qnf7/).
+We conducted two preregistered studies examining the role of expectations in moral evaluations (https://osf.io/86rsw/).
 Each study consisted of multiple trials where participants were presented brief descriptions of two events.
@@
 These trials were included to allow participants a chance to use the extremes of the response scale and to reduce any task demands that might drive them to make artificially fine-grained distinctions when generating their responses.
-All experimental materials for these studies are available as Supporting Information at https://osf.io/5qnf7/.
+All experimental materials for these studies are available as Supporting Information at https://osf.io/86rsw/.
 
@@
 We note that the most common response was to view both events, which involve identical harms, as equivalently upsetting.
-However, when participants did view one of the events as more upsetting, they showed a clear bias towards choosing unexpected events as more upsetting.
+However, across almost every item, participants showed a clear bias towards choosing unexpected events as more upsetting.
 Thus, Study 2 replicates the tendency to view unexpected moral events as more upsetting using a more conservative judgment task.
@@
 
-Evaluation (e.g., of utility) is often examined in the context of decision-making.
+Evaluation of utility is often examined in the context of decision-making.
 However, people often generate post hoc evaluations of events for their own sake.
-For instance, to reflect on the results of our choices and to evaluate others--to judge the wisdom of their decisions or the morality of their actions.
+For instance, people evaluate events to judge the wisdom of their decisions or the morality of their actions.
 
@@
 
-\begin{equation}
+\begin{equation*}
 E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
-\end{equation}
+\end{equation*}
 
@@
 
-Applying the expectation of conditional expectation, we can write:
+Our goal is to express this relation in terms of $p(S)$. First, we manipulate $E_X[u(X)]$ by applying the expectation of conditional expectation and some algebra.
 
-\begin{equation}
-V(s=1) = \sum_x u(x)p(x|s=1) - \sum_x \sum_s u(x)p(x|s)p(s)
-\end{equation}
+\begin{align*}
+\sum_x u(x)p(x) & = \sum_s u(x)p(x|s)p(s) \\
+ &= \begin{aligned}[t]
+    & \sum_x u(x)p(x|s=1)p(s=1) - \\
+    & \quad \sum_x u(x)p(x|s=0)p(s=0) \end{aligned}\\
+ &= \begin{aligned}[t]
+    & p(s=1)\sum_x u(x)p(x|s=1) -\\
+    & \quad p(s=0) \sum_x u(x)p(x|s=0) \end{aligned}\\
+ &= \begin{aligned}[t]
+    & p(s=1)\sum_x u(x)p(x|s=1) - \\
+    & \quad (1-p(s=1)) \sum_x u(x)p(x|s=0) \end{aligned}\\
+\end{align*}
 
-We then perform some algebra.
-We break out the summation over $S$ (6), pull the multipliers out of the last two sums (7), rewrite $p(s=0)$ in terms of $p(s=1)$ (8), and factor to combine the first two sums (9).
+Substituting this quantity back into Equation 2, we obtain:
 
-\begin{align}
-V(s=1) &= \sum_x u(x)p(x|s=1) - \sum_x u(x)p(x|s=1)p(s=1) - \sum_x u(x)p(x|s=0)p(s=0) \\
- &= \sum_x u(x)p(x|s=1) - p(s=1)\sum_x u(x)p(x|s=1) - p(s=0) \sum_x u(x)p(x|s=0) \\
- &= \sum_x u(x)p(x|s=1) - p(s=1)\sum_x u(x)p(x|s=1) - (1-p(s=1)) \sum_x u(x)p(x|s=0) \\
- &=(1-p(s=1))\sum_x u(x)p(x|s=1) - (1-p(s=1)) \sum_x u(x)p(x|s=0)
-\end{align}
+\begin{align*}
+V(s=1) = & \sum_x u(x)p(x|s=1) - \\ & p(s=1)\sum_x u(x)p(x|s=1) -\\ & (1-p(s=1)) \sum_x u(x)p(x|s=0)
+\end{align*}
 
-Finally, combining both sums by their common factor yields (10):
+Finally, we factor the first two sums and combine them with the last sum to yield:
 
@@
 
-Thus, it can be seen under this formalization that the role of expectations in the evaluation of events is fundamental--it does not, for instance, depend on any specifics concerning the utility function over states of the world.
+Thus, it can be seen under this formalization that the role of expectations in the evaluation of events is fundamental in the sense that it does not depend on any specifics concerning the utility function over states of the world.
 Instead, this conclusion depends on just two assumptions: first, that events are evaluated by comparing the utilities of the states of the world prior to and following the event; and second, that knowledge of those states of the world is uncertain.
@@
 Though our model of event evaluation is not an information-theoretic model in a direct sense, this interpretation still applies: the information carried by observing $S$ is a function of that observation’s prior probability, which we have in turn shown determines its evaluation.
-Also relevant, although not addressed here, is the mutual information between $X$ and $S$--the extent to which $S$ informs $X$.
 

### 2018-07-03 paper/paper.tex
updating pre-submission

@@
-\documentclass[9pt,twocolumn,twoside,lineno]{pnas-new}
+\documentclass[9pt,twocolumn,twoside,]{pnas-new}
 
@@
 % Please add here a significance statement to explain the relevance of your work
-\significancestatement{People living in geo-politically unstable regions or in the developing
-world are often those who are most affected by terrorism, famine, and
-natural disasters, and thus are the very people in greatest need of
-assistance and concern from the world at-large. However, for these very
-reasons, it is often unsurprising when harm befalls them. Two
-experiments confirm this intuition: we found that people's expectations
-bias their reactions to morally-significant events--people are more
-upset by surprising harms than more expected harms. This bias threatens
-to reduce observer's concern for victims most in need, making people
-less likely to donate time or money to aid victims, or to take political
-action to prevent further harms.}
+\significancestatement{In two behavioral studies, we found that people's expectations biased
+their reactions to morally-significant events--people are more upset by
+surprising harms than more expected harms. This bias appears deeply
+pernicious: because it is often unsurprising when people living in
+poverty, dangerous neighborhoods, or unstable regions become victims of
+crime, natural disasters, or terrorism, observers are likely to
+experience reduced concern for these victims. Yet people living in
+disadvantaged circumstances are those most in need of assistance and
+concern from the world at-large. By blunting people's concern for those
+victims who are most in need, the bias we have uncovered threatens to
+increase their suffering and further sustain inequality.}
 
-\authorcontributions{Please provide details of author contributions here.}
+\authorcontributions{Both authors jointly conceived of the project and theory, designed and
+conducted experiments, and analyzed results. D.P. developed the formal
+aspects of the theory.}
 
-\authordeclaration{Please declare any conflict of interest here.}
+\authordeclaration{The authors declare no conflicts of interest.}
 
@@
 angry, and driven to action. But our responses to these events differ
-depending on the context (1). In particular, our expectations play an
-important role in our reactions to events. This seems true of both the
-mundane--when a friend's loved-one passes, we often ask, ``was it
-sudden?''--the dramatic--a murder in a sleepy town prompts a very
-different reaction than a similar crime in a major city--and even the
-catastrophic: acts of terrorism seem to evoke more outrage, horror, and
-concern when they occurs in peaceful rather than unstable regions. For
-instance, the 2015 Paris attacks were felt around the world. Yet, most
-of those mourning had little to say 15 hours earlier, when another
-tragic attack killed at least 43 people in Beirut (2). Likewise, stories
-of the Virginia Tech massacre, which killed 32 people, wildly
-overshadowed coverage of attacks killing nearly 200 people in Iraq--one
-of the bloodiest days since the 2003 invasion (3). In each of these
-situations, the surprise elicited by an event appears to affect people's
-reactions to the event. In some contexts, harm is tragic and in others,
-``these things happen.''
+depending on the context in which these events occur (1). In particular,
+our expectations play an important role in our reactions to events. This
+seems true of both the mundane--when a friend's loved-one passes, we
+often ask, ``was it sudden?''--the dramatic--a murder in a sleepy town
+prompts a very different reaction than a similar crime in a major
+city--and even the catastrophic: acts of terrorism seem to evoke more
+outrage, horror, and concern when they occurs in peaceful rather than
+unstable regions. For instance, the 2015 Paris attacks were felt around
+the world. Yet, most of those mourning had little to say 15 hours
+earlier, when another tragic attack killed at least 43 people in Beirut
+(2). Likewise, stories of the Virginia Tech massacre, which killed 32
+people, wildly overshadowed coverage of attacks killing nearly 200
+people in Iraq--one of the bloodiest days since the 2003 invasion (3).
+In each of these situations, the surprise elicited by an event appears
+to affect people's reactions to the event. In some contexts, harm is
+tragic and in others, ``these things happen.''
 
@@
 psychological frameworks explicate the role of expectations in moral
-evaluations. Outside of the domain of morality, however, everyday
-experiences and controlled experiments show that people's evaluations of
-events unambiguously depend on their expectations about those events.
-There is a special thrill to having one's expectations exceeded and
+evaluations. Outside of the domain of morality, everyday experiences and
+controlled experiments show that people's evaluations of events
+unambiguously depend on their expectations about those events. There is
+a special thrill to having one's expectations exceeded and
 disappointment when events fail to meet expectations: we root for the
@@
 
-Applied to moral issues, the influences of expectations on evaluation
-would be a dangerous bias. Tragedies are tragic, whether expected or
+The influence of expectations on moral evaluations threatens to produce
+negative social consequences. Tragedies are tragic, whether expected or
 not, and failure to recognize that could lead to tacit endorsement of
@@
 following the event to the state of the world just prior to the event.
-As knowledge of the world is uncertain, our understanding of these
-states is probabilistic--that is, our model for the state of the world
-is just our expectations about what has happened, what is happening, and
-what will happen in the future. One direct implication of this proposal
-is that the evaluation of an event is linked to what is learned about
-the world as a result of that event's occurrence. That is, we evaluate
-events positively or negatively to the extent that they inform our
-understanding of the world in positive or negative ways.
+We further assume that people's knowledge of the world is uncertain, so
+we represent the states of the world probabilistically. This means that
+a person's model for the state of the world is just their expectations
+about what has happened, what is happening, and what will happen in the
+future. One direct implication of this proposal is that the evaluation
+of an event is linked to what is learned about the world as a result of
+that event's occurrence. That is, we evaluate events positively or
+negatively to the extent that they inform our understanding of the world
+in positive or negative ways.
 
-This account thereby provides an information-theoretic route to
-understanding how expectations directly influence evaluation. A
-fundamental insight of Information Theory is that the information
-carried by an event is a function of its prior probability; low
-probability events carry more information than high probability events
-(11). We propose that people learn more about the state of the world
-when their expectations are violated by shocking world events as
-compared to when they are affirmed by less surprising events. For
-example, if a bombing occurs in Paris--a less expected location--rather
-than Lebanon--a more expected location--we learn that the world is more
-dangerous than we had previously believed. This suggestion is novel in
-the moral domain, but these points are widely accepted elsewhere. For
-instance, expectations have long been recognized as fundamental to
-models of human and animal learning (12). Similarly, science progresses
-most when new findings impugn widely-accepted theories or when a
-theory's extreme predictions are validated.
+One striking implication of this account is that it provides an
+information-theoretic route to understanding how expectations directly
+influence evaluation. A fundamental insight of Information Theory is
+that the information carried by an event is a function of its prior
+probability; low probability events carry more information than high
+probability events (11). We propose that people learn more about the
+state of the world when their expectations are violated by shocking
+world events as compared to when they are affirmed by less surprising
+events. For example, if a bombing occurs in Paris--a less expected
+location--rather than Lebanon--a more expected location--we learn that
+the world is more dangerous than we had previously believed. This
+suggestion is novel in the moral domain, but these points are widely
+accepted elsewhere. For instance, expectations have long been recognized
+as fundamental to models of human and animal learning (12).
 
@@
 
-\begin{equation}
+\begin{equation*}
 E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
-\end{equation}
+\end{equation*}
 
-Given our construal of event evaluation as the comparison between states
-of the world following and prior to the event, we define the value
-assigned to the signalling event, \(V(S)\), as:
+Our construal of event evaluation is as a comparison between states of
+the world following and prior to an event. Consequently, we define the
+value assigned to the signalling event, \(V(S)\), as:
 
-\begin{equation}
+\begin{equation*}
 V(S) = E_{X|S}[u(X)] - E_X[u(X)]
-\end{equation}
+\end{equation*}
 
@@
 manipulated the prior probability of events by describing two otherwise
-identical events occurring in different contexts. This allowed us to
-test the role of expectations based on world knowledge, as well as
-providing a more naturalistic set of items.
+identical events occurring in different contexts. We explicitly chose
+contexts that differed in ways that could not be accounted for by
+construal-level explanations ((1)) or explanations based on biases
+towards one's ingroup (\textbf{???}). This allowed us to test the role
+of expectations based on world knowledge, as well as providing a more
+naturalistic set of items.
 
@@
 the largest differences, our statistical approach treated trial as a
-random effect (see Methods and Results), allowing us to generalize
+random effect (see Methods and Results), allowing us to generalize our
 conclusions to items we did not test (14).
@@
 Yet, however well-grounded the underlying principles, there's a clear
-moral danger whenever expectations influence our moral concerns. For
-instance, ``victim blaming'' describes the tendency to shift blame away
-from aggressors and onto victims of crimes like sexual assault (15) and
-police violence. Work by Janoff-Bulman and colleagues (16) has shown
-that merely increasing the perceived probability of a victim suffering a
-sexual assault induced participants to blame the victim for her assault.
-People's expectations may explain the flip side of victim blaming,
-whereby those who victim blame also tend to judge transgressors less
-harshly: When people perceive a victims' actions as increasing the
-probability of their eventual assault, they may perceive an aggressor's
-actions as producing a smaller change in utility, thus warranting weaker
-condemnation.
+moral danger whenever expectations influence our moral concerns (15,
+16). These effects pose a significant threat when considering our
+reactions to humanitarian challenges brought on by poverty or regional
+instability. In these cases, the influence of expectations on moral
+evaluations threatens to impose vicious and morally pernicious cycles.
+For instance, people living in geo-politically unstable regions or in
+the developing world are often those who are most affected by terrorism,
+famine, and natural disasters, and are the very people in greatest need
+of assistance and concern from the world at-large. However, for these
+very reasons, it is often unsurprising when harm befalls people living
+in these circumstances. Thus the influence of expectations on moral
+evaluations threatens to reduce observer's concern for these victims
+most in need, making people less likely to donate time or money to aid
+victims, or to take political action to prevent further harms. The
+present research identifies this potentially pernicious moral bias and
+grounds it in information theoretic principles. Hopefully, by further
+understanding these cognitive dynamics, psychologists and policymakers
+can begin to find ways to reduce or counteract people's tendency to
+ignore the plight of those in most need of attention.
 
-The influence of expectations on moral evaluations poses perhaps the
-greatest threat when considering our reactions to humanitarian
-challenges brought on by poverty or regional instability. In these
-cases, the influence of expectations on moral evaluations threatens to
-impose vicious and morally pernicious cycles. For instance, people
-living in geo-politically unstable regions or in the developing world
-are often those who are most affected by terrorism, famine, and natural
-disasters, and are the very people in greatest need of assistance and
-concern from the world at-large. However, for these very reasons, it is
-often unsurprising when harm befalls people living in these
-circumstances. Thus the influence of expectations on moral evaluations
-threatens to reduce observer's concern for these victims most in need,
-making people less likely to donate time or money to aid victims, or to
-take political action to prevent further harms. The present research
-identifies this potentially pernicious moral bias and grounds it in
-information theoretic principles. Hopefully, by further understanding
-these cognitive dynamics, psychologists and policymakers can begin to
-find ways to reduce or counteract people's tendency to ignore the plight
-of those in most need of attention.
-
-\section*{Materials and Methods}\label{materials-and-methods}
-\addcontentsline{toc}{section}{Materials and Methods}
+\matmethods{
 
 We conducted two preregistered studies examining the role of
-expectations in moral evaluations using two different judgment tasks
-(\url{https://osf.io/5qnf7/}). Each study consisted of multiple trials
-where participants were presented brief descriptions of two events.
-Their task was to decide which of the two events was more upsetting. In
-Study 1, they made a two-alternative forced-choice, and in Study 2 they
-made their judgments using a five-point rating scale. Both studies
-consisted of 16 experimental trials, five ``equivalent'' filler trials,
-and five ``non-equivalent'' filler trials.
+expectations in moral evaluations (\url{https://osf.io/5qnf7/}). Each
+study consisted of multiple trials where participants were presented
+brief descriptions of two events. Their task was to decide which of the
+two events was more upsetting. In Study 1, they made a two-alternative
+forced-choice, and in Study 2 they made their judgments using a
+five-point rating scale. Both studies consisted of 16 experimental
+trials, five ``equivalent'' filler trials, and five ``non-equivalent''
+filler trials.
 
@@
 We note that the most common response was to view both events, which
-involve identical harms, as equivalently upsetting. However, when
-participants did view one of the events as more upsetting, they showed a
-clear bias towards choosing unexpected events as more upsetting. Thus,
-Study 2 replicates the tendency to view unexpected moral events as more
-upsetting using a more conservative judgment task.
+involve identical harms, as equivalently upsetting. However, across
+almost every item, participants showed a clear bias towards choosing
+unexpected events as more upsetting. Thus, Study 2 replicates the
+tendency to view unexpected moral events as more upsetting using a more
+conservative judgment task.
 
@@
 
-Evaluation (e.g., of utility) is often examined in the context of
+Evaluation of utility is often examined in the context of
 decision-making. However, people often generate post hoc evaluations of
-events for their own sake. For instance, to reflect on the results of
-our choices and to evaluate others--to judge the wisdom of their
-decisions or the morality of their actions.
+events for their own sake. For instance, people evaluate events to judge
+the wisdom of their decisions or the morality of their actions.
 
@@
 
-Applying the expectation of conditional expectation, we can write:
+Our goal is to express this relation in terms of \(p(S)\). First, we
+manipulate \(E_X[u(X)]\) by applying the expectation of conditional
+expectation and some algebra.
 
-\begin{equation}
-V(s=1) = \sum_x u(x)p(x|s=1) - \sum_x \sum_s u(x)p(x|s)p(s)
-\end{equation}
+\begin{align*}
+\sum_x u(x)p(x) & = \sum_s u(x)p(x|s)p(s) \\
+ &= \begin{aligned}[t]
+    & \sum_x u(x)p(x|s=1)p(s=1) - \\
+    & \quad \sum_x u(x)p(x|s=0)p(s=0) \end{aligned}\\
+ &= \begin{aligned}[t]
+    & p(s=1)\sum_x u(x)p(x|s=1) -\\
+    & \quad p(s=0) \sum_x u(x)p(x|s=0) \end{aligned}\\
+ &= \begin{aligned}[t]
+    & p(s=1)\sum_x u(x)p(x|s=1) - \\
+    & \quad (1-p(s=1)) \sum_x u(x)p(x|s=0) \end{aligned}\\
+\end{align*}
 
-We then perform some algebra. We break out the summation over \(S\) (6),
-pull the multipliers out of the last two sums (7), rewrite \(p(s=0)\) in
-terms of \(p(s=1)\) (8), and factor to combine the first two sums (9).
+Substituting this quantity back into Equation 2, we obtain:
 
-\begin{align}
-V(s=1) &= \sum_x u(x)p(x|s=1) - \sum_x u(x)p(x|s=1)p(s=1) - \sum_x u(x)p(x|s=0)p(s=0) \\
- &= \sum_x u(x)p(x|s=1) - p(s=1)\sum_x u(x)p(x|s=1) - p(s=0) \sum_x u(x)p(x|s=0) \\
- &= \sum_x u(x)p(x|s=1) - p(s=1)\sum_x u(x)p(x|s=1) - (1-p(s=1)) \sum_x u(x)p(x|s=0) \\
- &=(1-p(s=1))\sum_x u(x)p(x|s=1) - (1-p(s=1)) \sum_x u(x)p(x|s=0)
-\end{align}
+\begin{align*}
+V(s=1) = & \sum_x u(x)p(x|s=1) - \\ & p(s=1)\sum_x u(x)p(x|s=1) -\\ & (1-p(s=1)) \sum_x u(x)p(x|s=0)
+\end{align*}
 
-Finally, combining both sums by their common factor yields (10):
+Finally, we factor the first two sums and combine them with the last sum
+to yield:
 
@@
 Thus, it can be seen under this formalization that the role of
-expectations in the evaluation of events is fundamental--it does not,
-for instance, depend on any specifics concerning the utility function
+expectations in the evaluation of events is fundamental in the sense
+that it does not depend on any specifics concerning the utility function
 over states of the world. Instead, this conclusion depends on just two
@@
 observation's prior probability, which we have in turn shown determines
-its evaluation. Also relevant, although not addressed here, is the
-mutual information between \(X\) and \(S\)--the extent to which \(S\)
-informs \(X\).
+its evaluation.
+}
 

### 2018-07-18 SuppInfo.Rmd
submitted to PNAS

@@
 output:
-  pdf_document: default
-  html_notebook: default
+  pdf_document:
+    keep_tex: true
+  # html_notebook: default
 ---
@@
 ```
+
 # Item Norming
@@
 We created and normed 20 items, each consisting of a pair of events that differed in their prior likelihood.
-The two events (outcomes) were manipulated so what one would see more expected than the other.
+The two events (outcomes) were manipulated so what one would seem more expected than the other.
 Participants (_N_ = `r nrow(df_wide)`) were asked to choose which event seemed more likely using a 5-point scale (1 = Event 1 seems a lot more likely, 2 = Event 1 seems a little more likely, 3 = Neither event seems more likely, 4 = Event 2 seems a little more likely, 5 = Event 2 seems much more likely).

### 2018-07-18 paper/paper.Rmd
submitted to PNAS

@@
  When we learn of tragic events--a violent killing or a terrorist attack--we are mournful and often driven to action.
-But our responses to these events can differ substantially depending on the context of the event.
+But responses to these events can differ substantially depending on the context in which they occur.
 When tragedy takes us by surprise, we are shocked and outraged, but when it strikes in unstable regions or rough neighborhoods, we are often resigned to feel that “these things happen.” Here we explain the role that expectations seem to play in moral evaluation by modeling evaluations as the comparison of the state of the world before and after an event.
-According to this model, expectations influence evaluations through information-theoretic principles: less expected events do more to inform us about the state of the world than do more expected events.
-Consistent with this theory, in two preregistered studies we found that people’s judgments of morally significant events were affected by the prior probability of that event, so that people were more upset about an event that was unexpected (e.g., a robbery at a clothing store) than an event that was more more probable (e.g., a robbery at a convenience store).
+According to this theory, expectations influence evaluations through information-theoretic principles: less expected events do more to inform us about the state of the world than do more expected events.
+Consistent with this theory, in two preregistered studies we found that people’s judgments of morally-significant events were affected by the prior probability of that event.
+People were more upset about an event that was unexpected (e.g., a robbery at a clothing store) than an event that was more more probable (e.g., a robbery at a convenience store).
 This bias threatens to impose a pernicious cycle: People living in poverty, geo-politically unstable regions, and other dangerous circumstances have the greatest need for assistance and concern.
 However, for these very reasons, it is often unsurprising when they are harmed.
-Consequently, the influence of expectations on moral evaluations threatens to reduce observer's concern for those people who are most in need.
+Consequently, the influence of expectations on moral evaluations threatens to reduce observers’ concern for those most in need.
 
 significance: |
-  In two behavioral studies, we found that people's expectations biased their reactions to morally-significant events--people are more upset by surprising harms than more expected harms.
-This bias appears deeply pernicious: because it is often unsurprising when people living in poverty, dangerous neighborhoods, or unstable regions become victims of crime, natural disasters, or terrorism, observers are likely to experience reduced concern for these victims.
-Yet people living in disadvantaged circumstances are those most in need of assistance and concern from the world at-large.
-By blunting people's concern for those victims who are most in need, the bias we have uncovered threatens to increase their suffering and further sustain inequality.
+  In two behavioral studies, we found that people's expectations biased their reactions to morally-significant events--people were more upset by surprising harms than more expected harms.
+This bias threatens to impose a pernicious cycle: People living in poverty, unstable regions, and other dangerous circumstances have the greatest need for assistance and concern from the world at-large.
+Yet for this very reason, it is often unsurprising when they are harmed.
+By blunting observers' concern for the disadvantaged, this bias could make people less likely to donate money, to volunteer, or to take political action, thereby increasing inequality and compounding suffering.
+Recognizing and understanding this bias will suggest ways to mitigate its harmful consequences.
 
@@
     x = "Trial",
-    y = "Probability of judging less\nexpected more upsetting"
+    y = "Prob. less expected\njudged more upsetting"
   ) 
@@
     x = "Trial",
-    y = '"Upsetting" Judgments'
+    y = 'Upset Judgments'
   )
@@
 When we learn of tragic events--a friend's loss of a loved-one, a violent killing, or a terrorist attack--we are mournful, concerned, angry, and driven to action.
-But our responses to these events differ depending on the context in which these events occur [e.g., @Trope2010].
+But our responses to these events differ depending on the context in which they occur [e.g., @Trope2010].
 In particular, our expectations play an important role in our reactions to events.
-This seems true of both the mundane--when a friend's loved-one passes, we often ask, "was it sudden?"--the dramatic--a murder in a sleepy town prompts a very different reaction than a similar crime in a major city--and even the catastrophic: acts of terrorism seem to evoke more outrage, horror, and concern when they occurs in peaceful rather than unstable regions.
+This seems true of the ordinary: when a friend's loved-one passes, we often ask, "was it sudden?"; the dramatic: a murder in a sleepy town prompts a very different reaction than a similar crime in a major city; and even the catastrophic: acts of terrorism seem to evoke more outrage, horror, and concern when they occur in peaceful rather than unstable regions.
 For instance, the 2015 Paris attacks were felt around the world.
@@
 Likewise, stories of the Virginia Tech massacre, which killed 32 people, wildly overshadowed coverage of attacks killing nearly 200 people in Iraq--one of the bloodiest days since the 2003 invasion [@Greenslade2007].
-In each of these situations, the surprise elicited by an event appears to affect people’s reactions to the event.
-In some contexts, harm is tragic and in others, “these things happen.”
+In each of these situations, the surprise elicited by an event appears to have affected people’s reactions to it.
+In some contexts, harm is tragic.
+In others, “these things happen.”
 
-Although this pattern seems so intuitive as to be obvious, no reasonable code of ethics would justify this pattern of judgment, nor do any moral psychological frameworks explicate the role of expectations in moral evaluations.
+Although this pattern seems so intuitive as to be obvious, no reasonable code of ethics would justify these judgments, nor do any moral psychological frameworks explicate the role of expectations in moral evaluations.
 Outside of the domain of morality, everyday experiences and controlled experiments show that people’s evaluations of events unambiguously depend on their expectations about those events.
 There is a special thrill to having one’s expectations exceeded and disappointment when events fail to meet expectations: we root for the underdog, hold surprise parties, and foreshadow bad news to ease its delivery [@Bell1985].
-Indeed, in a controlled study, Mellers and colleagues [-@Mellers1997] found that expectations influenced affective reactions during a gambling task: Given a gamble with a 10% chance to win \$30 and 90% chance to win \$0, participants felt little disappointment with the \$0 outcome, but were considerably more excited when they won \$30.
-Conversely, given a gamble with a 90% chance of winning \$60 and 10% chance of winning \$0, participants were disappointed with the \$0 outcome and showed more muted enjoyment of the \$60 outcome.
-In fact, in gambles similar to these, people were happier with the smaller unexpected gain than with the larger but more expected gain [also see @Shepperd2002].
+Indeed, in a controlled study, Mellers and colleagues [-@Mellers1997] found that expectations influenced affective reactions during a gambling task: Participants were often more excited by an small unexpected gain (e.g., winning \$30 with 10% probability) than they were by a larger but more expected gain (e.g., winning \$60 with 90% probability) [also see @Shepperd2002].
 
-The influence of expectations on moral evaluations threatens to produce negative social consequences.
-Tragedies are tragic, whether expected or not, and failure to recognize that could lead to tacit endorsement of the pain and suffering of others wherever pain and suffering are the status quo.
-Do expectations directly bias reactions to morally consequential events, as intuition and anecdote seem to suggest?
+It seems intuitive that surprise would shape people’s reactions to parties, sporting events, and games of chance.
+However, it seems wholly inappropriate that surprise would shape people’s reactions to events with serious moral implications.
+Tragedies are tragic, whether expected or not, and failure to recognize this could lead to tacit endorsement of the pain and suffering of others wherever pain and suffering are the status quo.
+If expectations truly do influence our reactions to morally-significant world events, as intuition and anecdote seem to suggest, there are potentially serious social consequences.
 To demonstrate the plausibility of this suggestion, we develop a theoretical account of how expectations influence evaluations, not as a strange quirk or uniquely human bias, but instead as a fundamental component of how people evaluate events.
 
-Most events of consequence consist of changes--a transition from one state to another.
+Events, as we will conceive of them, are changes in the state of the world--transitions from one state to another.
 For this reason, we construe event evaluation as a comparison between the states of the world before and after the event.
-Comparison plays a crucial role in many theories of decision-making, and especially in theories aimed at explaining the role of expectations in evaluation  [e.g., @Bell1985; @Gul1991; @Loomes1986; @Mellers1997].
+Similar comparison processes play crucial roles in many theories of decision-making [e.g., @Bell1985; @Gul1991; @Loomes1986; @Mellers1997].
 For instance, in Prospect Theory [@Kahneman1979; @Tversky1992], the utility of a prospect is evaluated by comparison to a current reference point.
 Our proposal is similar: we argue that expectations can set the reference points against which people compare future outcomes, so that assigning value to an event consists in comparing the state of the world following the event to the state of the world just prior to the event.
-We further assume that people’s knowledge of the world is uncertain, so we represent the states of the world probabilistically.
-This means that a person’s model for the state of the world is just their expectations about what has happened, what is happening, and what will happen in the future.
-One direct implication of this proposal is that the evaluation of an event is linked to what is learned about the world as a result of that event’s occurrence.
+Further, we assume that people’s knowledge of the world is uncertain, so we represent the states of the world probabilistically.
+This means that a person’s model for the state of the world is simply their expectations about what has happened, what is happening, and what will happen in the future.
+A direct implication of this proposal is that the evaluation of an event is linked to what is learned about the world as a result of that event’s occurrence.
 That is, we evaluate events positively or negatively to the extent that they inform our understanding of the world in positive or negative ways.
@@
 We propose that people learn more about the state of the world when their expectations are violated by shocking world events as compared to when they are affirmed by less surprising events.
-For example, if a bombing occurs in Paris--a less expected location--rather than Lebanon--a more expected location--we learn that the world is more dangerous than we had previously believed.
-This suggestion is novel in the moral domain, but these points are widely accepted elsewhere.
+For example, compared to a bombing in Lebanon, if a bombing occurs in Paris, a less expected location, we learn that the world is more dangerous than we had previously believed.
+Although this suggestion is novel in the moral domain, these points are widely accepted elsewhere.
 For instance, expectations have long been recognized as fundamental to models of human and animal learning [e.g., @Rescorla1972].
@@
 From this it can be shown that the value assigned to event $S$ is proportional to its prior probability (see Theoretical Formalization).
-Specifically, it is the difference in the expected utility under different values of $S$, weighted by the inverse of the prior probability of $S$..
+Specifically, it is the difference in the expected utility under different values of $S$, weighted by the inverse of the prior probability of $S$.
 [^1]
@@
 This conclusion does not depend on the nature of the event in question nor on the utility function over events, but instead holds generally wherever we conceive of evaluation as a holistic comparison between uncertain states of the world.
-Consequently, this account predicts expectations will equally affect reactions to gambles with clearly specified values and probabilities [@Mellers1997] and real-world events with ill-defined values and where reasoners' expectations are defined by their existing beliefs and world knowledge.
+Consequently, this account predicts expectations will equally affect reactions to gambles with clearly specified values and probabilities [@Mellers1997], as well as real-world events with ill-defined values, where reasoners' expectations are defined by their existing beliefs and world knowledge.
 
-Given the potentially fundamental role expectations play in event evaluation, and the potentially serious moral biases that may result, we sought to test empirically whether people’s evaluations of morally harmful events are affected by their expectations about those events.
+Given the potentially fundamental role that expectations play in event evaluation, and the potentially serious moral biases that may result, we sought to test empirically whether people’s evaluations of morally harmful events are affected by their expectations about those events.
 We conducted two preregistered studies examining the role of expectations in moral evaluations.
-In both studies, participants were presented with a series of trials where they read brief descriptions of two different events and were asked to indicate which of the two events seemed more upsetting.
+In both studies, participants were presented with a series of trials in which they read brief descriptions of two different events and were asked to indicate which of the two events seemed more upsetting.
 In these trials, the two events were highly similar, but were manipulated so that they differed in their perceived prior probabilities: one event was more expected and one more unexpected.
@@
 Asking participants to compare two similar events allowed us to test this directional prediction.
-Second, it is the informativeness of events, determined by one's prior expectations of the event, that we predict will shape evaluations.
-Thus, reasoners should not have to be induced to hold expectations experimentally [as in experiments by @Mellers1997], but instead their own prior beliefs will be sufficient to shape their expectations and affect their evaluations.
+Second, it is the informativeness of events, determined by one's prior expectations of the event, that we predicted would shape evaluations.
+Thus, reasoners should not have to be induced to hold expectations experimentally [as in experiments by @Mellers1997], but instead their own prior beliefs should be sufficient to shape their expectations and affect their evaluations.
 For this reason, we manipulated the prior probability of events by describing two otherwise identical events occurring in different contexts.
-We explicitly chose contexts that differed in ways that could not be accounted for by construal-level explanations (@Trope2010) or explanations based on biases towards one’s ingroup (@Brewer1979).
-This allowed us to test the role of expectations based on world knowledge, as well as providing a more naturalistic set of items.
+We explicitly chose contexts that differed in ways that could not be accounted for by construal-level explanations [@Trope2010] or explanations based on biases towards one’s ingroup [@Brewer1999].
+This both allowed us to test the role of expectations based on world knowledge and provided a more naturalistic set of items.
 
@@
 
-Consistent with our predictions, manipulating expectations consistently shaped evaluations across a range of morally-significant events.
+Consistent with our predictions, manipulating expectations shaped evaluations across a range of morally-significant events.
 We found that people viewed unexpected harmful events as more upsetting than expected harmful events, even when the harm a victim suffered was the same.
@@
 Note the scale has been constrained to show relevant region of results.
-In both plots, response of 3 indicates indifference, a response below 3 indicates a bias to choose the more expected event as more upsetting, and a response above 3 indicates a bias to choose the less expected event as more upsetting.
-Similar to Study 1, when participants judged that one of the events was more upsetting, they showed a robust bias toward choosing the less expected event as more upsetting.
+In both plots, a response of 3 indicates indifference, responses below 3 indicate a bias to choose the more expected event as more upsetting, and responses above 3 indicate a bias to choose the less expected event as more upsetting.
+Similar to Study 1, participants showed a robust bias toward choosing the less expected event as more upsetting.
 Trials are sorted by response to aid readability.
@@
 
-There's a sense in which it is clearly adaptive that the violation of our expectations would direct our attention to the events and situations about which our understanding is most in need of revision.
+In some contexts, it is clearly adaptive that the violation of our expectations would direct our attention to the events and situations about which our understanding is most in need of revision [@Rescorla1972].
 For instance, an act of terrorism in a peaceful nation might be less expected because of our beliefs about the effectiveness of security forces, the stability of national relationships, and positive relationships among people groups in that nation.
@@
 
-Yet, however well-grounded the underlying principles, there’s a clear moral danger whenever expectations influence our moral concerns [@Koss2000; -@Janoff-bulman1985].
-These effects pose a significant threat when considering our reactions to humanitarian challenges brought on by poverty or regional instability.
+Yet, however well-grounded the underlying principles, there is a clear moral danger whenever expectations influence our moral concerns.
+In particular, these effects pose a significant threat when considering our reactions to humanitarian challenges brought on by poverty or regional instability.
 In these cases, the influence of expectations on moral evaluations threatens to impose vicious and morally pernicious cycles.
 For instance, people living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and are the very people in greatest need of assistance and concern from the world at-large.
-However, for these very reasons, it is often unsurprising when harm befalls people living in these circumstances.
+However, for precisely these reasons, it is often unsurprising when harm befalls people living in these circumstances.
 Thus the influence of expectations on moral evaluations threatens to reduce observer's concern for these victims most in need, making people less likely to donate time or money to aid victims, or to take political action to prevent further harms.
@@
 We tested our preregistered predictions by computing Bayes Factors (i.e.
-BF01) using the hypothesis function in the R package brms.
+$BF_{01}$) using the hypothesis function in the R package brms.
 Bayes Factors express the ratio of the probability of data under the null hypothesis to the probability of the data under an alternative hypothesis.
 Larger Bayes Factors indicate that the data are more likely under the null hypothesis (i.e., that the intercept is not different from zero) than the alternative hypothesis (i.e., that the intercept is different from zero), and vice versa.
-As Bayes Factors can be influenced by prior choices (Gelman, Simpson, & Betancourt, 2017), we performed prior robustness checks to confirm that our conclusions would not depend substantially on the specification of the priors.
+As Bayes Factors can be influenced by prior choices [@Gelman2017], we performed prior robustness checks to confirm that our conclusions would not depend substantially on the specification of the priors.
 
@@
 Thus, by examining the population-level intercept, we can test whether participants were biased toward selecting the unexpected event ($\beta$ > 0), the expected event ($\beta$ < 0), or were unbiased ($\beta$ = 0).
-We found that people were considerably more likely to think that unexpected events were more upsetting than events that were expected, Intercept = `r round(coefs_exp1$estimate, 3)`, 95% CI [`r round(coefs_exp1$lower, 3)`, `r round(coefs_exp1$upper, 3)`], BF01 `r print_bf(bf_exp1)`.
+We found that people were considerably more likely to think that unexpected events were more upsetting than events that were expected, Intercept = `r round(coefs_exp1$estimate, 3)`, 95% CI [`r round(coefs_exp1$lower, 3)`, `r round(coefs_exp1$upper, 3)`], $BF_{01}$ `r print_bf(bf_exp1)`.
 Figure 1 shows participants' responses broken-down by individual items. Participants' bias toward selecting the unexpected event as more upsetting was consistent across the 16 experimental items.
@@
 To examine the predicted effect, we compared the third intercept coefficient (representing the log-odds of choosing 4 or higher) to the inverse of the second intercept coefficient (thereby representing the log-odds *not* choosing a 3 or higher--i.e., choosing a 1 or 2), allowing us to test whether participants were more likely to choose the expected or unexpected event as being more upsetting in cases where they did not choose the neither option.
-This analysis indicated that people were more likely to think that events that were unexpected were more upsetting than events that were expected, BF01 `r print_bf(bf_exp2)` (see supporting information for full model results).
+This analysis indicated that people were more likely to think that events that were unexpected were more upsetting than events that were expected, $BF_{01}$ `r print_bf(bf_exp2)` (see supporting information for full model results).
 We observed this difference consistently across nearly every item (Figure 2).
@@
 # Theoretical Formalization {#theory .unnumbered}
-
-Evaluation of utility is often examined in the context of decision-making.
-However, people often generate post hoc evaluations of events for their own sake.
-For instance, people evaluate events to judge the wisdom of their decisions or the morality of their actions.
 
@@
 
-Thus, it can be seen under this formalization that the role of expectations in the evaluation of events is fundamental in the sense that it does not depend on any specifics concerning the utility function over states of the world.
+Under this formalization, the role of expectations in the evaluation of events does not depend on any further assumptions about the states of the world or the utility function to be applied over them.
 Instead, this conclusion depends on just two assumptions: first, that events are evaluated by comparing the utilities of the states of the world prior to and following the event; and second, that knowledge of those states of the world is uncertain.

### 2018-07-18 paper/paper.tex
submitted to PNAS

@@
 \significancestatement{In two behavioral studies, we found that people's expectations biased
-their reactions to morally-significant events--people are more upset by
-surprising harms than more expected harms. This bias appears deeply
-pernicious: because it is often unsurprising when people living in
-poverty, dangerous neighborhoods, or unstable regions become victims of
-crime, natural disasters, or terrorism, observers are likely to
-experience reduced concern for these victims. Yet people living in
-disadvantaged circumstances are those most in need of assistance and
-concern from the world at-large. By blunting people's concern for those
-victims who are most in need, the bias we have uncovered threatens to
-increase their suffering and further sustain inequality.}
+their reactions to morally-significant events--people were more upset by
+surprising harms than more expected harms. This bias threatens to impose
+a pernicious cycle: People living in poverty, unstable regions, and
+other dangerous circumstances have the greatest need for assistance and
+concern from the world at-large. Yet for this very reason, it is often
+unsurprising when they are harmed. By blunting observers' concern for
+the disadvantaged, this bias could make people less likely to donate
+money, to volunteer, or to take political action, thereby increasing
+inequality and compounding suffering. Recognizing and understanding this
+bias will suggest ways to mitigate its harmful consequences.}
 
@@
 When we learn of tragic events--a violent killing or a terrorist
-attack--we are mournful and often driven to action. But our responses to
-these events can differ substantially depending on the context of the
-event. When tragedy takes us by surprise, we are shocked and outraged,
-but when it strikes in unstable regions or rough neighborhoods, we are
-often resigned to feel that ``these things happen.'' Here we explain the
-role that expectations seem to play in moral evaluation by modeling
-evaluations as the comparison of the state of the world before and after
-an event. According to this model, expectations influence evaluations
-through information-theoretic principles: less expected events do more
-to inform us about the state of the world than do more expected events.
-Consistent with this theory, in two preregistered studies we found that
-people's judgments of morally significant events were affected by the
-prior probability of that event, so that people were more upset about an
-event that was unexpected (e.g., a robbery at a clothing store) than an
-event that was more more probable (e.g., a robbery at a convenience
-store). This bias threatens to impose a pernicious cycle: People living
-in poverty, geo-politically unstable regions, and other dangerous
-circumstances have the greatest need for assistance and concern.
-However, for these very reasons, it is often unsurprising when they are
-harmed. Consequently, the influence of expectations on moral evaluations
-threatens to reduce observer's concern for those people who are most in
-need.
+attack--we are mournful and often driven to action. But responses to
+these events can differ substantially depending on the context in which
+they occur. When tragedy takes us by surprise, we are shocked and
+outraged, but when it strikes in unstable regions or rough
+neighborhoods, we are often resigned to feel that ``these things
+happen.'' Here we explain the role that expectations seem to play in
+moral evaluation by modeling evaluations as the comparison of the state
+of the world before and after an event. According to this theory,
+expectations influence evaluations through information-theoretic
+principles: less expected events do more to inform us about the state of
+the world than do more expected events. Consistent with this theory, in
+two preregistered studies we found that people's judgments of
+morally-significant events were affected by the prior probability of
+that event. People were more upset about an event that was unexpected
+(e.g., a robbery at a clothing store) than an event that was more more
+probable (e.g., a robbery at a convenience store). This bias threatens
+to impose a pernicious cycle: People living in poverty, geo-politically
+unstable regions, and other dangerous circumstances have the greatest
+need for assistance and concern. However, for these very reasons, it is
+often unsurprising when they are harmed. Consequently, the influence of
+expectations on moral evaluations threatens to reduce observers' concern
+for those most in need.
 \end{abstract}
@@
 
-When we learn of tragic events--a friend's loss of a loved-one, a
-violent killing, or a terrorist attack--we are mournful, concerned,
+When we learn of tragic events---a friend's loss of a loved-one, a
+violent killing, or a terrorist attack---we are mournful, concerned,
 angry, and driven to action. But our responses to these events differ
-depending on the context in which these events occur (1). In particular,
-our expectations play an important role in our reactions to events. This
-seems true of both the mundane--when a friend's loved-one passes, we
-often ask, ``was it sudden?''--the dramatic--a murder in a sleepy town
-prompts a very different reaction than a similar crime in a major
-city--and even the catastrophic: acts of terrorism seem to evoke more
-outrage, horror, and concern when they occurs in peaceful rather than
-unstable regions. For instance, the 2015 Paris attacks were felt around
-the world. Yet, most of those mourning had little to say 15 hours
-earlier, when another tragic attack killed at least 43 people in Beirut
-(2). Likewise, stories of the Virginia Tech massacre, which killed 32
-people, wildly overshadowed coverage of attacks killing nearly 200
-people in Iraq--one of the bloodiest days since the 2003 invasion (3).
-In each of these situations, the surprise elicited by an event appears
-to affect people's reactions to the event. In some contexts, harm is
-tragic and in others, ``these things happen.''
+depending on the context in which they occur (1). In particular, our
+expectations play an important role in our reactions to events. This
+seems true of the ordinary: when a friend's loved-one passes, we often
+ask, ``was it sudden?''; the dramatic: a murder in a sleepy town prompts
+a very different reaction than a similar crime in a major city; and even
+the catastrophic: acts of terrorism seem to evoke more outrage, horror,
+and concern when they occur in peaceful rather than unstable regions.
+For instance, the 2015 Paris attacks were felt around the world. Yet,
+most of those mourning had little to say 15 hours earlier, when another
+tragic attack killed at least 43 people in Beirut (2). Likewise, stories
+of the Virginia Tech massacre, which killed 32 people, wildly
+overshadowed coverage of attacks killing nearly 200 people in Iraq---one
+of the bloodiest days since the 2003 invasion (3). In each of these
+situations, the surprise elicited by an event appears to have affected
+people's reactions to it. In some contexts, harm is tragic. In others,
+``these things happen.''
 
 Although this pattern seems so intuitive as to be obvious, no reasonable
-code of ethics would justify this pattern of judgment, nor do any moral
+code of ethics would justify these judgments, nor do any moral
 psychological frameworks explicate the role of expectations in moral
@@
 found that expectations influenced affective reactions during a gambling
-task: Given a gamble with a 10\% chance to win \$30 and 90\% chance to
-win \$0, participants felt little disappointment with the \$0 outcome,
-but were considerably more excited when they won \$30. Conversely, given
-a gamble with a 90\% chance of winning \$60 and 10\% chance of winning
-\$0, participants were disappointed with the \$0 outcome and showed more
-muted enjoyment of the \$60 outcome. In fact, in gambles similar to
-these, people were happier with the smaller unexpected gain than with
-the larger but more expected gain (6).
+task: Participants were often more excited by an small unexpected gain
+(e.g., winning \$30 with 10\% probability) than they were by a larger
+but more expected gain (e.g., winning \$60 with 90\% probability) (6).
 
-The influence of expectations on moral evaluations threatens to produce
-negative social consequences. Tragedies are tragic, whether expected or
-not, and failure to recognize that could lead to tacit endorsement of
+It seems intuitive that surprise would shape people's reactions to
+parties, sporting events, and games of chance. However, it seems wholly
+inappropriate that surprise would shape people's reactions to events
+with serious moral implications. Tragedies are tragic, whether expected
+or not, and failure to recognize this could lead to tacit endorsement of
 the pain and suffering of others wherever pain and suffering are the
-status quo. Do expectations directly bias reactions to morally
-consequential events, as intuition and anecdote seem to suggest? To
+status quo. If expectations truly do influence our reactions to
+morally-significant world events, as intuition and anecdote seem to
+suggest, there are potentially serious social consequences. To
 demonstrate the plausibility of this suggestion, we develop a
@@
 
-Most events of consequence consist of changes--a transition from one
-state to another. For this reason, we construe event evaluation as a
-comparison between the states of the world before and after the event.
-Comparison plays a crucial role in many theories of decision-making, and
-especially in theories aimed at explaining the role of expectations in
-evaluation (4, 5, 7, 8). For instance, in Prospect Theory (9, 10), the
-utility of a prospect is evaluated by comparison to a current reference
-point. Our proposal is similar: we argue that expectations can set the
-reference points against which people compare future outcomes, so that
-assigning value to an event consists in comparing the state of the world
-following the event to the state of the world just prior to the event.
-We further assume that people's knowledge of the world is uncertain, so
-we represent the states of the world probabilistically. This means that
-a person's model for the state of the world is just their expectations
-about what has happened, what is happening, and what will happen in the
-future. One direct implication of this proposal is that the evaluation
-of an event is linked to what is learned about the world as a result of
-that event's occurrence. That is, we evaluate events positively or
-negatively to the extent that they inform our understanding of the world
-in positive or negative ways.
+Events, as we will conceive of them, are changes in the state of the
+world---transitions from one state to another. For this reason, we
+construe event evaluation as a comparison between the states of the
+world before and after the event. Similar comparison processes play
+crucial roles in many theories of decision-making (4, 5, 7, 8). For
+instance, in Prospect Theory (9, 10), the utility of a prospect is
+evaluated by comparison to a current reference point. Our proposal is
+similar: we argue that expectations can set the reference points against
+which people compare future outcomes, so that assigning value to an
+event consists in comparing the state of the world following the event
+to the state of the world just prior to the event. Further, we assume
+that people's knowledge of the world is uncertain, so we represent the
+states of the world probabilistically. This means that a person's model
+for the state of the world is simply their expectations about what has
+happened, what is happening, and what will happen in the future. A
+direct implication of this proposal is that the evaluation of an event
+is linked to what is learned about the world as a result of that event's
+occurrence. That is, we evaluate events positively or negatively to the
+extent that they inform our understanding of the world in positive or
+negative ways.
 
@@
 world events as compared to when they are affirmed by less surprising
-events. For example, if a bombing occurs in Paris--a less expected
-location--rather than Lebanon--a more expected location--we learn that
-the world is more dangerous than we had previously believed. This
-suggestion is novel in the moral domain, but these points are widely
-accepted elsewhere. For instance, expectations have long been recognized
-as fundamental to models of human and animal learning (12).
+events. For example, compared to a bombing in Lebanon, if a bombing
+occurs in Paris, a less expected location, we learn that the world is
+more dangerous than we had previously believed. Although this suggestion
+is novel in the moral domain, these points are widely accepted
+elsewhere. For instance, expectations have long been recognized as
+fundamental to models of human and animal learning (12).
 
@@
 different values of \(S\), weighted by the inverse of the prior
-probability of \(S\)..\footnote{It should be recognized that the result
+probability of \(S\).\footnote{It should be recognized that the result
   we obtain here bears a similarity to the ``disappointment function''
@@
 expectations will equally affect reactions to gambles with clearly
-specified values and probabilities (5) and real-world events with
-ill-defined values and where reasoners' expectations are defined by
+specified values and probabilities (5), as well as real-world events
+with ill-defined values, where reasoners' expectations are defined by
 their existing beliefs and world knowledge.
 
-Given the potentially fundamental role expectations play in event
+Given the potentially fundamental role that expectations play in event
 evaluation, and the potentially serious moral biases that may result, we
@@
 in moral evaluations. In both studies, participants were presented with
-a series of trials where they read brief descriptions of two different
-events and were asked to indicate which of the two events seemed more
-upsetting. In these trials, the two events were highly similar, but were
-manipulated so that they differed in their perceived prior
-probabilities: one event was more expected and one more unexpected.
+a series of trials in which they read brief descriptions of two
+different events and were asked to indicate which of the two events
+seemed more upsetting. In these trials, the two events were highly
+similar, but were manipulated so that they differed in their perceived
+prior probabilities: one event was more expected and one more
+unexpected.
 
@@
 theoretical points. First, the information-theoretic account we have
-described makes a solely directional prediction--all else being equal,
+described makes a solely directional prediction---all else being equal,
 more surprising events will elicit stronger reactions than less
@@
 informativeness of events, determined by one's prior expectations of the
-event, that we predict will shape evaluations. Thus, reasoners should
+event, that we predicted would shape evaluations. Thus, reasoners should
 not have to be induced to hold expectations experimentally (5), but
-instead their own prior beliefs will be sufficient to shape their
+instead their own prior beliefs should be sufficient to shape their
 expectations and affect their evaluations. For this reason, we
@@
 contexts that differed in ways that could not be accounted for by
-construal-level explanations ((1)) or explanations based on biases
-towards one's ingroup (\textbf{???}). This allowed us to test the role
-of expectations based on world knowledge, as well as providing a more
-naturalistic set of items.
+construal-level explanations (1) or explanations based on biases towards
+one's ingroup (14). This both allowed us to test the role of
+expectations based on world knowledge and provided a more naturalistic
+set of items.
 
@@
 
-Consistent with our predictions, manipulating expectations consistently
-shaped evaluations across a range of morally-significant events. We
-found that people viewed unexpected harmful events as more upsetting
-than expected harmful events, even when the harm a victim suffered was
-the same. For instance, people judged that a robbery at a clothing store
-(unexpected) was more upsetting than a robbery at a convenience store
-(expected). Furthermore, these differences were large, systematic, and
-robust. Although there was variability in the situations that elicited
-the largest differences, our statistical approach treated trial as a
-random effect (see Methods and Results), allowing us to generalize our
-conclusions to items we did not test (14).
+Consistent with our predictions, manipulating expectations shaped
+evaluations across a range of morally-significant events. We found that
+people viewed unexpected harmful events as more upsetting than expected
+harmful events, even when the harm a victim suffered was the same. For
+instance, people judged that a robbery at a clothing store (unexpected)
+was more upsetting than a robbery at a convenience store (expected).
+Furthermore, these differences were large, systematic, and robust.
+Although there was variability in the situations that elicited the
+largest differences, our statistical approach treated trial as a random
+effect (see Methods and Results), allowing us to generalize our
+conclusions to items we did not test (15).
 
@@
 scale has been constrained to show relevant region of results. In both
-plots, response of 3 indicates indifference, a response below 3
-indicates a bias to choose the more expected event as more upsetting,
-and a response above 3 indicates a bias to choose the less expected
-event as more upsetting. Similar to Study 1, when participants judged
-that one of the events was more upsetting, they showed a robust bias
+plots, a response of 3 indicates indifference, responses below 3
+indicate a bias to choose the more expected event as more upsetting, and
+responses above 3 indicate a bias to choose the less expected event as
+more upsetting. Similar to Study 1, participants showed a robust bias
 toward choosing the less expected event as more upsetting. Trials are
@@
 
-There's a sense in which it is clearly adaptive that the violation of
-our expectations would direct our attention to the events and situations
-about which our understanding is most in need of revision. For instance,
-an act of terrorism in a peaceful nation might be less expected because
-of our beliefs about the effectiveness of security forces, the stability
-of national relationships, and positive relationships among people
-groups in that nation. Therefore, a surprising act of terrorism casts
-doubt on our deeper understanding of the geopolitical situation, and
-thereby leads us to suspect similar acts are more likely to occur in the
-future.
+In some contexts, it is clearly adaptive that the violation of our
+expectations would direct our attention to the events and situations
+about which our understanding is most in need of revision (12). For
+instance, an act of terrorism in a peaceful nation might be less
+expected because of our beliefs about the effectiveness of security
+forces, the stability of national relationships, and positive
+relationships among people groups in that nation. Therefore, a
+surprising act of terrorism casts doubt on our deeper understanding of
+the geopolitical situation, and thereby leads us to suspect similar acts
+are more likely to occur in the future.
 
-Yet, however well-grounded the underlying principles, there's a clear
-moral danger whenever expectations influence our moral concerns (15,
-16). These effects pose a significant threat when considering our
+Yet, however well-grounded the underlying principles, there is a clear
+moral danger whenever expectations influence our moral concerns. In
+particular, these effects pose a significant threat when considering our
 reactions to humanitarian challenges brought on by poverty or regional
@@
 famine, and natural disasters, and are the very people in greatest need
-of assistance and concern from the world at-large. However, for these
-very reasons, it is often unsurprising when harm befalls people living
-in these circumstances. Thus the influence of expectations on moral
-evaluations threatens to reduce observer's concern for these victims
-most in need, making people less likely to donate time or money to aid
-victims, or to take political action to prevent further harms. The
-present research identifies this potentially pernicious moral bias and
-grounds it in information theoretic principles. Hopefully, by further
-understanding these cognitive dynamics, psychologists and policymakers
-can begin to find ways to reduce or counteract people's tendency to
-ignore the plight of those in most need of attention.
+of assistance and concern from the world at-large. However, for
+precisely these reasons, it is often unsurprising when harm befalls
+people living in these circumstances. Thus the influence of expectations
+on moral evaluations threatens to reduce observer's concern for these
+victims most in need, making people less likely to donate time or money
+to aid victims, or to take political action to prevent further harms.
+The present research identifies this potentially pernicious moral bias
+and grounds it in information theoretic principles. Hopefully, by
+further understanding these cognitive dynamics, psychologists and
+policymakers can begin to find ways to reduce or counteract people's
+tendency to ignore the plight of those in most need of attention.
 
@@
 We conducted two preregistered studies examining the role of
-expectations in moral evaluations (\url{https://osf.io/5qnf7/}). Each
+expectations in moral evaluations (\url{https://osf.io/86rsw/}). Each
 study consisted of multiple trials where participants were presented
@@
 their responses. All experimental materials for these studies are
-available as Supporting Information at \url{https://osf.io/5qnf7/}.
+available as Supporting Information at \url{https://osf.io/86rsw/}.
 
@@
 We analyzed our data by performing Bayesian estimation using the
-probabilistic programming language Stan (17). We tested our
-preregistered predictions by computing Bayes Factors (i.e.~BF01) using
-the hypothesis function in the R package brms. Bayes Factors express the
-ratio of the probability of data under the null hypothesis to the
-probability of the data under an alternative hypothesis. Larger Bayes
-Factors indicate that the data are more likely under the null hypothesis
-(i.e., that the intercept is not different from zero) than the
-alternative hypothesis (i.e., that the intercept is different from
+probabilistic programming language Stan (16). We tested our
+preregistered predictions by computing Bayes Factors (i.e. \(BF_{01}\))
+using the hypothesis function in the R package brms. Bayes Factors
+express the ratio of the probability of data under the null hypothesis
+to the probability of the data under an alternative hypothesis. Larger
+Bayes Factors indicate that the data are more likely under the null
+hypothesis (i.e., that the intercept is not different from zero) than
+the alternative hypothesis (i.e., that the intercept is different from
 zero), and vice versa. As Bayes Factors can be influenced by prior
-choices (Gelman, Simpson, \& Betancourt, 2017), we performed prior
-robustness checks to confirm that our conclusions would not depend
-substantially on the specification of the priors.
+choices (17), we performed prior robustness checks to confirm that our
+conclusions would not depend substantially on the specification of the
+priors.
 
@@
 wider population, much as we generalize from our sample of participants
-to the larger population (14).
+to the larger population (15).
 
@@
 events were more upsetting than events that were expected, Intercept =
-1.424, 95\% CI {[}1.147, 1.707{]}, BF01 \textless{} .001. Figure 1 shows
-participants' responses broken-down by individual items. Participants'
-bias toward selecting the unexpected event as more upsetting was
-consistent across the 16 experimental items.
+1.424, 95\% CI {[}1.147, 1.707{]}, \(BF_{01}\) \textless{} .001. Figure
+1 shows participants' responses broken-down by individual items.
+Participants' bias toward selecting the unexpected event as more
+upsetting was consistent across the 16 experimental items.
 
@@
 or higher) to the inverse of the second intercept coefficient (thereby
-representing the log-odds \emph{not} choosing a 3 or higher--i.e.,
+representing the log-odds \emph{not} choosing a 3 or higher---i.e.,
 choosing a 1 or 2), allowing us to test whether participants were more
@@
 that were unexpected were more upsetting than events that were expected,
-BF01 \textless{} .001 (see supporting information for full model
+\(BF_{01}\) \textless{} .001 (see supporting information for full model
 results). We observed this difference consistently across nearly every
@@
 \addcontentsline{toc}{section}{Theoretical Formalization}
-
-Evaluation of utility is often examined in the context of
-decision-making. However, people often generate post hoc evaluations of
-events for their own sake. For instance, people evaluate events to judge
-the wisdom of their decisions or the morality of their actions.
 
@@
 
-Thus, it can be seen under this formalization that the role of
-expectations in the evaluation of events is fundamental in the sense
-that it does not depend on any specifics concerning the utility function
-over states of the world. Instead, this conclusion depends on just two
-assumptions: first, that events are evaluated by comparing the utilities
-of the states of the world prior to and following the event; and second,
-that knowledge of those states of the world is uncertain.
+Under this formalization, the role of expectations in the evaluation of
+events does not depend on any further assumptions about the states of
+the world or the utility function to be applied over them. Instead, this
+conclusion depends on just two assumptions: first, that events are
+evaluated by comparing the utilities of the states of the world prior to
+and following the event; and second, that knowledge of those states of
+the world is uncertain.
 
@@
 }
-
 \showmatmethods
@@
 
+\hypertarget{ref-Brewer1999}{}
+14. Brewer MB (1999) The psychology of prejudice: Ingroup love or
+outgroup hate? \emph{Journal of Social Issues} 55(3):429--444.
+
 \hypertarget{ref-Judd2012}{}
-14. Judd CM, Westfall J, Kenny DA (2012) Treating stimuli as a random
+15. Judd CM, Westfall J, Kenny DA (2012) Treating stimuli as a random
 factor in social psychology: A new and comprehensive solution to a
@@
 
-\hypertarget{ref-Koss2000}{}
-15. Koss MP (2000) Blame, shame, and community: Justice responses to
-violence against women. \emph{American Psychologist} 55(11):1332--1343.
-
-\hypertarget{ref-Janoff-bulman1985}{}
-16. Janoff-Bulman R, Timko C, Carli LL (1985) Cognitive biases in
-blaming the victim. \emph{Journal of Experimental Social Psychology}
-21:161--177.
-
 \hypertarget{ref-Carpenter2017}{}
-17. Carpenter B, et al. (2017) Stan : A Probabilistic Programming
+16. Carpenter B, et al. (2017) Stan : A Probabilistic Programming
 Language. \emph{Journal of Statistical Software} 76(1).
 doi:\href{https://doi.org/10.18637/jss.v076.i01}{10.18637/jss.v076.i01}.
+
+\hypertarget{ref-Gelman2017}{}
+17. Gelman A, Simpson D, Betancourt M (2017) The Prior Can Often Only Be
+Understood in the Context of the Likelihood. \emph{Entropy} 19(10):555.
 

### 2018-09-04 paper/appendix.Rmd
paper in apa format

--- 2018-07-18 paper/paper.Rmd
@@
----
-title: Expectations bias moral evaluations
-
-# Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
-author:
-  - name: Derek Powell
-    affiliation: a,1
-  - name: Zachary Horne
-    affiliation: b
-address:
-  - code: a
-    address: Stanford University, Department of Psychology, 450 Serra Mall, Stanford, CA, 94305
-  - code: b
-    address: Arizona State University, School of Social and Behavioral Sciences, 13591 N 47th Ave, Phoenix, AZ 85051
-
-corresponding_author:
-  - code: 1
-    text: "To whom correspondence should be addressed. E-mail: derekpowell@stanford.edu"
-
-# For footer text
-lead_author_surname: Powell
-
-## Remove this if not required - talk to zach 
-# equal_authors:
-#   - code: 0
-#     text:
-
-author_contributions: |
-  Both authors jointly conceived of the project and theory, designed and conducted experiments, and analyzed results. D.P. developed the formal aspects of the theory.
-
-## Remove this if not required
-conflict_of_interest: |
-  The authors declare no conflicts of interest.
-
-abstract: |
- When we learn of tragic events--a violent killing or a terrorist attack--we are mournful and often driven to action.
-But responses to these events can differ substantially depending on the context in which they occur.
-When tragedy takes us by surprise, we are shocked and outraged, but when it strikes in unstable regions or rough neighborhoods, we are often resigned to feel that “these things happen.” Here we explain the role that expectations seem to play in moral evaluation by modeling evaluations as the comparison of the state of the world before and after an event.
-According to this theory, expectations influence evaluations through information-theoretic principles: less expected events do more to inform us about the state of the world than do more expected events.
-Consistent with this theory, in two preregistered studies we found that people’s judgments of morally-significant events were affected by the prior probability of that event.
-People were more upset about an event that was unexpected (e.g., a robbery at a clothing store) than an event that was more more probable (e.g., a robbery at a convenience store).
-This bias threatens to impose a pernicious cycle: People living in poverty, geo-politically unstable regions, and other dangerous circumstances have the greatest need for assistance and concern.
-However, for these very reasons, it is often unsurprising when they are harmed.
-Consequently, the influence of expectations on moral evaluations threatens to reduce observers’ concern for those most in need.
-
-significance: |
-  In two behavioral studies, we found that people's expectations biased their reactions to morally-significant events--people were more upset by surprising harms than more expected harms.
-This bias threatens to impose a pernicious cycle: People living in poverty, unstable regions, and other dangerous circumstances have the greatest need for assistance and concern from the world at-large.
-Yet for this very reason, it is often unsurprising when they are harmed.
-By blunting observers' concern for the disadvantaged, this bias could make people less likely to donate money, to volunteer, or to take political action, thereby increasing inequality and compounding suffering.
-Recognizing and understanding this bias will suggest ways to mitigate its harmful consequences.
-
-acknowledgements: |
-  We would like to acknowledge the help of Larisa Hussak, Keith Holyoak, Alan Fiske, Hongjing Lu, John Hummel, Andrei Cimpian, and Ellen Markman for their comments and support.
-
-keywords:
-  - evaluation
-  - moral judgment
-  - information theory
-  # - optional
-  # - optional
-
-## must be one of: pnasresearcharticle (usual two-column layout), pnasmathematics (one column layout), or pnasinvited (invited submissions only)
-pnas_type: pnasresearcharticle
-
-bibliography: wiw-biblio.bib
-csl: pnas.csl
-
-## change to true to add optional line numbering
-lineno: FALSE
-
-output: rticles::pnas_article
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-```{r load data, include=FALSE}
-# analysis
-
-library(tidyverse)
-
-d_raw <- read.csv("../data/WIW+Judgment+Task+Preregistered_March+30%2C+2018_11.17.csv")
-
-d_wide <- d_raw %>%
-  filter(DistributionChannel=="anonymous") %>%
-  rename(duration = Duration..in.seconds.) %>%
-  select(workerId, age, sex, attn, duration, contains("_trial"), -contains("_trial_")) %>%
-  filter(Check_trial == 0, attn == 1) %>%
-  rename(
-    Corruption_trial = corruption_trial,
-    Clerk_trial = clerk_trial,
-    Flattire_trial = flattire_trial,
-    Rape_trial = rape_trial) %>%
-  mutate(sex = if_else(sex == 1, "male", "female"))
-
-reverse_items <- c( # sadly, only 6 of 10 items were reversed! :(
-  "Cancer_trial",
-  "Heater_trial",
-  "Flattire_trial",
-  "Clerk_trial",
-  "Corruption_trial",
-  "Rape_trial"
-)
-
-d_tidy <- d_wide %>%
-  gather(trial, response, Earthquake_trial:BigDiff5_trial) %>%
-  mutate(trial_type = ifelse(grepl("[0-9]", trial), "filler", "experimental")) %>%
-  mutate(trial_type = ifelse(grepl("BigDiff",trial), "filler-diff", trial_type)) %>%
-  mutate(likely_2nd = ifelse((trial %in% reverse_items), 1, 0))
-
-d <- d_tidy %>%
-  filter(trial_type == "experimental", trial != "Check_trial") %>%
-  mutate(response = ifelse(likely_2nd==1, (1-response), response)) %>%
-  mutate(response = abs(response-1)) %>%
-  mutate(trial = gsub("_trial","",trial))
-
-d2_raw <- read.csv("../data/WIW+Judgment+Task+Preregistered+-+Likert_April+20%2C+2018_17.08.csv")
-
-d2_wide <- d2_raw %>%
-  filter(DistributionChannel=="anonymous") %>%
-  rename(duration = Duration..in.seconds.) %>%
-  select(workerId, age, sex, attn, duration, contains("_trial"), -contains("_trial_")) %>%
-  filter(
-    Check1_trial == -1, 
-    Check2_trial == -2, 
-    Check3_trial == 1, 
-    attn == 1,
-    !is.na(workerId)
-    ) %>%
-  rename(
-    Corruption_trial = corruption_trial,
-    Clerk_trial = clerk_trial,
-    Flattire_trial = flattire_trial,
-    Rape_trial = rape_trial) %>%
-  mutate(sex = if_else(sex == 1, "male", "female"))
-
-# normal = more likely is 1
-# reverse = less likely is 1
-
-reverse_items2 <- c(
-  "Cancer_trial",
-  "Heater_trial",
-  "Flattire_trial",
-  "Clerk_trial",
-  "Corruption_trial",
-  "Rape_trial",
-  "Earthquake_trial",
-  "Fastfood_trial",
-  "Police_trial"
-)
-
-d2_tidy <- d2_wide %>%
-  select(-Check1_trial, -Check2_trial, -Check3_trial) %>%
-  gather(trial, response, Tornado_trial:BigDiff5_trial) %>%
-  mutate(trial_type = ifelse(grepl("[0-9]", trial), "filler", "experimental")) %>%
-  mutate(trial_type = ifelse(grepl("BigDiff",trial), "filler-diff", trial_type)) %>%
-  mutate(likely_2nd = ifelse((trial %in% reverse_items2), 1, 0))
-
-d2 <- d2_tidy %>%
-  filter(trial_type == "experimental") %>%
-  mutate(response = ifelse(likely_2nd==1, response*-1, response)) %>%
-  mutate(response = -1*response) %>%
-  mutate(response = response + 3) %>%
-  mutate(trial = gsub("_trial","",trial))
-```
-
-```{r compute descriptives, include=FALSE}
-gender_exp1 <- d %>% 
-  group_by(workerId) %>% 
-  summarize(sex=first(sex)) %>% 
-  ungroup() %>% 
-  count(sex) %>% 
-  spread(sex, n)
-
-gender_exp2 <- d2 %>% 
-  group_by(workerId) %>% 
-  summarize(sex=first(sex)) %>% 
-  ungroup() %>% 
-  count(sex) %>% 
-  spread(sex, n)
-
-age_exp1 <- d %>% 
-  group_by(workerId) %>% 
-  summarize(age=first(age)) %>% 
-  ungroup() %>% 
-  summarize(age = median(age)) %>% 
-  .$age
-
-age_exp2 <- d2 %>% 
-  group_by(workerId) %>% 
-  summarize(age=first(age)) %>% 
-  ungroup() %>% 
-  summarize(age = median(age)) %>% 
-  .$age
-```
-
-```{r fit brms models, include=FALSE}
-library(brms)
-devtools::source_gist(id = "f1994c0f8325abbc5d300600744af39d", filename="cbrm.R")
-
-fit_exp1 <- cbrm(
-  response ~ 0 + intercept + trial + (1|workerId), # another model adds likely_2nd
-  data = d,
-  family = bernoulli(),
-  prior = prior(normal(0,3), class="b"),
-  control = list(adapt_delta = .80),
-  cores = parallel::detectCores(),
-  sample_prior = TRUE,
-  iter = 2000,
-  cached_file = "fit_exp1.rds"
-  )
-
-fit_exp2 <- cbrm(
-  response ~ trial + (1|workerId), # another model adds likely_2nd
-  data = d2,
-  family = cumulative(),
-  prior = prior(normal(0,3), class="Intercept"),
-  control = list(adapt_delta = .80),
-  cores = parallel::detectCores(),
-  sample_prior = TRUE,
-  iter = 2000,
-  cached_file = "fit_exp2.rds"
-  )
-
-fit_exp1rand <- cbrm(
-  response ~ 0 + intercept + (1|trial) + (1|workerId), # another model adds likely_2nd
-  data = d,
-  family = bernoulli(),
-  prior = prior(normal(0,3), class="b"),
-  control = list(adapt_delta = .80),
-  sample_prior = TRUE,
-  cores = parallel::detectCores(),
-  iter = 2000,
-  cached_file = "fit_exp1rand.rds"
-  )
-
-fit_exp2rand <- cbrm(
-  response ~ (1|trial) + (1|workerId), # another model adds likely_2nd
-  data = d2,
-  family = cumulative(),
-  prior = prior(normal(0,3), class="Intercept"),
-  control = list(adapt_delta = .80),
-  cores = parallel::detectCores(),
-  sample_prior = TRUE,
-  iter = 2000,
-  cached_file = "fit_exp2rand.rds"
-  )
-```
-
-```{r hypothesis tests, include=FALSE}
-print_bf <- function(hypothesis) {
-
-  bf <- hypothesis$hypothesis$Evid.Ratio
-
-  if (is.na(bf)) {
-    return("< .001")
-  }
-  else if (bf < .001) {
-    return("< .001")
-  }
-  else if (bf > 1000) {
-    return("> 1000")
-  }
-  else {
-    return(paste("=", as.character(bf)))
-  }
-}
-
-coefs_exp1 <- broom::tidy(fit_exp1rand) %>%
-  filter(term=="b_intercept")
-
-# coefs_exp2 <- broom::tidy(fit_exp2) %>%
-#   filter(term=="b_intercept")
-
-bf_exp1 <- hypothesis(fit_exp1rand, "intercept = 0")
-# bf_exp1 <- h1$hypothesis$Evid.Ratio
-
-bf_exp2 <- hypothesis(fit_exp2rand, "-Intercept[2] = Intercept[3]")
-# bf_exp2 <- h2$hypothesis$Evid.Ratio
-```
-
-```{r make figures, include=FALSE}
-fig1 <- plot(marginal_effects(fit_exp1, "trial"))
-fig_exp1 <- fig1$trial$data %>%
-  ggplot() +
-  aes(x = reorder(trial, -estimate__), y = estimate__, ymin = lower__, ymax = upper__) +
-  geom_errorbar(width=.25) +
-  geom_point(shape=17, size = 2) +
-  coord_cartesian(ylim=c(0,1)) +
-  geom_hline(yintercept = .5, linetype = "dashed", color = "grey") +
-  theme_bw() +
-  theme(
-    panel.grid.major = element_blank(),
-    panel.grid.minor = element_blank(),
-    aspect.ratio = .75,
-    axis.text.x = element_text(angle = 45, hjust = 1, vjust = .7)
-    ) +
-  labs(
-    x = "Trial",
-    y = "Prob. less expected\njudged more upsetting"
-  ) 
-# library(gridExtra)
-# grid.arrange(ggplotGrob(fig_exp1))
-
-ggsave("fig1.eps",fig_exp1, height=8.7, width=8.7, unit="cm")
-```
-
-```{r fig2, include=FALSE}
-
-fig2_cont <- marginal_effects(fit_exp2, "trial", plot=FALSE)
-fig2 <- plot(marginal_effects(fit_exp2, "trial", ordinal=TRUE))
-
-plt_cont <- fig2_cont$trial %>%
-  ggplot() +
-  aes(x = reorder(trial, -estimate__), y = estimate__, ymin = lower__, ymax = upper__) +
-  geom_errorbar(width=0) +
-  geom_point(shape=17, size = 1) +
-  coord_cartesian(ylim=c(2,4)) +
-  geom_hline(yintercept = 3, linetype = "dashed", color = "grey") +
-  theme_bw() +
-  theme(
-    panel.grid.major = element_blank(),
-    panel.grid.minor = element_blank(),
-    aspect.ratio = 10/16,
-    axis.text.x = element_text(angle = 45, hjust = 1, vjust = .7)
-  ) +
-  labs(
-    x = "Trial",
-    y = 'Upset Judgments'
-  )
-
-plt_heatmap <- fig2$trial$data %>%
-  left_join(
-    fig2_cont$trial %>% 
-              rename(mean_estimate = estimate__) %>%
-              select(trial, mean_estimate)
-            , by = "trial"
-    ) %>%
-  ggplot(aes(y = cats__, x = reorder(trial, -mean_estimate), fill = estimate__)) +
-  geom_tile(color="white") + 
-  scale_fill_viridis_c(
-    option="B", 
-    limits=c(0,1), 
-    breaks = c(0, .25, .5, .75, 1), 
-    labels= c(0, .25, .5, .75, 1),
-    guide = guide_colorbar(
-      title.position="top", 
-      barheight = unit(.5, "cm")
-      )
-    ) +
-  theme_minimal() +
-  theme(
-    panel.grid.major = element_blank(),
-    panel.grid.minor = element_blank(),
-    # aspect.ratio = 1,
-    legend.position = "top",
-    legend.key.width = unit(.75,"cm"),
-    axis.text.x = element_text(angle = 45, hjust = 1, vjust = .7)
-  ) +
-  labs(
-    x = "Trial",
-    y = "Response",
-    fill = "Probability"
-  ) +
-  coord_fixed(ratio=1)
-
-library(gtable)
-g2 <- ggplotGrob(plt_cont + theme(
-  axis.title.y = element_text(margin = margin(t=0,b=.5,l=0,r=.25,unit="cm")) # this is a kludge to get shared axis
-))
-g3 <- ggplotGrob(plt_heatmap + theme(
-  axis.title.x = element_blank(),
-  axis.title.y = element_blank(),
-  axis.text.x = element_blank()
-  ))
-
-g <- rbind(g3, g2, size = "last")
-# grid.arrange(g)
-
-library(gridExtra)
-
-ggsave("fig2.eps", grid.arrange(g), height=11.4, width=8.7, units = "cm")
-```
-
-When we learn of tragic events--a friend's loss of a loved-one, a violent killing, or a terrorist attack--we are mournful, concerned, angry, and driven to action.
-But our responses to these events differ depending on the context in which they occur [e.g., @Trope2010].
-In particular, our expectations play an important role in our reactions to events.
-This seems true of the ordinary: when a friend's loved-one passes, we often ask, "was it sudden?"; the dramatic: a murder in a sleepy town prompts a very different reaction than a similar crime in a major city; and even the catastrophic: acts of terrorism seem to evoke more outrage, horror, and concern when they occur in peaceful rather than unstable regions.
-For instance, the 2015 Paris attacks were felt around the world.
-Yet, most of those mourning had little to say 15 hours earlier, when another tragic attack killed at least 43 people in Beirut [@Graham2015].
-Likewise, stories of the Virginia Tech massacre, which killed 32 people, wildly overshadowed coverage of attacks killing nearly 200 people in Iraq--one of the bloodiest days since the 2003 invasion [@Greenslade2007].
-In each of these situations, the surprise elicited by an event appears to have affected people’s reactions to it.
-In some contexts, harm is tragic.
-In others, “these things happen.”
-
-Although this pattern seems so intuitive as to be obvious, no reasonable code of ethics would justify these judgments, nor do any moral psychological frameworks explicate the role of expectations in moral evaluations.
-Outside of the domain of morality, everyday experiences and controlled experiments show that people’s evaluations of events unambiguously depend on their expectations about those events.
-There is a special thrill to having one’s expectations exceeded and disappointment when events fail to meet expectations: we root for the underdog, hold surprise parties, and foreshadow bad news to ease its delivery [@Bell1985].
-Indeed, in a controlled study, Mellers and colleagues [-@Mellers1997] found that expectations influenced affective reactions during a gambling task: Participants were often more excited by an small unexpected gain (e.g., winning \$30 with 10% probability) than they were by a larger but more expected gain (e.g., winning \$60 with 90% probability) [also see @Shepperd2002].
-
-It seems intuitive that surprise would shape people’s reactions to parties, sporting events, and games of chance.
-However, it seems wholly inappropriate that surprise would shape people’s reactions to events with serious moral implications.
-Tragedies are tragic, whether expected or not, and failure to recognize this could lead to tacit endorsement of the pain and suffering of others wherever pain and suffering are the status quo.
-If expectations truly do influence our reactions to morally-significant world events, as intuition and anecdote seem to suggest, there are potentially serious social consequences.
-To demonstrate the plausibility of this suggestion, we develop a theoretical account of how expectations influence evaluations, not as a strange quirk or uniquely human bias, but instead as a fundamental component of how people evaluate events.
-
-Events, as we will conceive of them, are changes in the state of the world--transitions from one state to another.
-For this reason, we construe event evaluation as a comparison between the states of the world before and after the event.
-Similar comparison processes play crucial roles in many theories of decision-making [e.g., @Bell1985; @Gul1991; @Loomes1986; @Mellers1997].
-For instance, in Prospect Theory [@Kahneman1979; @Tversky1992], the utility of a prospect is evaluated by comparison to a current reference point.
-Our proposal is similar: we argue that expectations can set the reference points against which people compare future outcomes, so that assigning value to an event consists in comparing the state of the world following the event to the state of the world just prior to the event.
-Further, we assume that people’s knowledge of the world is uncertain, so we represent the states of the world probabilistically.
-This means that a person’s model for the state of the world is simply their expectations about what has happened, what is happening, and what will happen in the future.
-A direct implication of this proposal is that the evaluation of an event is linked to what is learned about the world as a result of that event’s occurrence.
-That is, we evaluate events positively or negatively to the extent that they inform our understanding of the world in positive or negative ways.
-
-One striking implication of this account is that it provides an information-theoretic route to understanding how expectations directly influence evaluation.
-A fundamental insight of Information Theory is that the information carried by an event is a function of its prior probability; low probability events carry more information than high probability events [@Shannon1948].
-We propose that people learn more about the state of the world when their expectations are violated by shocking world events as compared to when they are affirmed by less surprising events.
-For example, compared to a bombing in Lebanon, if a bombing occurs in Paris, a less expected location, we learn that the world is more dangerous than we had previously believed.
-Although this suggestion is novel in the moral domain, these points are widely accepted elsewhere.
-For instance, expectations have long been recognized as fundamental to models of human and animal learning [e.g., @Rescorla1972].
-
-A more formal treatment of this theory establishes the potential for its generality across domains.
-Let the state of the world prior to some event be represented by a random variable, $X$.
-We'll assume $X$ is discrete, and that every realization $x$ represents some potential specific state of the world in all relevant aspects under consideration.
-An agent has some utility function $u$, that applies over the different realizations of $X$.
-Thus the expected utility of the present state of the world is:
-
-\begin{equation*}
-E_X[u(X)] = \sum_x u(x)p(x)
-\end{equation*}
-
-When we learn an event has occurred, we can represent this as an observation of a second, binary random variable, $S$.
-We can think of $S$ as a "signalling event" that informs us about about $X$ [following the formalism deployed by @Arrow1996].
-That is, we assume there is a dependency between $S$ and $X$, so that observing $S$ informs us about the state of the world, $X$.
-For instance, reading a news report of a terror attack in Brussels informs us about the state of the world, telling us about what realizations of $X$ are no longer possible (e.g., any wherein it was a peaceful day in Brussels) and which are more or less likely.
-Conditioned on $S$, the expected utility of the world is:
-
-\begin{equation*}
-E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
-\end{equation*}
-
-Our construal of event evaluation is as a comparison between states of the world following and prior to an event. Consequently, we define the value assigned to the signalling event, $V(S)$, as:
-
-\begin{equation*}
-V(S) = E_{X|S}[u(X)] - E_X[u(X)]
-\end{equation*}
-
-From this it can be shown that the value assigned to event $S$ is proportional to its prior probability (see Theoretical Formalization).
-Specifically, it is the difference in the expected utility under different values of $S$, weighted by the inverse of the prior probability of $S$.
-[^1]
-
-\begin{equation*}
-V(s=1)=(1-p(s=1))\sum_x u(x)p(x|s=1) - u(x)p(x|s=0)
-\end{equation*}
-
-This conclusion does not depend on the nature of the event in question nor on the utility function over events, but instead holds generally wherever we conceive of evaluation as a holistic comparison between uncertain states of the world.
-Consequently, this account predicts expectations will equally affect reactions to gambles with clearly specified values and probabilities [@Mellers1997], as well as real-world events with ill-defined values, where reasoners' expectations are defined by their existing beliefs and world knowledge.
-
-Given the potentially fundamental role that expectations play in event evaluation, and the potentially serious moral biases that may result, we sought to test empirically whether people’s evaluations of morally harmful events are affected by their expectations about those events.
-We conducted two preregistered studies examining the role of expectations in moral evaluations.
-In both studies, participants were presented with a series of trials in which they read brief descriptions of two different events and were asked to indicate which of the two events seemed more upsetting.
-In these trials, the two events were highly similar, but were manipulated so that they differed in their perceived prior probabilities: one event was more expected and one more unexpected.
-
-The design of these studies was chosen with special consideration to two theoretical points.
-First, the information-theoretic account we have described makes a solely directional prediction--all else being equal, more surprising events will elicit stronger reactions than less surprising events.
-Asking participants to compare two similar events allowed us to test this directional prediction.
-Second, it is the informativeness of events, determined by one's prior expectations of the event, that we predicted would shape evaluations.
-Thus, reasoners should not have to be induced to hold expectations experimentally [as in experiments by @Mellers1997], but instead their own prior beliefs should be sufficient to shape their expectations and affect their evaluations.
-For this reason, we manipulated the prior probability of events by describing two otherwise identical events occurring in different contexts.
-We explicitly chose contexts that differed in ways that could not be accounted for by construal-level explanations [@Trope2010] or explanations based on biases towards one’s ingroup [@Brewer1999].
-This both allowed us to test the role of expectations based on world knowledge and provided a more naturalistic set of items.
-
-![Probability with which participants chose the less expected event as more upsetting across trials in Study 1.
-As predicted, participants exhibited a robust bias toward choosing the less expected event as more upsetting across all items.
-Error bars represent 95% credible intervals calculated from the hierarchical logistic regression model.
-See Supporting Information for full description of items.
-<span data-label="fig:fig1"></span>](fig1.eps)
-
-Consistent with our predictions, manipulating expectations shaped evaluations across a range of morally-significant events.
-We found that people viewed unexpected harmful events as more upsetting than expected harmful events, even when the harm a victim suffered was the same.
-For instance, people judged that a robbery at a clothing store (unexpected) was more upsetting than a robbery at a convenience store (expected).
-Furthermore, these differences were large, systematic, and robust.
-Although there was variability in the situations that elicited the largest differences, our statistical approach treated trial as a random effect (see Methods and Results), allowing us to generalize our conclusions to items we did not test [@Judd2012].
-
-![Participants’ responses by item in Study 2.
-Upper: heatmap plot displaying probability of choosing each response on the 1 to 5 scale across items.
-Lower: marginal expected response across items, with 95% credible interval calculated from the ordinal regression model.
-Note the scale has been constrained to show relevant region of results.
-In both plots, a response of 3 indicates indifference, responses below 3 indicate a bias to choose the more expected event as more upsetting, and responses above 3 indicate a bias to choose the less expected event as more upsetting.
-Similar to Study 1, participants showed a robust bias toward choosing the less expected event as more upsetting.
-Trials are sorted by response to aid readability.
-<span data-label="fig:fig1"></span>](fig2.eps)
-
-We have argued that expectations shape our reactions to events because they determine how informative those events are in shaping our understanding of the world.
-Thus, we argue that well-grounded information theoretic principles can account for people’s divergent moral reactions to likely and unlikely events.
-
-In some contexts, it is clearly adaptive that the violation of our expectations would direct our attention to the events and situations about which our understanding is most in need of revision [@Rescorla1972].
-For instance, an act of terrorism in a peaceful nation might be less expected because of our beliefs about the effectiveness of security forces, the stability of national relationships, and positive relationships among people groups in that nation.
-Therefore, a surprising act of terrorism casts doubt on our deeper understanding of the geopolitical situation, and thereby leads us to suspect similar acts are more likely to occur in the future.
-
-Yet, however well-grounded the underlying principles, there is a clear moral danger whenever expectations influence our moral concerns.
-In particular, these effects pose a significant threat when considering our reactions to humanitarian challenges brought on by poverty or regional instability.
-In these cases, the influence of expectations on moral evaluations threatens to impose vicious and morally pernicious cycles.
-For instance, people living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and are the very people in greatest need of assistance and concern from the world at-large.
-However, for precisely these reasons, it is often unsurprising when harm befalls people living in these circumstances.
-Thus the influence of expectations on moral evaluations threatens to reduce observer's concern for these victims most in need, making people less likely to donate time or money to aid victims, or to take political action to prevent further harms.
-The present research identifies this potentially pernicious moral bias and grounds it in information theoretic principles.
-Hopefully, by further understanding these cognitive dynamics, psychologists and policymakers can begin to find ways to reduce or counteract people's tendency to ignore the plight of those in most need of attention.
-
-# Materials and Methods {#materials-and-methods .unnumbered}
-
-We conducted two preregistered studies examining the role of expectations in moral evaluations (https://osf.io/86rsw/).
-Each study consisted of multiple trials where participants were presented brief descriptions of two events.
-Their task was to decide which of the two events was more upsetting.
-In Study 1, they made a two-alternative forced-choice, and in Study 2 they made their judgments using a five-point rating scale.
-Both studies consisted of 16 experimental trials, five “equivalent” filler trials, and five “non-equivalent” filler trials.
-
-In experimental trials, the two events described identical harms, but one event was more expected and one more unexpected, as determined in a prior norming study (see SI).
-For example, participants considered the following stimulus:
-
-* “A 30 year old man in California dies in an earthquake” [Expected]
-
-* “A 30 year old man in Oklahoma dies in an earthquake” [Unexpected]
-
-In each of these events, the harm to the victim is the same (death) but one event is more probable than the other, given the different likelihoods of earthquakes occurring in California versus Oklahoma.
-In addition to being more naturalistic, manipulating expectations through (otherwise irrelevant) contextual details helps ensure that the harm suffered by the victim is understood to be identical.
-In the context of gambles, it is possible to manipulate probabilities specifically and directly [@Mellers1997].
-However, in the context of more realistic moral harms, any attempt to directly manipulate probabilities risks also changing participants’ perceptions of the degree of harm.
-For instance, suppose we describe two cancer patients, one with a 5% chance of survival and one with a 50% chance of survival.
-Even if they both ultimately meet the same fate, participants might fairly assume the patient with the lower chance of survival had a more serious illness, was more debilitated, suffered more severely, and so forth.
-In contrast, changing the context of the two events explains away these likelihood differences, better equating the events in other respects.
-
-Both studies also included “equivalent” filler trials.
-In these trials, the two events differed in trivial contextual details that we did not expect would affect participants’ judgments.
-These filler trials were meant to prevent participants from becoming explicitly aware of the structure of the experimental trials.
-Finally, both studies included “non-equivalent” filler trials, the two events differed substantially in the degree of harm suffered by a victim, so that one event was expected to be seen as considerably more upsetting than the other.
-These trials were included to allow participants a chance to use the extremes of the response scale and to reduce any task demands that might drive them to make artificially fine-grained distinctions when generating their responses.
-All experimental materials for these studies are available as Supporting Information at https://osf.io/86rsw/.
-
-## Exclusions {#exclusions .unnumbered}
-
-In both studies, we excluded participants who failed to correctly answer attention-check questions or who indicated at the end of the study they had not taken their participation seriously.
-
-## Data Analysis {#data-analysis .unnumbered}
-
-We analyzed our data by performing Bayesian estimation using the probabilistic programming language Stan [@Carpenter2017].
-We tested our preregistered predictions by computing Bayes Factors (i.e.
-$BF_{01}$) using the hypothesis function in the R package brms.
-Bayes Factors express the ratio of the probability of data under the null hypothesis to the probability of the data under an alternative hypothesis.
-Larger Bayes Factors indicate that the data are more likely under the null hypothesis (i.e., that the intercept is not different from zero) than the alternative hypothesis (i.e., that the intercept is different from zero), and vice versa.
-As Bayes Factors can be influenced by prior choices [@Gelman2017], we performed prior robustness checks to confirm that our conclusions would not depend substantially on the specification of the priors.
-
-To ensure the generality of our findings, we tested participants’ reactions to a variety of items and treated both participants and items as random effects throughout our analyses.
-The use of random effects over items allows us to consider these items as a sample from a larger population of possible items, and to generalize our findings to that wider population, much as we generalize from our sample of participants to the larger population [@Judd2012].
-
-## Study 1 {#study1 .unnumbered}
-
-### Participants {#s1-participants .unnumbered}
-
-A total of `r n_distinct(d_raw$workerId)` participants were recruited from Amazon’s Mechanical Turk work distribution website (mTurk).
-Of these, `r n_distinct(d$workerId)` passed attention checks and were included in the final analyses (`r gender_exp1$male` male, `r gender_exp1$female` female, median age = `r age_exp1` years old).
-All participants were paid $0.60 for their participation.
-
-### Materials and procedure {#s1-materials-procedure .unnumbered}
-
-Participants judged 16 experimental event-pairs and 10 equivalent filler event-pairs.
-On each trial (an example is shown above), participants were presented with the event-pair stimulus and had to judge which outcome was more upsetting in a two-alternative forced choice task.
-The two events were labeled “Outcome 1” and “Outcome 2” and the order, and whether Outcome 1 or 2 was the unexpected event, was randomized and counter-balanced.
-
-### Results and discussion {#s1-results .unnumbered}
-
-We predicted that people would judge that events where unexpected harm occurred were more upsetting than events where expected harm occurred.
-As indicated in Figure 1, we observed this trend.
-We estimated the magnitude of this effect by fitting a Bayesian logistic random effects model with participants’ responses as the dependent variable (1 = unexpected event is more upsetting; 0 = expected event is more upsetting) and random intercepts for item and subject, allowing us to generalize both to items and subjects we did not test.
-The intercept in this model represents the log-odds of selecting the unexpected event as being more upsetting.
-Thus, by examining the population-level intercept, we can test whether participants were biased toward selecting the unexpected event ($\beta$ > 0), the expected event ($\beta$ < 0), or were unbiased ($\beta$ = 0).
-We found that people were considerably more likely to think that unexpected events were more upsetting than events that were expected, Intercept = `r round(coefs_exp1$estimate, 3)`, 95% CI [`r round(coefs_exp1$lower, 3)`, `r round(coefs_exp1$upper, 3)`], $BF_{01}$ `r print_bf(bf_exp1)`.
-Figure 1 shows participants' responses broken-down by individual items. Participants' bias toward selecting the unexpected event as more upsetting was consistent across the 16 experimental items.
-
-## Study 2 {#study2 .unnumbered}
-
-In Study 1 we found that people’s judgments of events were biased by their expectations about those events.
-When forced to choose between two events, participants decided that unexpected events were more upsetting than expected events.
-In Study 2, we sought to test our hypothesis using a more conservative method: we provided participants with a more expressive response scale so that if they viewed the events under consideration as equally harmful, their responses could reflect their attitude.
-
-### Participants {#s2-participants .unnumbered}
-
-A total of `r n_distinct(d2_raw$workerId)` participants were recruited from Amazon’s Mechanical Turk work distribution website (mTurk).
-Of these, `r n_distinct(d2$workerId)` passed attention checks and were included in the final analyses (`r gender_exp2$male` male, `r gender_exp2$female` female, median age = `r age_exp2` years old).
-All participants were paid $0.60 for their participation.
-
-### Materials and procedure {#s2-materials-procedure .unnumbered}
-
-The materials in Study 2 were the same as those in Study 1.
-As in Study 1, on each trial of the study, participants were presented with a pair of actions labeled “Outcome 1” and “Outcome 2” and were asked, “Which outcome seems more upsetting?” In Study 2 participants made their rating on a five-point scale (Outcome 1 seems more upsetting, Outcome 1 seems a little more upsetting, neither seems more upsetting than the other, Outcome 2 seems a little more upsetting, Outcome 2 seems more upsetting).
-
-### Results and discussion {#s2-results .unnumbered}
-
-We predicted that participants would evaluate unexpected events as more upsetting than expected events.
-We estimated the magnitude of this effect by fitting a cumulative (ordinal) random effects model with participants’ scale responses as the dependent variable (1 to 5) and random intercepts for item and subject.
-This model produces four intercept coefficients, representing the cumulative log-odds of responses at each scale point or lower.
-For instance, the second coefficient represents the log-odds participants chose a 2 ("outcome 2 seems slightly more upsetting") or lower on the scale.
-Similarly, the third intercept coefficient represents the log-odds participants chose a 3 or lower on the scale.
-To examine the predicted effect, we compared the third intercept coefficient (representing the log-odds of choosing 4 or higher) to the inverse of the second intercept coefficient (thereby representing the log-odds *not* choosing a 3 or higher--i.e., choosing a 1 or 2), allowing us to test whether participants were more likely to choose the expected or unexpected event as being more upsetting in cases where they did not choose the neither option.
-This analysis indicated that people were more likely to think that events that were unexpected were more upsetting than events that were expected, $BF_{01}$ `r print_bf(bf_exp2)` (see supporting information for full model results).
-We observed this difference consistently across nearly every item (Figure 2).
-
-We note that the most common response was to view both events, which involve identical harms, as equivalently upsetting.
-However, across almost every item, participants showed a clear bias towards choosing unexpected events as more upsetting.
-Thus, Study 2 replicates the tendency to view unexpected moral events as more upsetting using a more conservative judgment task.
-
-# Theoretical Formalization {#theory .unnumbered}
+# Theoretical Formalization
 
@@
 [^1]: It should be recognized that the result we obtain here bears a similarity to the “disappointment function” included in Mellers and colleagues’ [-@Mellers1997] Decision Affect Theory.
-
-<!-- Leave these lines as they are at the end of your .Rmd file to ensure placement of methods & acknowledgements sections before the references-->
-\showmatmethods
-\showacknow
-\pnasbreak

### 2018-09-04 paper/paper-apa.Rmd
paper in apa format

--- 2018-07-18 paper/paper.Rmd
@@
 ---
-title: Expectations bias moral evaluations
+title             : "Expectations bias moral evaluations"
+shorttitle        : "Expectations bias moral evaluations"
 
-# Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
 author: 
-  - name: Derek Powell
-    affiliation: a,1
-  - name: Zachary Horne
-    affiliation: b
-address:
-  - code: a
-    address: Stanford University, Department of Psychology, 450 Serra Mall, Stanford, CA, 94305
-  - code: b
-    address: Arizona State University, School of Social and Behavioral Sciences, 13591 N 47th Ave, Phoenix, AZ 85051
+  - name          : "Derek Powell"
+    affiliation   : "1"
+    corresponding : yes    # Define only one corresponding author
+    address       : "Stanford University, Department of Psychology, 450 Serra Mall, Stanford, CA, 94305"
+    email         : "derekpowell@stanford.edu"
+  - name          : "Zachary Horne"
+    affiliation   : "2"
 
-corresponding_author:
-  - code: 1
-    text: "To whom correspondence should be addressed. E-mail: derekpowell@stanford.edu"
+affiliation:
+  - id            : "1"
+    institution   : "Stanford University"
+  - id            : "2"
+    institution   : "Arizona State University"
 
-# For footer text
-lead_author_surname: Powell
+authornote: |
+  Derek Powell, Department of Psychology, Stanford University; Zachary Horne, School of Social and Behavioral Sciences, Arizona State University.
 
-## Remove this if not required - talk to zach 
-# equal_authors:
-#   - code: 0
-#     text:
+  The authors would like to acknowledge the help of Larisa Hussak, Keith Holyoak, Alan Fiske, Hongjing Lu, John Hummel, Andrei Cimpian, and Ellen Markman for their comments and support.
 
-author_contributions: |
   Both authors jointly conceived of the project and theory, designed and conducted experiments, and analyzed results. D.P. developed the formal aspects of the theory.
-
-## Remove this if not required
-conflict_of_interest: |
-  The authors declare no conflicts of interest.
 
@@
 
-significance: |
-  In two behavioral studies, we found that people's expectations biased their reactions to morally-significant events--people were more upset by surprising harms than more expected harms.
-This bias threatens to impose a pernicious cycle: People living in poverty, unstable regions, and other dangerous circumstances have the greatest need for assistance and concern from the world at-large.
-Yet for this very reason, it is often unsurprising when they are harmed.
-By blunting observers' concern for the disadvantaged, this bias could make people less likely to donate money, to volunteer, or to take political action, thereby increasing inequality and compounding suffering.
-Recognizing and understanding this bias will suggest ways to mitigate its harmful consequences.
+  <!-- https://tinyurl.com/ybremelq -->
 
-acknowledgements: |
-  We would like to acknowledge the help of Larisa Hussak, Keith Holyoak, Alan Fiske, Hongjing Lu, John Hummel, Andrei Cimpian, and Ellen Markman for their comments and support.
+keywords          : "Evaluation; Moral judgment; Information theory"
+wordcount         : "X"
 
-keywords:
-  - evaluation
-  - moral judgment
-  - information theory
-  # - optional
-  # - optional
+bibliography      : ["wiw-biblio.bib"]
 
-## must be one of: pnasresearcharticle (usual two-column layout), pnasmathematics (one column layout), or pnasinvited (invited submissions only)
-pnas_type: pnasresearcharticle
+floatsintext      : yes
+figurelist        : no
+tablelist         : no
+footnotelist      : no
+linenumbers       : no
+mask              : no
+draft             : no
 
-bibliography: wiw-biblio.bib
-csl: pnas.csl
-
-## change to true to add optional line numbering
-lineno: FALSE
-
-output: rticles::pnas_article
+documentclass     : "apa6"
+classoption       : "man"
+output:
+  papaja::apa6_pdf:
+    includes:
+      after_body: "appendix.tex"
 ---
 
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+```{r setup, include = FALSE}
+library("papaja")
+```
+
+```{r analysis-preferences}
+# Seed for random number generation
+set.seed(42)
+knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
 ```
@@
   iter = 2000,
-  cached_file = "fit_exp1.rds"
+  file = "brms_models/fit_exp1.rds"
   )
@@
   iter = 2000,
-  cached_file = "fit_exp2.rds"
+  file = "brms_models/fit_exp2.rds"
   )
@@
   iter = 2000,
-  cached_file = "fit_exp1rand.rds"
+  file = "brms_models/fit_exp1rand.rds"
   )
@@
   iter = 2000,
-  cached_file = "fit_exp2rand.rds"
+  file = "brms_models/fit_exp2rand.rds"
   )
@@
 
+# Study Overview
+
+We conducted two preregistered studies examining the role of expectations in moral evaluations (https://osf.io/86rsw/).
+Each study consisted of multiple trials where participants were presented brief descriptions of two events.
+Their task was to decide which of the two events was more upsetting.
+In Study 1, they made a two-alternative forced-choice, and in Study 2 they made their judgments using a five-point rating scale.
+Both studies consisted of 16 experimental trials, five “equivalent” filler trials, and five “non-equivalent” filler trials.
+
 The design of these studies was chosen with special consideration to two theoretical points.
@@
 This both allowed us to test the role of expectations based on world knowledge and provided a more naturalistic set of items.
-
-![Probability with which participants chose the less expected event as more upsetting across trials in Study 1.
-As predicted, participants exhibited a robust bias toward choosing the less expected event as more upsetting across all items.
-Error bars represent 95% credible intervals calculated from the hierarchical logistic regression model.
-See Supporting Information for full description of items.
-<span data-label="fig:fig1"></span>](fig1.eps)
-
-Consistent with our predictions, manipulating expectations shaped evaluations across a range of morally-significant events.
-We found that people viewed unexpected harmful events as more upsetting than expected harmful events, even when the harm a victim suffered was the same.
-For instance, people judged that a robbery at a clothing store (unexpected) was more upsetting than a robbery at a convenience store (expected).
-Furthermore, these differences were large, systematic, and robust.
-Although there was variability in the situations that elicited the largest differences, our statistical approach treated trial as a random effect (see Methods and Results), allowing us to generalize our conclusions to items we did not test [@Judd2012].
-
-![Participants’ responses by item in Study 2.
-Upper: heatmap plot displaying probability of choosing each response on the 1 to 5 scale across items.
-Lower: marginal expected response across items, with 95% credible interval calculated from the ordinal regression model.
-Note the scale has been constrained to show relevant region of results.
-In both plots, a response of 3 indicates indifference, responses below 3 indicate a bias to choose the more expected event as more upsetting, and responses above 3 indicate a bias to choose the less expected event as more upsetting.
-Similar to Study 1, participants showed a robust bias toward choosing the less expected event as more upsetting.
-Trials are sorted by response to aid readability.
-<span data-label="fig:fig1"></span>](fig2.eps)
-
-We have argued that expectations shape our reactions to events because they determine how informative those events are in shaping our understanding of the world.
-Thus, we argue that well-grounded information theoretic principles can account for people’s divergent moral reactions to likely and unlikely events.
-
-In some contexts, it is clearly adaptive that the violation of our expectations would direct our attention to the events and situations about which our understanding is most in need of revision [@Rescorla1972].
-For instance, an act of terrorism in a peaceful nation might be less expected because of our beliefs about the effectiveness of security forces, the stability of national relationships, and positive relationships among people groups in that nation.
-Therefore, a surprising act of terrorism casts doubt on our deeper understanding of the geopolitical situation, and thereby leads us to suspect similar acts are more likely to occur in the future.
-
-Yet, however well-grounded the underlying principles, there is a clear moral danger whenever expectations influence our moral concerns.
-In particular, these effects pose a significant threat when considering our reactions to humanitarian challenges brought on by poverty or regional instability.
-In these cases, the influence of expectations on moral evaluations threatens to impose vicious and morally pernicious cycles.
-For instance, people living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and are the very people in greatest need of assistance and concern from the world at-large.
-However, for precisely these reasons, it is often unsurprising when harm befalls people living in these circumstances.
-Thus the influence of expectations on moral evaluations threatens to reduce observer's concern for these victims most in need, making people less likely to donate time or money to aid victims, or to take political action to prevent further harms.
-The present research identifies this potentially pernicious moral bias and grounds it in information theoretic principles.
-Hopefully, by further understanding these cognitive dynamics, psychologists and policymakers can begin to find ways to reduce or counteract people's tendency to ignore the plight of those in most need of attention.
-
-# Materials and Methods {#materials-and-methods .unnumbered}
-
-We conducted two preregistered studies examining the role of expectations in moral evaluations (https://osf.io/86rsw/).
-Each study consisted of multiple trials where participants were presented brief descriptions of two events.
-Their task was to decide which of the two events was more upsetting.
-In Study 1, they made a two-alternative forced-choice, and in Study 2 they made their judgments using a five-point rating scale.
-Both studies consisted of 16 experimental trials, five “equivalent” filler trials, and five “non-equivalent” filler trials.
 
@@
 
-## Exclusions {#exclusions .unnumbered}
+## Exclusions
 
@@
 
-## Data Analysis {#data-analysis .unnumbered}
+## Data Analysis
 
@@
 
-## Study 1 {#study1 .unnumbered}
+## Study 1
 
-### Participants {#s1-participants .unnumbered}
+### Participants
 
@@
 
-### Materials and procedure {#s1-materials-procedure .unnumbered}
+### Materials and procedure
 
@@
 
-### Results and discussion {#s1-results .unnumbered}
+### Results and discussion
 
@@
 
-## Study 2 {#study2 .unnumbered}
+(ref:fig1-caption)
+Probability with which participants chose the less expected event as more upsetting across trials in Study 1.
+As predicted, participants exhibited a robust bias toward choosing the less expected event as more upsetting across all items.
+Error bars represent 95% credible intervals calculated from the hierarchical logistic regression model.
+See Supporting Information for full description of items.
+
+```{r figure1, fig.cap="(ref:fig1-caption)"}
+knitr::include_graphics("fig1.eps") 
+```
+
+## Study 2
 
@@
 
-### Participants {#s2-participants .unnumbered}
+### Participants
 
@@
 
-### Results and discussion {#s2-results .unnumbered}
+(ref:fig2-caption)
+Participants’ responses by item in Study 2.
+Upper: heatmap plot displaying probability of choosing each response on the 1 to 5 scale across items.
+Lower: marginal expected response across items, with 95% credible interval calculated from the ordinal regression model.
+Note the scale has been constrained to show relevant region of results.
+In both plots, a response of 3 indicates indifference, responses below 3 indicate a bias to choose the more expected event as more upsetting, and responses above 3 indicate a bias to choose the less expected event as more upsetting.
+Similar to Study 1, participants showed a robust bias toward choosing the less expected event as more upsetting.
+Trials are sorted by response to aid readability.
+
+```{r figure2, fig.cap="(ref:fig2-caption)", fig.align = "center", out.width = "\\textwidth", fig.pos = "!h"}
+# plot(arrangeGrob(g))
+knitr::include_graphics("fig2.eps") 
+```
+
+### Results and discussion
 
@@
 
-# Theoretical Formalization {#theory .unnumbered}
+### Discussion
 
-We construe event evaluation as a comparison between the state of affairs prior to an event and the state of affairs following the event.
-We represent the state of the world as a random variable, $X$.
-We assume $X$ is discrete, and that every realization $x$ represents some potential specific state of the world in all relevant aspects under consideration.
-An agent has some utility function $u$, that applies over the different realizations of $X$.
-Thus the expected utility of the present state of the world is:
+Consistent with our predictions, manipulating expectations shaped evaluations across a range of morally-significant events.
+We found that people viewed unexpected harmful events as more upsetting than expected harmful events, even when the harm a victim suffered was the same.
+For instance, people judged that a robbery at a clothing store (unexpected) was more upsetting than a robbery at a convenience store (expected).
+Furthermore, these differences were large, systematic, and robust.
+Although there was variability in the situations that elicited the largest differences, our statistical approach treated trial as a random effect (see Methods and Results), allowing us to generalize our conclusions to items we did not test [@Judd2012].
 
-\begin{equation*}
-E_X[u(X)] = \sum_x u(x)p(x)
-\end{equation*}
+We have argued that expectations shape our reactions to events because they determine how informative those events are in shaping our understanding of the world.
+Thus, we argue that well-grounded information theoretic principles can account for people’s divergent moral reactions to likely and unlikely events.
 
-When we learn an event has occurred, we can represent this as an observation of another random variable, $S$.
-We can think of $S$ as a "signalling event" because it informs (signals) to us something about $X$ [this formalism follows @Arrow1996].
-For simplicity, we treat $S$ as binary, either 1 or 0.
-Importantly, we assume there is a dependency between $S$ and $X$, so that observing $S$ informs our knowledge of $X$.
-For instance, reading a news report of a terror attack in Brussels informs us about the state of the world, telling us about what realizations of X are no longer possible (e.g., any wherein it was a peaceful day in Brussels) and which are more or less likely.
-Conditioned on $S$, the expected utility of the world is:
+In some contexts, it is clearly adaptive that the violation of our expectations would direct our attention to the events and situations about which our understanding is most in need of revision [@Rescorla1972].
+For instance, an act of terrorism in a peaceful nation might be less expected because of our beliefs about the effectiveness of security forces, the stability of national relationships, and positive relationships among people groups in that nation.
+Therefore, a surprising act of terrorism casts doubt on our deeper understanding of the geopolitical situation, and thereby leads us to suspect similar acts are more likely to occur in the future.
 
-\begin{equation*}
-E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
-\end{equation*}
+Yet, however well-grounded the underlying principles, there is a clear moral danger whenever expectations influence our moral concerns.
+In particular, these effects pose a significant threat when considering our reactions to humanitarian challenges brought on by poverty or regional instability.
+In these cases, the influence of expectations on moral evaluations threatens to impose vicious and morally pernicious cycles.
+For instance, people living in geo-politically unstable regions or in the developing world are often those who are most affected by terrorism, famine, and natural disasters, and are the very people in greatest need of assistance and concern from the world at-large.
+However, for precisely these reasons, it is often unsurprising when harm befalls people living in these circumstances.
+Thus the influence of expectations on moral evaluations threatens to reduce observer's concern for these victims most in need, making people less likely to donate time or money to aid victims, or to take political action to prevent further harms.
+The present research identifies this potentially pernicious moral bias and grounds it in information theoretic principles.
+Hopefully, by further understanding these cognitive dynamics, psychologists and policymakers can begin to find ways to reduce or counteract people's tendency to ignore the plight of those in most need of attention.
 
-We define the value assigned to the signalling event, $V(S)$, as:
+```{r render_appendix, include=FALSE}
+render_appendix("appendix.Rmd")
+```
 
-\begin{equation}
-V(S) = E_{X|S}[u(X)] - E_X[u(X)]
-\end{equation}
+\newpage
 
-Which we write more specifically as:
+# References
+```{r create_r-references}
+r_refs(file = "r-references.bib")
+```
 
-\begin{equation}
-V(s=1) = \sum_x u(x)p(x|s=1) - \sum_x u(x)p(x)
-\end{equation}
+\begingroup
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
 
-Our goal is to express this relation in terms of $p(S)$. First, we manipulate $E_X[u(X)]$ by applying the expectation of conditional expectation and some algebra.
-
-\begin{align*}
-\sum_x u(x)p(x) & = \sum_s u(x)p(x|s)p(s) \\
- &= \begin{aligned}[t]
-    & \sum_x u(x)p(x|s=1)p(s=1) - \\
-    & \quad \sum_x u(x)p(x|s=0)p(s=0) \end{aligned}\\
- &= \begin{aligned}[t]
-    & p(s=1)\sum_x u(x)p(x|s=1) -\\
-    & \quad p(s=0) \sum_x u(x)p(x|s=0) \end{aligned}\\
- &= \begin{aligned}[t]
-    & p(s=1)\sum_x u(x)p(x|s=1) - \\
-    & \quad (1-p(s=1)) \sum_x u(x)p(x|s=0) \end{aligned}\\
-\end{align*}
-
-Substituting this quantity back into Equation 2, we obtain:
-
-\begin{align*}
-V(s=1) = & \sum_x u(x)p(x|s=1) - \\ & p(s=1)\sum_x u(x)p(x|s=1) -\\ & (1-p(s=1)) \sum_x u(x)p(x|s=0)
-\end{align*}
-
-Finally, we factor the first two sums and combine them with the last sum to yield:
-
-\begin{equation}
-V(s=1)=(1-p(s=1))\sum_x u(x)p(x|s=1) - u(x)p(x|s=0)
-\end{equation}
-
-Under this formalization, the role of expectations in the evaluation of events does not depend on any further assumptions about the states of the world or the utility function to be applied over them.
-Instead, this conclusion depends on just two assumptions: first, that events are evaluated by comparing the utilities of the states of the world prior to and following the event; and second, that knowledge of those states of the world is uncertain.
-
-We have appealed to the informativeness of an event to illustrate intuitively how expectations should influence evaluations.
-More expected events carry less information, and thereby engender weaker reactions.
-Though our model of event evaluation is not an information-theoretic model in a direct sense, this interpretation still applies: the information carried by observing $S$ is a function of that observation’s prior probability, which we have in turn shown determines its evaluation.
-
-[^1]: It should be recognized that the result we obtain here bears a similarity to the “disappointment function” included in Mellers and colleagues’ [-@Mellers1997] Decision Affect Theory.
-
-<!-- Leave these lines as they are at the end of your .Rmd file to ensure placement of methods & acknowledgements sections before the references-->
-\showmatmethods
-\showacknow
-\pnasbreak
+<div id = "refs"></div>
+\endgroup

### 2018-09-04 paper/paper-apa.Rmd
added R version and cites

@@
 
-bibliography      : ["wiw-biblio.bib"]
+bibliography      : ["wiw-biblio.bib", "r-references.bib"]
 
@@
 
+```{r}
+my_citations <- cite_r(
+  file = "r-references.bib"
+  , pkgs = c("tidyverse", "brms", "gridExtra", "papaja")
+  , withhold = FALSE
+  # , footnote = TRUE
+)
+```
+
 When we learn of tragic events--a friend's loss of a loved-one, a violent killing, or a terrorist attack--we are mournful, concerned, angry, and driven to action.
@@
 
+All analyses were carried out in `r my_citations`.
 We analyzed our data by performing Bayesian estimation using the probabilistic programming language Stan [@Carpenter2017].
 We tested our preregistered predictions by computing Bayes Factors (i.e.
-$BF_{01}$) using the hypothesis function in the R package brms.
+$BF_{01}$) using the *hypothesis* function in the R-package *brms* [Version 2.2.0; @R-brms].
 Bayes Factors express the ratio of the probability of data under the null hypothesis to the probability of the data under an alternative hypothesis.

### 2018-09-06 paper/paper-apa.Rmd
added reg table

@@
 library(tidyverse)
+library(broom)
+library(brms)
 
@@
 ```{r fit brms models, include=FALSE}
-library(brms)
+
 devtools::source_gist(id = "f1994c0f8325abbc5d300600744af39d", filename="cbrm.R")
@@
 
-coefs_exp1 <- broom::tidy(fit_exp1rand) %>%
+coefs_exp1 <- tidy(fit_exp1rand) %>%
   filter(term=="b_intercept")
@@
   file = "r-references.bib"
-  , pkgs = c("tidyverse", "brms", "gridExtra", "papaja")
+  , pkgs = c("tidyverse", "brms", "broom", "gridExtra", "papaja")
   , withhold = FALSE
@@
 We found that people were considerably more likely to think that unexpected events were more upsetting than events that were expected, Intercept = `r round(coefs_exp1$estimate, 3)`, 95% CI [`r round(coefs_exp1$lower, 3)`, `r round(coefs_exp1$upper, 3)`], $BF_{01}$ `r print_bf(bf_exp1)`.
-Figure 1 shows participants' responses broken-down by individual items. Participants' bias toward selecting the unexpected event as more upsetting was consistent across the 16 experimental items.
+Figure 1 shows participants' responses broken-down by individual items.
+Participants' bias toward selecting the unexpected event as more upsetting was consistent across the 16 experimental items.
 
@@
 We estimated the magnitude of this effect by fitting a cumulative (ordinal) random effects model with participants’ scale responses as the dependent variable (1 to 5) and random intercepts for item and subject.
-This model produces four intercept coefficients, representing the cumulative log-odds of responses at each scale point or lower.
+This model produces four intercept coefficients, representing the cumulative log-odds of responses at each scale point or lower (summarized in Table 1).
 For instance, the second coefficient represents the log-odds participants chose a 2 ("outcome 2 seems slightly more upsetting") or lower on the scale.
@@
 To examine the predicted effect, we compared the third intercept coefficient (representing the log-odds of choosing 4 or higher) to the inverse of the second intercept coefficient (thereby representing the log-odds *not* choosing a 3 or higher--i.e., choosing a 1 or 2), allowing us to test whether participants were more likely to choose the expected or unexpected event as being more upsetting in cases where they did not choose the neither option.
-This analysis indicated that people were more likely to think that events that were unexpected were more upsetting than events that were expected, $BF_{01}$ `r print_bf(bf_exp2)` (see supporting information for full model results).
+This analysis indicated that people were more likely to think that events that were unexpected were more upsetting than events that were expected, $BF_{01}$ `r print_bf(bf_exp2)`.
 We observed this difference consistently across nearly every item (Figure 2).
@@
 Thus, Study 2 replicates the tendency to view unexpected moral events as more upsetting using a more conservative judgment task.
+
+```{r, results="asis"}
+tidy(fit_exp2rand) %>%
+  filter(grepl("b_Intercept", term), !grepl("prior", term)) %>%
+  mutate(
+    Term = gsub("b_","",term),
+    `95\\% conf. interval` = paste0("[",round(lower,3), ", ", round(upper,3), "]"),
+    B = estimate,
+    `SE B` = std.error,
+    `$e^B$` = exp(B)
+    ) %>%
+  select(Term, B, `SE B`, `95\\% conf. interval`) %>%
+  apa_table(caption="Bayesian regression results for Experiment 2.", 
+            escape=FALSE,
+            align = c("r","c","c","c"))
+```
 

### 2018-09-06 SuppInfo.Rmd
updated readme
paper edits pre submission

@@
 ---
-title: "Supporting Information"
+title: "Expectations bias moral evaluations: Supporting information"
 output:
   pdf_document:
-    keep_tex: true
+bibliography: "paper/wiw-biblio.bib"
+csl: apa6.csl
+    # keep_tex: true
   # html_notebook: default
@@
 ```
+
+# Theoretical Formalization
+
+We construe event evaluation as a comparison between the state of affairs prior to an event and the state of affairs following the event.
+We represent the state of the world as a random variable, $X$.
+We assume $X$ is discrete, and that every realization $x$ represents some potential specific state of the world in all relevant aspects under consideration.
+An agent has some utility function $u$, that applies over the different realizations of $X$.
+Thus the expected utility of the present state of the world is:
+
+\begin{equation*}
+E_X[u(X)] = \sum_x u(x)p(x)
+\end{equation*}
+
+When we learn an event has occurred, we can represent this as an observation of another random variable, $S$.
+We can think of $S$ as a "signalling event" because it informs (signals) to us something about $X$ [this formalism follows @Arrow1996].
+For simplicity, we treat $S$ as binary, either 1 or 0.
+Importantly, we assume there is a dependency between $S$ and $X$, so that observing $S$ informs our knowledge of $X$.
+For instance, reading a news report of a terror attack in Brussels informs us about the state of the world, telling us about what realizations of X are no longer possible (e.g., any wherein it was a peaceful day in Brussels) and which are more or less likely.
+Conditioned on $S$, the expected utility of the world is:
+
+\begin{equation*}
+E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
+\end{equation*}
+
+We define the value assigned to the signalling event, $V(S)$, as:
+
+\begin{equation}
+V(S) = E_{X|S}[u(X)] - E_X[u(X)]
+\end{equation}
+
+Which we write more specifically as:
+
+\begin{equation}
+V(s=1) = \sum_x u(x)p(x|s=1) - \sum_x u(x)p(x)
+\end{equation}
+
+Our goal is to express this relation in terms of $p(S)$. First, we manipulate $E_X[u(X)]$ by applying the expectation of conditional expectation and some algebra.
+
+\begin{align*}
+\sum_x u(x)p(x) & = \sum_s u(x)p(x|s)p(s) \\
+ &= \begin{aligned}[t]
+    & \sum_x u(x)p(x|s=1)p(s=1) - \\
+    & \quad \sum_x u(x)p(x|s=0)p(s=0) \end{aligned}\\
+ &= \begin{aligned}[t]
+    & p(s=1)\sum_x u(x)p(x|s=1) -\\
+    & \quad p(s=0) \sum_x u(x)p(x|s=0) \end{aligned}\\
+ &= \begin{aligned}[t]
+    & p(s=1)\sum_x u(x)p(x|s=1) - \\
+    & \quad (1-p(s=1)) \sum_x u(x)p(x|s=0) \end{aligned}\\
+\end{align*}
+
+Substituting this quantity back into Equation 2, we obtain:
+
+\begin{align*}
+V(s=1) = & \sum_x u(x)p(x|s=1) - \\ & p(s=1)\sum_x u(x)p(x|s=1) -\\ & (1-p(s=1)) \sum_x u(x)p(x|s=0)
+\end{align*}
+
+Finally, we factor the first two sums and combine them with the last sum to yield:
+
+\begin{equation}
+V(s=1)=(1-p(s=1))\sum_x u(x)p(x|s=1) - u(x)p(x|s=0)
+\end{equation}
+
+Under this formalization, the role of expectations in the evaluation of events does not depend on any further assumptions about the states of the world or the utility function to be applied over them.
+Instead, this conclusion depends on just two assumptions: first, that events are evaluated by comparing the utilities of the states of the world prior to and following the event; and second, that knowledge of those states of the world is uncertain.
+
+We have appealed to the informativeness of an event to illustrate intuitively how expectations should influence evaluations.
+More expected events carry less information, and thereby engender weaker reactions.
+Though our model of event evaluation is not an information-theoretic model in a direct sense, this interpretation still applies: the information carried by observing $S$ is a function of that observation’s prior probability, which we have in turn shown determines its evaluation.
+
+\newpage
 
@@
 | brokeninto | Three cars on the same street are broken into while parked in the suburbs. | Three cars on the same street are broken into while parked downtown. |
+
+\newpage
+
+# References
+
+\begingroup
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
+\noindent

### 2018-09-06 paper/paper-apa.Rmd
paper edits pre submission

@@
 classoption       : "man"
-output:
-  papaja::apa6_pdf:
-    includes:
-      after_body: "appendix.tex"
+output: papaja::apa6_pdf
+    # includes:
+    #   after_body: "appendix.tex"
 ---
@@
 
-From this it can be shown that the value assigned to event $S$ is proportional to its prior probability (see Theoretical Formalization).
+From this it can be shown that the value assigned to event $S$ is proportional to its prior probability (see Theoretical Formalization section in supplemental materials).
 Specifically, it is the difference in the expected utility under different values of $S$, weighted by the inverse of the prior probability of $S$.
@@
 
-# Study Overview
+# General Methods
 
@@
 
-A total of `r n_distinct(d_raw$workerId)` participants were recruited from Amazon’s Mechanical Turk work distribution website (mTurk).
+In light of meta-meta analyses of effect sizes in social psychology [@Richard2003], we sought to achieve 80% power to detect an effect of cohen's D of .2, requiring approximately 200 participants.
+Anticipating losing some participants to attention checks or other forms of attrition, a total of `r n_distinct(d_raw$workerId)` participants were recruited from Amazon’s Mechanical Turk work distribution website (mTurk).
 Of these, `r n_distinct(d$workerId)` passed attention checks and were included in the final analyses (`r gender_exp1$male` male, `r gender_exp1$female` female, median age = `r age_exp1` years old).
@@
 
+Due to the nature of this task, we anticipated the effects in this study might be smaller than in Study 1.
+Consequently, we approximately doubled our sample size to ensure adequate power to detect any effect of interest.
 A total of `r n_distinct(d2_raw$workerId)` participants were recruited from Amazon’s Mechanical Turk work distribution website (mTurk).
@@
 
-### Discussion
+# Discussion
 
@@
 ```{r render_appendix, include=FALSE}
-render_appendix("appendix.Rmd")
+# render_appendix("appendix.Rmd")
 ```

### 2018-09-13 paper/appendix.Rmd
preprint

@@
 
-We construe event evaluation as a comparison between the state of affairs prior to an event and the state of affairs following the event.
-We represent the state of the world as a random variable, $X$.
-We assume $X$ is discrete, and that every realization $x$ represents some potential specific state of the world in all relevant aspects under consideration.
-An agent has some utility function $u$, that applies over the different realizations of $X$.
-Thus the expected utility of the present state of the world is:
-
-\begin{equation*}
-E_X[u(X)] = \sum_x u(x)p(x)
-\end{equation*}
-
-When we learn an event has occurred, we can represent this as an observation of another random variable, $S$.
-We can think of $S$ as a "signalling event" because it informs (signals) to us something about $X$ [this formalism follows @Arrow1996].
-For simplicity, we treat $S$ as binary, either 1 or 0.
-Importantly, we assume there is a dependency between $S$ and $X$, so that observing $S$ informs our knowledge of $X$.
-For instance, reading a news report of a terror attack in Brussels informs us about the state of the world, telling us about what realizations of X are no longer possible (e.g., any wherein it was a peaceful day in Brussels) and which are more or less likely.
-Conditioned on $S$, the expected utility of the world is:
-
-\begin{equation*}
-E_{X|S}[u(X)] = \sum_x u(x)p(x|s)
-\end{equation*}
-
-We define the value assigned to the signalling event, $V(S)$, as:
+We defined the value assigned to the signalling event, $V(S)$, as:
 
@@
 \end{equation}
-
-Under this formalization, the role of expectations in the evaluation of events does not depend on any further assumptions about the states of the world or the utility function to be applied over them.
-Instead, this conclusion depends on just two assumptions: first, that events are evaluated by comparing the utilities of the states of the world prior to and following the event; and second, that knowledge of those states of the world is uncertain.
-
-We have appealed to the informativeness of an event to illustrate intuitively how expectations should influence evaluations.
-More expected events carry less information, and thereby engender weaker reactions.
-Though our model of event evaluation is not an information-theoretic model in a direct sense, this interpretation still applies: the information carried by observing $S$ is a function of that observation’s prior probability, which we have in turn shown determines its evaluation.
-
-[^1]: It should be recognized that the result we obtain here bears a similarity to the “disappointment function” included in Mellers and colleagues’ [-@Mellers1997] Decision Affect Theory.

### 2018-09-13 paper/paper-apa.Rmd
preprint

--- 2018-09-06 paper/paper-apa.Rmd
@@
 abstract: |
-  When we learn of tragic events--a violent killing or a terrorist attack--we are mournful and often driven to action.
-But responses to these events can differ substantially depending on the context in which they occur.
-When tragedy takes us by surprise, we are shocked and outraged, but when it strikes in unstable regions or rough neighborhoods, we are often resigned to feel that “these things happen.” Here we explain the role that expectations seem to play in moral evaluation by modeling evaluations as the comparison of the state of the world before and after an event.
-According to this theory, expectations influence evaluations through information-theoretic principles: less expected events do more to inform us about the state of the world than do more expected events.
-Consistent with this theory, in two preregistered studies we found that people’s judgments of morally-significant events were affected by the prior probability of that event.
-People were more upset about an event that was unexpected (e.g., a robbery at a clothing store) than an event that was more more probable (e.g., a robbery at a convenience store).
-This bias threatens to impose a pernicious cycle: People living in poverty, geo-politically unstable regions, and other dangerous circumstances have the greatest need for assistance and concern.
-However, for these very reasons, it is often unsurprising when they are harmed.
-Consequently, the influence of expectations on moral evaluations threatens to reduce observers’ concern for those most in need.
+  People’s expectations play an important role in their reactions to events.
+There is often disappointment when events fail to meet expectations and a special thrill to having one’s expectations exceeded.
+We propose that expectations influence evaluations through information-theoretic principles: less expected events do more to inform us about the state of the world than do more expected events.
+An implication of this proposal is that people may have inappropriately muted responses to morally significant but expected events.
+In two preregistered experiments, we found that people’s judgments of morally-significant events were affected by the likelihood of that event.
+People were more upset about events that were unexpected (e.g., a robbery at a clothing store) than events  that were more expected (e.g., a robbery at a convenience store).
+We argue that this bias has pernicious moral consequences, including leading to reduced concern for victims in most need of help.
 
@@
 keywords          : "Evaluation; Moral judgment; Information theory"
-wordcount         : "X"
+wordcount         : "1911"
 
@@
 
-A total of `r n_distinct(d_raw$workerId)` participants were recruited from Amazon’s Mechanical Turk work distribution website (mTurk).
+In light of meta-meta analyses of effect sizes in social psychology [@Richard2003], we sought to achieve 80% power to detect an effect of cohen's _d_ of .2, requiring approximately 200 participants.
+As we anticipated excluding participants who failed attention checks, a total of `r n_distinct(d_raw$workerId)` participants were recruited from Amazon’s Mechanical Turk work distribution website (mTurk).
 Of these, `r n_distinct(d$workerId)` passed attention checks and were included in the final analyses (`r gender_exp1$male` male, `r gender_exp1$female` female, median age = `r age_exp1` years old).
@@
 On each trial (an example is shown above), participants were presented with the event-pair stimulus and had to judge which outcome was more upsetting in a two-alternative forced choice task.
-The two events were labeled “Outcome 1” and “Outcome 2” and the order, and whether Outcome 1 or 2 was the unexpected event, was randomized and counter-balanced.
+The two events were labeled “Outcome 1” and “Outcome 2” and the order, and whether Outcome 1 or 2 was the unexpected event, was randomized and counterbalanced.
 
@@
 
-We predicted that participants would evaluate unexpected events as more upsetting than expected events.
-We estimated the magnitude of this effect by fitting a cumulative (ordinal) random effects model with participants’ scale responses as the dependent variable (1 to 5) and random intercepts for item and subject.
-This model produces four intercept coefficients, representing the cumulative log-odds of responses at each scale point or lower (summarized in Table 1).
-For instance, the second coefficient represents the log-odds participants chose a 2 ("outcome 2 seems slightly more upsetting") or lower on the scale.
-Similarly, the third intercept coefficient represents the log-odds participants chose a 3 or lower on the scale.
-To examine the predicted effect, we compared the third intercept coefficient (representing the log-odds of choosing 4 or higher) to the inverse of the second intercept coefficient (thereby representing the log-odds *not* choosing a 3 or higher--i.e., choosing a 1 or 2), allowing us to test whether participants were more likely to choose the expected or unexpected event as being more upsetting in cases where they did not choose the neither option.
-This analysis indicated that people were more likely to think that events that were unexpected were more upsetting than events that were expected, $BF_{01}$ `r print_bf(bf_exp2)`.
-We observed this difference consistently across nearly every item (Figure 2).
-
-We note that the most common response was to view both events, which involve identical harms, as equivalently upsetting.
-However, across almost every item, participants showed a clear bias towards choosing unexpected events as more upsetting.
-Thus, Study 2 replicates the tendency to view unexpected moral events as more upsetting using a more conservative judgment task.
-
 ```{r, results="asis"}
@@
 
-### Discussion
+We predicted that participants would evaluate unexpected events as more upsetting than expected events.
+We estimated the magnitude of this effect by fitting a cumulative (ordinal) random effects model with participants’ scale responses as the dependent variable (1 to 5) and random intercepts for item and subject.
+This model produces four intercept coefficients, representing the cumulative log-odds of responses at each scale point or lower (summarized in Table 1).
+For instance, the second coefficient represents the log-odds participants chose a 2 ("outcome 2 seems slightly more upsetting") or lower on the scale.
+Similarly, the third intercept coefficient represents the log-odds participants chose a 3 or lower on the scale.
+To examine the predicted effect, we compared the third intercept coefficient (representing the log-odds of choosing 4 or higher) to the inverse of the second intercept coefficient (thereby representing the log-odds *not* choosing a 3 or higher--i.e., choosing a 1 or 2), allowing us to test whether participants were more likely to choose the expected or unexpected event as being more upsetting in cases where they did not choose the neither option.
+This analysis indicated that people were more likely to think that events that were unexpected were more upsetting than events that were expected, $BF_{01}$ `r print_bf(bf_exp2)`.
+We observed this difference consistently across nearly every item (Figure 2).
+
+We note that the most common response was to view both events, which involve identical harms, as equivalently upsetting.
+However, across almost every item, participants showed a clear bias towards choosing unexpected events as more upsetting.
+Thus, Study 2 replicates the tendency to view unexpected moral events as more upsetting using a more conservative judgment task.
+
+# Discussion
 
@@
 Thus, we argue that well-grounded information theoretic principles can account for people’s divergent moral reactions to likely and unlikely events.
-
 In some contexts, it is clearly adaptive that the violation of our expectations would direct our attention to the events and situations about which our understanding is most in need of revision [@Rescorla1972].

### 2018-09-29 paper/paper-apa.Rmd
moved theoretical stuff to appendix

@@
 keywords          : "Evaluation; Moral judgment; Information theory"
-wordcount         : "1911"
+wordcount         : "3898 (total), 1988 (excluding methods and results)"
 
@@
 
-From this it can be shown that the value assigned to event $S$ is proportional to its prior probability (see Theoretical Formalization section in supplemental materials).
+From this it can be shown that the value assigned to event $S$ is proportional to its prior probability (see Appendix).
 Specifically, it is the difference in the expected utility under different values of $S$, weighted by the inverse of the prior probability of $S$.
@@
 
-# General Methods
+# General Method
 
@@
 
-## Study 1
+# Study 1
+
+## Method
 
@@
 
-### Results and discussion
+## Results
 
@@
 
-## Study 2
+# Study 2
 
@@
 In Study 2, we sought to test our hypothesis using a more conservative method: we provided participants with a more expressive response scale so that if they viewed the events under consideration as equally harmful, their responses could reflect their attitude.
+## Method
 
-### Participants
+## Participants
 
@@
 
-### Materials and procedure {#s2-materials-procedure .unnumbered}
+### Materials and procedure
 
